### 2020-05-22 output/paper/appendix.tex
added latex version of paper

% \documentclass[11pt, a4paper]{article}

% \usepackage{upref,amsmath,amssymb,amsthm}
% \usepackage{amsmath}
% \usepackage{graphicx}
% %\usepackage{caption}
% \usepackage[normalem]{ulem}
% %\documentclass[12pt,a4paper]{article}
% %\usepackage{lipsum}
% \usepackage{amsmath,amssymb,upref,amsthm,eucal}
% %\usepackage{pstricks-add}
% %\usepackage{pst-node}
% %\usepackage[T1]{fontenc}
% \usepackage{graphics}
% \usepackage{mathrsfs}
% \usepackage{natbib}

% %%%ADDED BY KRYS
% \usepackage{marginnote} %Package helping with margin notes within equations
%                         %This package works well with ''pdflatex'' but not so well with standard ''latex''
%                         %It is enough to comment the second line below to get rid of all margin-notes

% %The following was added by K.P.
% \usepackage[usenames, dvipsnames]{xcolor}
% \definecolor{darkblue}{HTML}{2C326E}
% %%%
% %%%
% \usepackage{marginnote} %Package helping with margin notes
% %To change the width of paragraph on the margin
% \setlength{\marginparwidth}{2.2cm}
% \newcommand\mn[1]{
% \marginnote{\tiny #1}
% }

% \newcommand\omn[1]{
% \reversemarginpar
% \mn{ #1}
% \normalmarginpar
% }
% %% Label's name on the margin.
% %% To remove label's names from margin simply comment the third line.
% %% It is important not to use any 'Latex' meaningful sequences of characters in labeling.
% %% For example `_` would lead to subscripting in the shown label.
% \newcommand\labmarg[1]{
% \label{#1}\mbox{} %\mbox{} is a trick to make it work in subsection environment
% \mn{#1}
% }

% %\theoremstyle{plain}
% \newtheorem{theorem}{Theorem}
% \newtheorem{lemma}[theorem]{Lemma}
% \newtheorem{remark}[theorem]{Remark}
% \newtheorem{proposition}[theorem]{Proposition}
% %\newtheorem{example}{example}

% %\theoremstyle{definition}
% \newtheorem{definition}[theorem]{Definition}%[section]
% %\newtheorem{definition}{Definition}[subsection]
% \theoremstyle{remark}
% \newtheorem{example}{Example}

% \DeclareMathOperator{\ran}{Ran}
% \DeclareMathOperator{\cov}{Cov}
% \DeclareMathOperator{\vect}{vec}
% \DeclareMathOperator{\tr}{tr}
% \DeclareMathOperator{\var}{Var}
% \DeclareMathOperator{\spant}{Span}
% \newcommand{\XB}{\mathbf{X}}
% \newcommand{\YB}{\mathbf{Y}}
% \newcommand{\UB}{\mathbf{U}}
% \newcommand{\VB}{\mathbf{V}}
% \newcommand{\PB}{\mathbf{P}}
% \newcommand{\PH}{\mathbf{h}}
% \newcommand{\obsy}{\mathscr{Y}}
% \newcommand{\eps}{{\epsilon}}
% \newcommand{\calH}{{\cal H}}
% %%
% % JW added
% %%
% \newcommand{\jw}[1]{\textcolor{red}{#1}}
\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}
% \usepackage{cleveref}
% \newcommand{\Ordo}{\mathcal{O}}
% \newcommand{\E}{\mathbb{E}}
% \usepackage{algorithm}
% \usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
% \algnewcommand{\ffor}[1]{\State\algorithmicfor\ #1\ \algorithmicdo}
% \algnewcommand{\Endffor}{\unskip\ \algorithmicend\ \algorithmicfor}

% \newtheorem{thm}{Theorem}[section]
% \newtheorem{alg}[thm]{Algorithm}

% \usepackage{array,multirow,graphicx}
% \newtheorem{claim}[theorem]{Claim}
% \begin{document}

% \title{Supp}

%  \maketitle

\section{Model}

\subsection{Notation}
Before presenting the model we describe some notation used through out the article for a $m \times n$ matrix $r$ we use the following broadcasting notation $\mv{r}_{k,j:l}=[ r_{k,j}, r_{k,j+1}, \ldots, r_{k,l}]$.
Further $x | y \sim \pi(.)$ implies that the random variable $x$ if we conditioning on $y$ follows distribution $\pi(.)$.
The relevant variables in the model are the following:

	\begin{tabularx}{\linewidth}{ccL}
		Variable name & Dimension & Description \\  \hline
		$\mv{d}$ & $T \times 1$ & $d_i$ is the number of deaths that occurred day $i$. \\
		$\mv{r}$ & $T \times T$ & $r_{ij}$ is number of death recorded for day $i$ at day $j$.  Note that $r_{ij}$ for $i<j$ is not defined.   \\
		$\mv{p}$ & $T \times T$ & $p_{ij}$ is the probability of that a death for day $i$ not yet recorded is recorded at day $j$.
		  Note that $p_{ij}$ for $i<j$ is not defined.  \\
		$\mv{\alpha}$ & $K \times 1$ & Latent prior parameter for $\mv{p}$ \\
		$\mv{\beta}$ & $K \times 1$ & Latent prior parameter for $\mv{p}$ \\
		$\mv{\alpha}^H$ & $2 \times 1$ & parameter for the probability, $\mv{p}$ for holiday adjustment. \\
		$\mv{\beta}^H$ & $2 \times 1$ & parameter for the probability, $\mv{p}$ for holiday adjustment. \\
		$\mv{\lambda}$ &  $T \times 1$ &  $\lambda_i$  is the intensity of the expected number of deaths at day $i$. \\
		$\sigma^2$ & $1\times 1$ & Variation of the random walk prior for the log intensity.

	\end{tabularx}
\subsection{likelihood}
The most complex part of our model is the likelihood, i.e. the density of the observations given the parameters. Here the data consist the daily report of recored deaths for the past days. This can conveniently be represented upper triangular matrix, $\mv{r}$, where $r_{i,j}$ represents number of new reported deaths for day $i$ reported at day $j$. This matrix is displayed on the left in Table \ref{tab:Data}.

\begin{table}
	\centering
	\begin{tabular}{cccccc}
		\multicolumn{1}{c}{} & \multicolumn{5}{c}{Reported date}                                             \\
		\parbox[t]{2mm}{\multirow{5}{*}{\rotatebox[origin=c]{90}{Death date}}}   & $r_{11}$ & $r_{12}$ & $\cdots$ &$\cdots$  &  $r_{1T}$\\
		& & $r_{22}$ &  $\cdots$ & $\cdots$   &$r_{2T}$ \\
		& & &$r_{33}$ &  $\cdots$ &  $r_{3T}$ \\
		& & & &  $\ddots$ & $\vdots$  \\
		& & & &  &  $r_{TT}$ \\

	\end{tabular}

	\caption{The table describes the observations data.}
	\label{tab:Data}
\end{table}

 We assume that given the true number of deaths at day $i$, $d_i$, that each reported day $j$ the remaining death $d_i - \sum_{k=1}^{j-1}r_{i,k}$ each recored with probability $p_{ij}$, i.e. $$r_{i,j}|D_i,r_{1,1:j}.p \sim Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}).$$

Typically in removal sampling one let $p_{i,j}:=p$ however for this data this is clearly not realistic-- given delaying in the reporting. Instead we assume that we have $k$ different probabilities. Further to account for overdispertion we assume that each probability instead follows a Beta distribution. The Beta distibution has two parameters $\alpha$ and $\beta$. This resulting the following distribution for the probabilities
$$
p_{i,j}| \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H  \sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}).
$$
Here, we let $H$ denote holidays and weekends and the parameters above are
$$
\alpha^H_j = \begin{cases}
\alpha_1^H \alpha_2^H & \mbox{if }  \{j\in H \}\cup  \{j-1\in H \},  \\
\alpha_1^H & \mbox{if }  \{j\in H \}\cup  \{j-1\in H^c \}, \\
\alpha_2^H & \mbox{if }  \{j\in H^c \}\cup  \{j-1\in H \}, \\
1 & \mbox{else,}
\end{cases}
$$
and
$$
\beta^H_j = \begin{cases}
\beta_1^H \beta_2^H & \mbox{if }  \{j\in H \}\cup  \{j-1\in H \},  \\
\beta_1^H & \mbox{if }  \{j\in H \}\cup  \{j-1\in H^c \}, \\
\beta_2^H & \mbox{if }  \{j\in H^c \}\cup  \{j-1\in H \}, \\
1 & \mbox{else.}
\end{cases}
$$
These extra parameters are created to account for the under-reporting that occurs during weekend and holidays.

\subsection{Priors}
For the $\mv{\alpha}$ and $\mv{\beta}$ parameters we use an (improper) uniform prior. For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model. However, here we just assume a log-Gaussian Cox processes \cite{Moller1998_log_gaussian} where the Gaussian processes has a intrinsic random walk distribution \cite{Rue2005_gaussian_markov} i.e.
\begin{align*}
\log(\lambda_i) - \log(\lambda_{i-1}) &\sim N(0,\sigma^2),\\
d_i| \lambda_i  &\sim Po(\lambda_i).
\end{align*}
This model is created to create a temporal smoothing between the reported deaths.
For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution.

\subsection{Full model}
Putting the likelihood and priors together we get the following hierarchical Bayesian model
\begin{align*}
\sigma^2 &\sim \Gamma^{-1}(0.01,0.01) \\
\alpha_k &\sim U[0,\infty] \\
\beta_k &\sim U[0,\infty] \\
\alpha_k^H &\sim U[0,\infty] \\
\beta_k^H &\sim U[0,\infty] \\
\log(\lambda_i) - \log(\lambda_{i-1}) &\sim N(0,\sigma^2)\\
d_i| \lambda_i  &\sim Po(\lambda_i) \\
p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
\end{align*}
where where and $j\leq i$ and $i=1,\ldots,T$.

\section{Inference}
As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \cite{Brooks2011_handbook_markov}. In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
\begin{itemize}
	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution. Here to we use an adaptive MALA \cite{Atchade2006_adaptive_version} to sample from these parameters.
	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
	\item Finally we sample $\sigma^2|\mv{d}$ directly since this distribution is explicit.
\end{itemize}

\section{Model Benchmark}
In this section, we present additional comparison of the model to a simple constant model. The benchmark model is the sum of average reporting lag for the preceding 14 days.

% TODO: Add figures

% \bibliographystyle{chicago}
% \bibliography{researchplan}
% \end{document}

### 2020-05-22 output/paper/paper.tex
added latex version of paper

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode
% !BIB TS-program = biber
% !TeX root = ./paper.tex
\documentclass[a4paper,11pt,article,oneside,openany,american]{memoir}
\usepackage[memoir,bibliography]{setupscript}
\addbibresource{bibliography.bib}

% Customization
\usepackage{array}
\usepackage{multirow}
\usepackage{tabularx}
\newcolumntype{L}{>{\raggedright\arraybackslash}X}

% TITLE

\title{Nowcasting Covid-19 statistics reported with delay: a case-study of Sweden}
\date{\today}
\author{
	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
	Joacim Rocklöv \and
	Jonas Wallin
}

\begin{document}
\frontmatter%
\begin{titlingpage}
    \maketitle
    \begin{abstract}
        The new corona virus disease - COVID-2019- is rapidly spreading through the world. The availability of unbiased timely statistics of trends in disease events are a key to effective responses. But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend. Here we describe a statistical methodology for predicting the actual daily number occurring events a specific day, and its uncertainty, based on the daily reported event frequency. The methodology takes into account the observed distribution pattern of the delay and can be derived from the “removal method”, a well-established method in the field of ecology.
	\end{abstract}

	\noindent Keywords:

\end{titlingpage}
%\tableofcontents*
%%%%%%%%%%%%%%%%%%%
% MAIN TEXT SECTION
\mainmatter%

\chapter{Pandemic response demands timely data}
The new corona virus pandemic is affecting societies all around the world. As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions (\cite{No_author_2020_coronavirus_three}). Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}). The latest data are often not finalized, but change as new statistics are reported. In fact, since these updates are usually in the form of delayed reporting. Since the most recent days then have the least cases, such systematic underreporting gives the dangerous picture of an always improving situation.

Still, these unfinished statistics offer crucial information. If the pandemic is indeed slowing, we should not wait for the data to be finalized before using it. Rather, we argue that actual case counts and deaths should be nowcasted to account for any reporting delay and ensure policymakers are using the most accurate numbers available.

Such predictions provide an additional feature that is perhaps even more important. They explicitly model the uncertainty about these unknown quantities, ensuring that all users of these data have the same view of the current state of the epidemic.

In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and the degree of uncertainty, based on the daily reported event frequency and the observed distribution pattern of the reporting delay. The prediction model is building on methods developed in ecology, referred to as the “removal methods” or “capture-retain” models (\cite{Pollock1991_review_papers}).

\section{The current situation of COVID-19 in Sweden}
Each day at 14:00 the Swedish Public Health Agency holds a press conference where new COVID-19 statistics are presented\footnote{The data is then published on https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/ where we download it every day.}. Deaths is one of the main indicators to follow for understanding the impact of the pandemic on public health in Sweden, but also the number of new admissions to critical care, to hospitals, and the new confirmed cases are reported. One of the reasons for following these indicators is to enable public health professionals and the public to observe the patterns of the evolving, flattened or suppressed epidemic (\cite{Anderson2020_how_will}). In relation to policy, it is of further interest to understand if growth rates change, which could indicate a potential response. However, at the daily presentation only a proportion of the number deaths for each of the most recent days is yet known, and this bias causes an artificial, downward, trend in the data.

The death counts suffer from the longest reporting day. In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trendline 10 days before the latest date. But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information from the most recent days creating a sense of a downward trend. In fact, this might be the reason why the number of deaths has been underestimated repeatedly. At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual level was closer to 100 (\cite{Ohman2020_antalet_virusdoda}).

\chapter{The removal method}
We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty. The method has a long history dating back at least to the 1930s (\cite{Leslie1939_attempt_determine}). However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today (\cite{Matechou2016_open_models}). It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.

The removal method that has three major advantages over simply reporting moving averages:
\begin{itemize}
	\item it does not relay previous trend in the data,
	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day,
	\item the uncertainty in the estimate can be carried over to epidemiological models that uses the estimate as input, and hence give more realistic models.
\end{itemize}

A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals (\cite{Pollock1991_review_papers}). Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline. This pattern of declining number of trapped animals allows one to draw inference of the underlying population size. Here we replace the animal population with the true number of deaths or cases in a given day. Instead of traps we have the new reports of COVID-19 events. As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day. In fact, if we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.

Suppose for example that on day one, 4 individuals are reported dead for that day. On the second day, 10 deaths are recorded for day two. Then, with no further information, it is reasonable to assume that more people died on day two. If the proportion reported on the first day is 3\%, the actual number of deaths would be 133 for day one and 333 for day two.

If additionally, 60 deaths are reported during the second day to have happened during day one, and on the third day, only 40 are reported for day two, we now have conflicting information. From the first-day reports it seemed like more people had died during day two, but the second day-reports gave the opposite indication. The model we propose systematically deals with such data, and handles many other sources of systematic variation in reporting delay. In fact, the Swedish reporting lag follows a calendar pattern. The number of events reported during weekends is much smaller. To account for this, we  allow the estimated proportions of daily reported cases to follow a probability distribution taking into consideration what type of day it is.

\chapter{Applying the model to COVID-19 in Sweden}
We prorpose a Bayesian version of the removal model that assumes anoverdispersed binomial distribution for the  the daily observations of deaths in Sweden in COVID-19. We then calculate the posterior distribution and prediction median and 95\% uncertainty intervals of the expected deaths from the reported deaths on each specific day. The method and algorithm is thoroughly described in the Supplementary Information.

In Figure 1 we illustrate the similarity and difference between the 7-day moving average and the new Bayesian prediction model with 95\% prediction intervals to the reported number of deaths from COVID-19 in Sweden. The model provides estimates of actual deaths considerably above the reported number of deaths and the uncertainty of the estimate.

% TODO: FIGURE 1 HERE
%Figure 1. The 7-day moving average (blue) and the prediction from the Bayesian model (black) with prediction intervals overlaid. The grey bars represent the number of reported deaths up to May the 12th, 2020, in COVID-19 in Sweden.

%In Figure 2 we use the method to predict the number of new deaths that will be reported in the most recent report and how their date of death will be distributed over time. It is clear from the model prediction that the delay is substantial and therefore interpretations of the daily reported number of deaths without these models to adjust it are difficult.

% TODO: FIGURE 2 here
%Figure 2. The bars represent deaths in COVID-19 in Sweden and the prediction intervals represent the Bayesian predictions of the number of deaths for the specific day in the latest report.

\chapter{Implications and limitations}
The model proposed here has much better ability to estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden. To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks. In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data. The method and algorithm proposed overcome major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic. It also provides valuable measures of uncertainty around these estimates, showing readers how large the range of possible outcomes can be.

Whenever case statistics are collected from multiple sources and attributed to its actual event date in the middle of a public health emergency, similar reporting delays to the ones in Sweden will necessarily occur. The method described thus has implications and value beyond Sweden, for any situation where nowcasts of disease event frequencies are of relevance to public health.

Nevertheless, the method also has its limitations. As presented, the model assumes that all deaths are reported in the same manner, given there exists many regions in Sweden this is unlikely to be the case. For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure compared to Stockholm. Building a model for each region separately would most likely give better results and make the assumptions more reasonable. Unfortunately we do not currently have access to the high resolution data required to do so. Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days. The number of such cases is very small, however, and its removal should not make much difference. The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates). If actual reporting standards change over time and its not explicitly modelled by a covariate, the model will not be able to account for this. But reporting likely becomes faster as the crisis infrastructure improves. One can imagine that after a while the reporting improves, or is changed, if this is not accounted for by a covariate in the model, it will report incorrect numbers. Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages.

\chapter{Conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
\printbibliography%
\backmatter%
\appendix%
\chapter{Appendix}
\input{appendix.tex}%

\end{document}

### 2020-05-22 src/Beta_GP_benchmark_build.R
updating progg

@@ -8,14 +8,15 @@ source(file.path("src", "util.r"))
 source(file.path("src", "MH.R"))
 source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
+source(file.path("src", "modelbenchmarkutil.R"))
 nclust <- 6
-MCMC_sim <- 30000
-burnin_p = 0.5
-deaths_sim <- 10
+MCMC_sim <- 2000
+burnin_p = 0.3
+deaths_sim <- 5
 maxusage.day = 20 #must be less then N
 unique.days  = 7
 true.day = 5
-start.predict.day = 26#16 # more then unique days
+start.predict.day = 16#16 # more then unique days
 
 result <- readRDS(file.path("data", "processed", "processed_data.rds"))
 Reported_T = result$detected
@@ -29,113 +30,20 @@ data_T <- newDeaths(deaths_est_T,
 X_T <- setup_data(N_T, maxusage.day, result$dates_report, unique.days)
 predicition.list <-list()
 
+
 cl <- parallel::makeCluster(nclust,setup_strategy = "sequential")
 doParallel::registerDoParallel(cl)
 foreach(j = start.predict.day:N_T)  %dopar% {
-  library(Matrix)
   result <- readRDS(file.path("data", "processed", "processed_data.rds"))
-  Reported <- result$detected[1:j,1:j]
-  dates_report <- result$dates_report[1:j]
-  N <- dim(Reported)[1]
-  deaths_est <- apply(Reported, 1, max, na.rm=T)
-  deaths_est[1:true.day] = deaths_est_T[1:true.day] #days with known deaths
+  res_save<-benchmark_BetaGP_j(j,
+                               result,true.day,
+                               maxusage.day,
+                               unique.days,
+                               MCMC_sim,
+                               burnin_p,
+                               deaths_sim = deaths_sim,
+                               prior = c(0,0))
 
-
-  data_ <- newDeaths(deaths_est,
-                     Reported,
-                     maxusage.day =maxusage.day)
-
-
-  ##
-  # building covariate matrix
-  ##
-  X <- setup_data(N, maxusage.day,  dates_report, unique.days)
-
-  ##
-  # seting up MCMC
-  ##
-  MH_obj <- MH_setup()
[…]

### 2020-05-22 src/GPutil.R

@@ -1,14 +1,45 @@
 library(Matrix)
+library(invgamma)
+#'
+#' Measument error prior
+#' Q = (tau Q_0 + tau_E^-1 I)^{-1}
+#'  @param x - (2x1) tau tau_E
+#'  @param Ltheta    - (n x 1) KL base elements
+#'  @param Q_E       - (n x 1) eigenvalues of Q_0
+tau_sigma_prior <- function(x, L_theta, Q_E){
+  tau <- x[1]
+  tau_E <- x[2]
+  if(min(tau,tau_E) <0)
+    return(list(loglik=-Inf))
+  D <- sqrt(1/(1/(tau*Q_E) + 1/tau_E))
+  lik <- -1/2 * sum((D*L_theta)^2 )
+  lik <- lik + sum(log(D[abs(D)>10e-14]))
+  lik <- lik +  dinvgamma(tau, shape= 0.01, rate = 0.01, log = T)
+  lik <- lik +  dinvgamma(tau_E, shape= 0.01, rate = 0.01, log = T)
+  return(list(loglik = lik))
+}
 
 #'
-#' second order random walk model (intrisinct)
+#' General Gaussian prior
+#'
+#'  @param theta - (nx1) the processes
+#'  @param L     - (nxn) cholesky preicison factor
+prior_GP <- function(theta, L){
+  lik <- -1/2 * sum((L%*%theta)^2)
+  Q = t(L)%*%L
+  grad <-  -(Q%*%theta)
+  Hessian <- - Q
+  return(list(loglik = lik, grad = grad, Hessian = Hessian))
+}
+#'
+#' first order random walk model (intrisinct)
 #'  d theta /dt = N(0,\tau)
 #'  @param theta - (nx1)
 #'  @param tau   - precision
 #'  @param alpha - prior parameter
 #'  @param beta  - prior parameter
 prior_s1d <-function(theta, tau, alpha = 0.01, beta = 0.01){
-  
+
   n <- length(theta)
   L <- toeplitz(c(-1,1, rep(0,n-2)))
   L[lower.tri(L,diag=F)] <- 0
@@ -26,13 +57,38 @@ prior_s1d <-function(theta, tau, alpha = 0.01, beta = 0.01){
   return(list(loglik = lik, grad = grad, Hessian = Hessian, tau=tau_sample))
 }
 
+#'
+#' second order random walk model (intrisinct)
+#'  d theta /dt = N(0,\tau)
[…]

### 2020-05-22 src/MultinomialBetaMCMC_with_GP.R

@@ -10,7 +10,7 @@ source(file.path("src", "GPutil.R"))
 
 path.to.files <- file.path("..","data")
 MCMC_sim <- 20000
-burnin_p = 0.5
+burnin_p = 0.33
 deaths_sim <- 10
 maxusage.day = 20 #must be less then N
 unique.days  = 7

### 2020-05-22 src/plots_all.R

@@ -1,0 +1,77 @@
+library(data.table)
+library(fst)
+library(forcats)
+library(ggplot2)
+library(hrbrthemes)
+library(wesanderson)
+source("src/functions.R")
+
+#
+w <- 11 # plot width (inches)
+
+# Load data
+benchmark <- read_fst(file.path("data", "processed", "constant_benchmark.fst"), as.data.table = TRUE)
+model <- read_fst(file.path("data", "processed", "model_benchmark.fst"), as.data.table = TRUE)
+
+# Fix data tables so they look the same
+reported_dead <- benchmark[, .(state, date, days_left = as.integer(days_left), reported_dead)]
+benchmark <- benchmark[!is.na(target), .(state, date, target, days_left = as.integer(days_left), ci_upper, ci_lower, predicted_deaths, SCRPS)]
+model[, days_left := as.integer(days_left)]
+# Add a final 14-day prediction to model (just equal to truth)
+model <- rbindlist(list(model, data.table(state = model[, unique(date)] + 14, date = model[, unique(date)], days_left = 14)), use.names = TRUE, fill = TRUE)
+
+# Days left as factor
+reported_dead[, days_left := forcats::fct_rev(factor(14 - days_left))]
+benchmark[, days_left := forcats::fct_rev(factor(14 - days_left))]
+model[, days_left := forcats::fct_rev(factor(14 - days_left))]
+
+# Order correctly
+setkey(reported_dead, date, state, days_left)
+setkey(benchmark, date, state, days_left)
+setkey(model, date, state, days_left)
+
+# Add target as prediction for last day
+model[is.na(target), predicted_deaths := model[!is.na(target), unique(target), by = date][, V1]]
+
+
+plot_data <- rbindlist(list("Historical Avg." = benchmark, "Capture-Retain" = model), idcol = "type")
+plot_data[, type := factor(type)]
+
+dates <- seq(as.Date("2020-04-15"), model[days_left == 13, max(date)], 1)
+plot_data_full <- plot_data
+for(i in 1:length(dates)){
+    plot_data <- plot_data_full[date %in% dates[i]]
+
+colors <- c("#ECCBAE", "#046C9A", "gray50")
+colors <- setNames(colors, c(levels(plot_data$type), "Reported"))
+
+plot <- ggplot(data = plot_data,
+               aes(x = days_left,
+                   y = predicted_deaths,
+                   color = type,
+                   group = type)) +
[…]

### 2020-05-22 src/data_processing.R
no percentage diff needed

@@ -62,8 +62,8 @@ deaths_dt[date == "2020-04-02" & publication_date == "2020-04-02", days_since_pu
 
 deaths_dt[!is.na(date), paste0("n_m", 1) := shift(N, n = 1, type = "lag", fill = 0L), by = date]
 deaths_dt[!is.na(date), n_diff := N - n_m1]
-deaths_dt[!is.na(date) & n_m1 > 0 & !is.na(n_m1), n_diff_pct := N/n_m1 - 1]
-deaths_dt[!is.na(date) & n_m1 == 0 & N == 0, n_diff_pct := 0]
+# deaths_dt[!is.na(date) & n_m1 > 0 & !is.na(n_m1), n_diff_pct := N/n_m1 - 1]
+# deaths_dt[!is.na(date) & n_m1 == 0 & N == 0, n_diff_pct := 0]
 deaths_dt[, n_m1 := NULL]
 
 setkey(deaths_dt, date, publication_date)

### 2020-05-22 src/plots.R
added daily plots

@@ -34,54 +34,73 @@ setkey(model, date, state, days_left)
 model[is.na(target), predicted_deaths := model[!is.na(target), unique(target), by = date][, V1]]
 
 #
-## PLOT 1: Performance on 4 random dates ##
+## PLOT 1: Performance per day, PLOT1=4 random dates ##
 #
+
+# Plot a specific day
+day_plot <- function(DT, reported, plot.title) {
+    colors <- c("#ECCBAE", "#046C9A", "gray50")
+    colors <- setNames(colors, c(levels(DT$type), "Reported"))
+
+    plot <- ggplot(data = DT,
+                aes(x = days_left,
+                    y = predicted_deaths,
+                    color = type,
+                    group = type)) +
+        geom_hline(data = DT[!is.na(target), .(unique(target)), by = .(date)],
+                aes(yintercept = V1), color = "grey50") +
+        geom_line(data = reported,
+                aes(y = reported_dead, group = "Reported", color = "Reported"), linetype = "dashed") +
+        geom_point(data = reported,
+                aes(y = reported_dead, group = "Reported", color = "Reported")) +
+        # The actual models
+        geom_line(position = position_dodge(width = 0.6)) +
+        geom_point(data = DT[days_left != "0"], position = position_dodge(width = 0.6)) +
+        geom_errorbar(data = DT[days_left != "0"], aes(ymin = ci_lower, ymax = ci_upper),
+                    width = 0.7, position = position_dodge(width = 0.6)) +
+        # Converging with a grey point on the last day
+        geom_point(data = DT[days_left == "0" & type == "Historical Avg."], color = "grey50") +
+        # Theming
+        set_default_theme() +
+        # scale_fill_manual(values = fill_colors, limits = label_order, drop = FALSE) +
+        # scale_color_manual(values = wes_palette("Darjeeling2")) +
+        scale_color_manual(values = colors) +
+        scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(0, 5))) +
+        labs(title = plot.title,
+            subtitle = "",
+            caption = "",
+            color = "Model",
+            x = "Days of lag to predict",
+            y = "Number of deaths")
+
+    if (DT[, uniqueN(date)] > 1) {
+        # One plot for each day
+        plot <- plot + facet_wrap(~date)
+    }
+    return(plot)
+}
+
[…]

### 2020-05-22 src/Beta_GP_benchmark_build.R
prediciton output

@@ -10,7 +10,7 @@ source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
 source(file.path("src", "modelbenchmarkutil.R"))
 nclust <- 6
-MCMC_sim <- 2000
+MCMC_sim <- 20000
 burnin_p = 0.3
 deaths_sim <- 5
 maxusage.day = 20 #must be less then N
@@ -42,7 +42,7 @@ foreach(j = start.predict.day:N_T)  %dopar% {
                                MCMC_sim,
                                burnin_p,
                                deaths_sim = deaths_sim,
-                               prior = c(0,0))
+                               prior = c(1,0))
 
   save(res_save,
        file = file.path("data", "simulation_results", paste0("param_", dates_report[j], ".rds")))

### 2020-05-22 src/MultinomialBetaMCMC_with_GP.R

@@ -3,21 +3,23 @@
 # then profile likelihood of N
 ##
 set.seed(2)
+library(data.table)
 source(file.path("src", "util.r"))
 source(file.path("src", "MH.R"))
 source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
+source(file.path("src", "modelbenchmarkutil.R"))
+
 
 path.to.files <- file.path("..","data")
-MCMC_sim <- 20000
+MCMC_sim <- 10000
 burnin_p = 0.33
 deaths_sim <- 10
 maxusage.day = 20 #must be less then N
 unique.days  = 7
 true.day = 5
-start.predict.day = 14# more then unique days
-
 result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+
 Reported = result$detected
 N <- dim(Reported)[1]
 
@@ -26,151 +28,20 @@ deaths_est <- apply(Reported, 1, max, na.rm=T)
 data_ <- newDeaths(deaths_est,
                    Reported,
                    maxusage.day =maxusage.day)
-X <- setup_data(N, maxusage.day, result$dates_report, unique.days)
 
-##
-# seting up MCMC
-##
-MH_obj <- MH_setup()
-MH_obj$sigma <- 0.1
-MH_obj$theta <- rep(0,2*dim(X)[2])
+res_save<-benchmark_BetaGP_j(N,
+                             result,true.day,
+                             maxusage.day,
+                             unique.days,
+                             MCMC_sim,
+                             burnin_p,
+                             deaths_sim = deaths_sim,
+                             prior = c(0,0))
 
+CI <-apply(res_save$Death_est,2 , function(x){ quantile(x,c(0.05,0.5,0.95))})
 
-MH_obj$Lik <- loglikProbBB
-
-##
-# seting up MCMC GP
-##
-MH_obj_GP <- MH_setup()
-MH_obj_GP$sigma <- 0.1
-#probably better do something smarter
-MH_obj_GP$theta <- log(deaths_est+1)
-tau <- 0.1
-MH_obj_GP$Lik <- function(x, N, tau) {
-  prior_list <- prior_s1d(x, tau)
-  lik_list   <- lik_poisson(N, x)
-  return(list(loglik = lik_list$loglik + prior_list$loglik,
-              grad = lik_list$grad  + prior_list$grad,
[…]

### 2020-05-22 src/modelbenchmarkutil.R

@@ -108,6 +108,16 @@ benchmark_BetaGP_j <- function(j,
     Alpha <- matrix(NA, ncol=N, nrow=N)
     Beta <- matrix(NA, ncol=N, nrow=N)
     for(i in 1:(MCMC_sim+burnin-1)){
+
+        if(i%%100==0){
+            cat('*')
+        }
+        if(i%%1000==0){
+            cat('+')
+        }
+        if(i%%10000==0){
+            cat(' ',i/10000,' ')
+        }
         MH_obj <- MALAiter(MH_obj, TRUE,
                            death.remain = data_$death.remain,
                            report.new   = data_$report.new,
@@ -134,14 +144,13 @@ benchmark_BetaGP_j <- function(j,
         ##
         # build tau sampler
         ##
-        E_Q <- t(EQ$vectors)
-        D_Q <- EQ$values
+
         if(prior[2] == 0){
-            L_theta          <- sqrt(D_Q)*(E_Q%*%MH_obj_GP$theta)
+            L_theta          <- L%*%MH_obj_GP$theta
 
             tau_div   <- 1/rgamma(1, shape= (1 + length(L_theta)-1)/2 + 0.01, rate = sum(L_theta^2)/2 + 0.01)
             tau <- 1/tau_div
-            L_in <- diag(sqrt(tau*D_Q))%*%E_Q
+            L_in <- sqrt(tau) * L
         }else{
             L_theta          <- (E_Q%*%MH_obj_GP$theta)
             MH_obj_tau <- MHiter(MH_obj_tau,

### 2020-05-22 src/util.r

@@ -586,7 +586,7 @@ death.givenParamBB <- function(alpha, beta, Reported,Predict.day = Inf,sim=c(200
 # unique.prob - (int) how many of days shall we have unique probabilites
 ##
 setup_data <- function(N, Predict.day, dates_report, unique.prob=NULL){
-  holidays.Sweden <- as.Date(c("2020-04-10","2020-04-13","2020-05-01"))
+  holidays.Sweden <- as.Date(c("2020-04-10","2020-04-13","2020-05-01","2020-05-21"))
   X <- buildXdayeffect(N,Predict.day)
   #holidays and weekends
 

### 2020-05-23 src/plots.R
added model predictions

@@ -9,9 +9,43 @@ source("src/functions.R")
 #
 w <- 11 # plot width (inches)
 
+# Plot 1 = Predictions and current stats
+deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
+model_predict <- read_fst(file.path("data", "processed", "model_predict.fst"), as.data.table = TRUE)
+
+DT1 <- model_predict[date >= Sys.Date() - 28]
+DT2 <- deaths_dt[!is.na(N) & !is.na(date) & date >= "2020-03-15" &
+                 publication_date == max(publication_date)]
+
+colors <- c("gray80", "#ECCBAE")
+colors <- setNames(colors, c("Reported dead", "Model prediction"))
+
+plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
+    geom_bar(aes(fill = "Model prediction"), stat = "identity") +
+    geom_bar(data = DT2,
+             aes(y = N, fill = "Reported dead"), stat = "identity") +
+    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
+                  width = 0.4, color = "#e0ac7e") +
+    set_default_theme() +
+    scale_fill_manual(values = colors) +
+    scale_x_date(date_breaks = "3 day", date_labels = "%b %d", expand = expansion(add = 0.8)) +
+    theme(axis.text.x = element_text(angle = 35, hjust = 1.3, vjust = 1.1)) +
+    scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(0, 5))) +
+    labs(title = paste0("Reported deaths as of ", deaths_dt[, max(publication_date)], " and model prediction"),
+        subtitle = "",
+        caption = "",
+        fill = "",
+        x = "Death date",
+        y = "Number of deaths")
+
+ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
+       plot = plot, device = cairo_pdf, width = w, height = w/1.9)
+
+
 # Load data
 benchmark <- read_fst(file.path("data", "processed", "constant_benchmark.fst"), as.data.table = TRUE)
 model <- read_fst(file.path("data", "processed", "model_benchmark.fst"), as.data.table = TRUE)
+
 
 # Fix data tables so they look the same
 reported_dead <- benchmark[, .(state, date, days_left = as.integer(days_left), reported_dead)]

### 2020-05-23 src/plots.R
added moving average

@@ -14,8 +14,9 @@ deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.t
 model_predict <- read_fst(file.path("data", "processed", "model_predict.fst"), as.data.table = TRUE)
 
 DT1 <- model_predict[date >= Sys.Date() - 28]
-DT2 <- deaths_dt[!is.na(N) & !is.na(date) & date >= "2020-03-15" &
-                 publication_date == max(publication_date)]
+DT2 <- deaths_dt[!is.na(N) & !is.na(date) & publication_date == max(publication_date)]
+DT2[, avg := frollmean(N, 7, algo = "exact", align = "center")]
+DT2 <- DT2[date >= "2020-03-15"]
 
 colors <- c("gray80", "#ECCBAE")
 colors <- setNames(colors, c("Reported dead", "Model prediction"))
@@ -24,6 +25,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
     geom_bar(aes(fill = "Model prediction"), stat = "identity") +
     geom_bar(data = DT2,
              aes(y = N, fill = "Reported dead"), stat = "identity") +
+    geom_line(data = DT2[!is.na(avg)], aes(y = avg, linetype = "7-day moving average"), color = "grey50") +
     geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                   width = 0.4, color = "#e0ac7e") +
     set_default_theme() +
@@ -35,6 +37,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
         subtitle = "",
         caption = "",
         fill = "",
+        linetype = "",
         x = "Death date",
         y = "Number of deaths")
 

### 2020-05-23 output/paper/paper.tex
Updates from Overleaf

@@
     \begin{abstract}
-        The new corona virus disease - COVID-2019- is rapidly spreading through the world.
+        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world.
 The availability of unbiased timely statistics of trends in disease events are a key to effective responses.
@@
 Here we describe a statistical methodology for predicting the actual daily number occurring events a specific day, and its uncertainty, based on the daily reported event frequency.
-The methodology takes into account the observed distribution pattern of the delay and can be derived from the “removal method”, a well-established method in the field of ecology.
+The methodology takes into account the observed distribution pattern of the delay and can be derived from the ``removal method'', a well-established method in the field of ecology.
 	\end{abstract}
@@
 In fact, since these updates are usually in the form of delayed reporting.
-Since the most recent days then have the least cases, such systematic underreporting gives the dangerous picture of an always improving situation.
+Since the most recent days then have the least cases, such systematic under reporting gives the dangerous picture of an always improving situation.
 
@@
 In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and the degree of uncertainty, based on the daily reported event frequency and the observed distribution pattern of the reporting delay.
-The prediction model is building on methods developed in ecology, referred to as the “removal methods” or “capture-retain” models (\cite{Pollock1991_review_papers}).
+The prediction model is building on methods developed in ecology, referred to as the ``removal methods'' or ``capture-retain'' models (\cite{Pollock1991_review_papers}).
 
 \section{The current situation of COVID-19 in Sweden}
-Each day at 14:00 the Swedish Public Health Agency holds a press conference where new COVID-19 statistics are presented\footnote{The data is then published on https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/ where we download it every day.}.
+Each day at 14:00 the Swedish Public Health Agency holds a press conference where new COVID-19 statistics are presented\footnote{The data is then published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/} where we download it every day.}.
 Deaths is one of the main indicators to follow for understanding the impact of the pandemic on public health in Sweden, but also the number of new admissions to critical care, to hospitals, and the new confirmed cases are reported.
@@
 The death counts suffer from the longest reporting day.
-In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trendline 10 days before the latest date.
+In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trend line 10 days before the latest date.
 But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information from the most recent days creating a sense of a downward trend.
@@
 Here we replace the animal population with the true number of deaths or cases in a given day.
-Instead of traps we have the new reports of COVID-19 events.
+Instead of traps we have the new reports of COVID--19 events.
 As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day.
@@
 \chapter{Applying the model to COVID-19 in Sweden}
-We prorpose a Bayesian version of the removal model that assumes anoverdispersed binomial distribution for the  the daily observations of deaths in Sweden in COVID-19.
-We then calculate the posterior distribution and prediction median and 95\% uncertainty intervals of the expected deaths from the reported deaths on each specific day.
+We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19.
+We then calculate the posterior distribution, prediction median and 95\% uncertainty intervals of the expected deaths from the reported deaths on each specific day.
 The method and algorithm is thoroughly described in the Supplementary Information.
+
+In Figure 1 we apply the model to the latest statistics from Sweden.
+The graph shows reported and predicted deaths (with uncertainty intervals) as bars and a dashed line plots the 7-day (centered) moving average.
 
@@
 The model provides estimates of actual deaths considerably above the reported number of deaths and the uncertainty of the estimate.
+
+\begin{figure}
+    \centering
+    \includegraphics{../plots/}
+    \caption{Caption}
+    \label{fig:my_label}
+\end{figure}
 
@@
 Unfortunately we do not currently have access to the high resolution data required to do so.
-Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days.
+Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or mis classification of days.
 The number of such cases is very small, however, and its removal should not make much difference.

### 2020-05-23 src/plots.R
removed titles

@@ -33,7 +33,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
     scale_x_date(date_breaks = "3 day", date_labels = "%b %d", expand = expansion(add = 0.8)) +
     theme(axis.text.x = element_text(angle = 35, hjust = 1.3, vjust = 1.1)) +
     scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(0, 5))) +
-    labs(title = paste0("Reported deaths as of ", deaths_dt[, max(publication_date)], " and model prediction"),
+    labs(#title = paste0("Reported deaths as of ", deaths_dt[, max(publication_date)], " and model prediction"),
         subtitle = "",
         caption = "",
         fill = "",
@@ -125,7 +125,8 @@ set.seed(1234)
 example_dates <- sample(seq(as.Date("2020-04-15"), model[days_left == 13, max(date)], 1), 4)
 plot <- day_plot(plot_data[date %in% example_dates],
                  reported_dead[date %in% example_dates],
-                 "Predicting the number of deaths in a given day reported within 14 days.")
+                 "")
+                 #"Predicting the number of deaths in a given day reported within 14 days.")
 ggsave(filename = file.path("output", "plots", "lag_prediction_by_date.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w)
 
@@ -167,7 +168,7 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
     facet_wrap(~variable, scales = "free_y") +
     set_default_theme() +
     scale_color_manual(values = wes_palette("Darjeeling2")) +
-    labs(title = "Model metrics",
+    labs(#title = "Model metrics",
          subtitle = "",
          caption = "",
          color = "Model",
@@ -193,7 +194,7 @@ plot <- ggplot(data = plot_data, aes(x = state, y = V1, color = type, group = ty
     geom_line() + geom_point() +
     set_default_theme() +
     scale_color_manual(values = wes_palette("Darjeeling2")) +
-    labs(title = "Model metrics",
+    labs(#title = "Model metrics",
          subtitle = "",
          caption = "",
          color = "Model",
@@ -223,7 +224,7 @@ plot <- ggplot(data = plot_data, aes(x = dayofweek, y = V1, color = type, group 
     # facet_wrap(~days_left) +
     set_default_theme() +
[…]

### 2020-05-23 src/Beta_GP_benchmark.R
model update

@@ -6,7 +6,8 @@ library(foreach)
 library(fst)
 library(data.table)
 source("src/util.r")
-nclust <- 6
+nclust <- 1
+maxusage.day <- 20
 path.to.files <- file.path("data")
 files <- list.files(paste(path.to.files,"/simulation_results",sep=""),
                              pattern = "^param",
@@ -48,9 +49,9 @@ result_par <- foreach(i = index_files)  %dopar% {
                       Reported_T,
                       maxusage.day =Inf)
   detected_mod<- result$detected
-  for(k in 1:dim(data_T$report.new)[1]){
-    detected_mod[i,] <- cumsum(replace_na(data_mod$report.new[i,],0))
-  }
+  for(k in 1:dim(data_T$report.new)[1])
+    detected_mod[k,] <- cumsum(replace_na(data_mod$report.new[k,],0))
+
   detected_mod[is.na(result$detected)] <- NA
 
 

### 2020-05-23 src/Beta_GP_benchmark_build.R

@@ -31,10 +31,10 @@ X_T <- setup_data(N_T, maxusage.day, result$dates_report, unique.days)
 predicition.list <-list()
 
 
-cl <- parallel::makeCluster(nclust,setup_strategy = "sequential")
+cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
 doParallel::registerDoParallel(cl)
 foreach(j = start.predict.day:N_T)  %dopar% {
-  result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+    result <- readRDS(file.path("data", "processed", "processed_data.rds"))
   res_save<-benchmark_BetaGP_j(j,
                                result,true.day,
                                maxusage.day,
@@ -42,10 +42,10 @@ foreach(j = start.predict.day:N_T)  %dopar% {
                                MCMC_sim,
                                burnin_p,
                                deaths_sim = deaths_sim,
-                               prior = c(1,0))
+                               prior = c(0,0))
 
   save(res_save,
-       file = file.path("data", "simulation_results", paste0("param_", dates_report[j], ".rds")))
+       file = file.path("data", "simulation_results", paste0("param_", result$dates_report[j], ".rds")))
 }
 
 parallel::stopCluster(cl)

### 2020-05-23 src/GPutil.R

@@ -155,3 +155,94 @@ lik_poisson2<- function(N, theta){
   return(list(loglik = lik, grad = grad, Hessian = Hessian))
 }
 
+
+dnegbin <- function(x, mu, phi, log.p=T){
+  lik <- lgamma(x+phi)-lgamma(phi) - lgamma(x+1)
+  lik <- lik + x*log(mu)
+  lik <- lik + phi * log(phi)
+  lik <- lik - (phi+x) * log(phi + mu)
+  if(log.p)
+    return(lik)
+  return(exp(lik))
+}
+
+
+#'
+#' Negative Biniomial log likelihood (phi)
+#' N     - (n x 1) Poisson observation
+#' theta - (m x 1) log intens
+#' phi   - (1 x 1) over disperstion
+#'
+lik_negbin_phi <- function(N, theta, phi){
+  mu <- exp(theta)
+  lik <- dnegbin(N, mu, phi)
+  grad <- digamma(N + phi) -digamma(phi)
+  grad <- grad  + log(phi) + 1
+  grad <- grad  - log(phi + mu)
+  grad <- grad  - (phi+N)/(phi + mu)
+  return(list(loglik = sum(lik), grad = sum(grad)))
+}
+
+#'
+#' Negative Biniomial log likelihood (theta)
+#' N     - (n x 1) Poisson observation
+#' theta - (m x 1) log intens
+#' phi   - (1 x 1) over disperstion
+#'
+lik_negbin_theta <- function(N, theta, phi){
+  mu <- exp(theta)
+  lik <- dnegbin(N, mu, phi)
+  grad <- mu * (N/mu - (phi+N)/(phi + mu))
+  return(list(loglik = sum(lik), grad = grad))
+}
+
+##
+#' testing lik_negbin_theta
+#' N     - (n x 1) Poisson observation
+#' theta - (m x 1) log intens
+#' phi   - (1 x 1) over disperstion
+##
+test_negbin <- function(N, theta, phi){
+
+  base <- lik_negbin_theta(N, theta, phi)
+  base2 <- lik_negbin_phi(N, theta, phi)
+  lik <- base$loglik
+  grad_theta <- base$grad
+  grad_phi <- base2$grad
+  eps <- 1e-6
+  grad_num_theta <- rep(0,length(theta))
+  grad_num_phi   <- (lik_negbin_theta(N, theta, phi+eps)$loglik - lik)/eps
+  for(i in 1:length(theta)){
[…]

### 2020-05-23 src/MH.R

@@ -253,12 +253,12 @@ MH_MALA <- function(MH_obj,...)
 
   if(Le==FALSE)
   {
-    mu_old <- MH_obj$theta + MH_obj$grad_old * MH_obj$sigma^2/2
-    theta_star <- mu_old  + MH_obj$sigma * rnorm(length(MH_obj$theta))
+    mu_old <- as.vector(MH_obj$theta + MH_obj$grad_old * MH_obj$sigma^2/2)
+    theta_star <- as.vector(mu_old  + MH_obj$sigma * rnorm(length(MH_obj$theta)))
 
   }else{
-    mu_old <-  MH_obj$theta +  MH_obj$sigma^2/2 *( MH_obj$Sigma%*%MH_obj$grad_old)
-    theta_star <- mu_old + MH_obj$sigma * (MH_obj$L%*%rnorm(length(MH_obj$theta)))
+    mu_old <-  as.vector(MH_obj$theta +  MH_obj$sigma^2/2 *( MH_obj$Sigma%*%MH_obj$grad_old))
+    theta_star <- as.vector(mu_old + MH_obj$sigma * (MH_obj$L%*%rnorm(length(MH_obj$theta))))
   }
   res <- MH_obj$Lik(theta_star,...)
   lik_star <- res$loglik

### 2020-05-23 src/MultinomialBetaMCMC_with_GP.R

@@ -2,19 +2,19 @@
 # fitting maximum likelhiood multinomial using logit p
 # then profile likelihood of N
 ##
-set.seed(2)
 library(data.table)
 source(file.path("src", "util.r"))
 source(file.path("src", "MH.R"))
 source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
 source(file.path("src", "modelbenchmarkutil.R"))
-
+set.seed(4)
 
 path.to.files <- file.path("..","data")
+save.file = F
 MCMC_sim <- 10000
 burnin_p = 0.33
-deaths_sim <- 10
+deaths_sim <- 5
 maxusage.day = 20 #must be less then N
 unique.days  = 7
 true.day = 5
@@ -29,7 +29,8 @@ data_ <- newDeaths(deaths_est,
                    Reported,
                    maxusage.day =maxusage.day)
 
-res_save<-benchmark_BetaGP_j(N,
+day <- 16
+res_save<-benchmark_BetaGP_j(day,
                              result,true.day,
                              maxusage.day,
                              unique.days,
@@ -40,8 +41,16 @@ res_save<-benchmark_BetaGP_j(N,
 
 CI <-apply(res_save$Death_est,2 , function(x){ quantile(x,c(0.05,0.5,0.95))})
 
-out <- data.table(date = result$dates,
-                      ci_upper  = CI[1,],
-                      ci_lower  = CI[3,],
+out <- data.table(date = result$dates[1:day],
+                      ci_upper  = CI[3,],
+                      ci_lower  = CI[1,],
                       predicted_deaths = CI[2,])
-write_fst(out, file.path("data", "processed", "model_predict.fst"))
+plot(out$date, out$predicted_deaths, type='l', ylim = c(0,max(out$ci_upper)),col='blue')
+points(out$date, result$detected[1:day,day])
+points(out$date, deaths_est[1:day], pch=3)
+lines(out$date, out$ci_upper,  ylim = c(0,max(out$ci_upper)),col='red')
+lines(out$date, out$ci_lower, ylim = c(0,max(out$ci_upper)),col='red')
+if(save.file)
+    write_fst(out, file.path("data", "processed", "model_predict.fst"))
+
+

### 2020-05-23 src/modelbenchmarkutil_old.R

@@ -55,10 +55,6 @@ benchmark_BetaGP_j <- function(j,
     ##
     # seting up MCMC GP
     ##
-
-    phi.samples <- 10
-    phi <- 100
-    alpha.phi <- 10
     Start_theta <- log(deaths_est+1)
     tau <- 0.1
     MH_obj_GP <- MH_setup()
@@ -87,11 +83,12 @@ benchmark_BetaGP_j <- function(j,
         MH_obj_tau$theta <- c(tau, 1)
         Gp.prior <- matrix(NA, nrow=MCMC_sim, ncol=2)
     }
-    MH_obj_GP$Lik <- function(x, N, L, phi) {
+    MH_obj_GP$Lik <- function(x, N, L) {
         prior_list <- prior_GP(x, L)
-        lik_list   <- lik_negbin_theta(N, x, phi)
+        lik_list   <- lik_poisson(N, x)
         return(list(loglik = lik_list$loglik + prior_list$loglik,
-                    grad = as.vector(lik_list$grad  + prior_list$grad)))
+                    grad = lik_list$grad  + prior_list$grad,
+                    Hessian = lik_list$Hessian + prior_list$Hessian))
     }
 
     ##
@@ -104,7 +101,6 @@ benchmark_BetaGP_j <- function(j,
     Death_est <- matrix(NA, nrow=MCMC_sim, ncol=N)
     alpha.MCMC <- rep(4, N)
     pred_set <-array(NA, dim = c(j,N_T-j,MCMC_sim))
-    phis <- rep(0, MCMC_sim)
     p <- dim(X)[2]
     Alpha <- matrix(NA, ncol=N, nrow=N)
     Beta <- matrix(NA, ncol=N, nrow=N)
@@ -128,8 +124,8 @@ benchmark_BetaGP_j <- function(j,
         beta_2 <- MH_obj$theta[(p+1):(2*p)]
         Alpha[upper.tri(data_$report.new,diag=T)] <- exp(X%*%beta_1)
         Beta[upper.tri(data_$report.new,diag=T)]  <-  exp(X%*%beta_2)
-        #prior_N <- function(N, i){ dnegbin(N, exp(MH_obj_GP$theta[i]) - lgamma(N+1)}
-        prior_N <- function(N, i){ dnegbin(N, exp(MH_obj_GP$theta[i]),phi) }
+
+        prior_N <- function(N, i){ N *MH_obj_GP$theta[i] - lgamma(N+1)}
         res <- sample.deathsBB(deaths_sim,
[…]

### 2020-05-23 src/time_series_experiment.R

@@ -1,0 +1,159 @@
+##
+# simple stan code for data
+#
+##
+library(data.table)
+library(rstan)
+source(file.path("src", "GPutil.R"))
+source(file.path("src", "MH.R"))
+sim <- 2000
+rstan_options(auto_write = TRUE)
+result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+
+Reported = result$detected
+N <- dim(Reported)[1]
+
+deaths_est <- apply(Reported, 1, max, na.rm=T)
+
+death_est <- deaths_est[1:(N-15)]
+
+N <- length(death_est)
+data.list <- list(N=N,
+                  Y = death_est)
+
+plot(death_est,xlab='days',ylab='deaths')
+tau <- 1000
+for(i in 15 ){
+    data.list_temp <- list(N=i, Y = death_est[1:i])
+    stan_out = stan(file = "src/stan/RW1.stan",
+                    data= data.list_temp,
+                    iter = sim,
+                    warmup = ceiling(0.5*sim),
+                    chains =1)
+    Theta_stan <- extract(stan_out)$theta
+    #tau        <- extract(stan_out)$tau
+
+
+    CI <- apply(exp(Theta_stan),2,quantile,c(0.1,0.5,0.9))
+    lines(CI[2,],col='blue')
+    lines(CI[1,],col='red')
+    lines(CI[3,],col='red')
+    stan_out = stan(file = "src/stan/RW1_negbin.stan",
+                    data= data.list_temp,
+                    iter = sim,
+                    warmup = ceiling(0.5*sim),
+                    chains =1)
+    Theta_stan <- extract(stan_out)$theta
+    #tau        <- extract(stan_out)$tau
+
+    CI <- apply(exp(Theta_stan),2,quantile,c(0.1,0.5,0.9))
+    lines(CI[2,],col='blue',lty=2)
+    lines(CI[1,],col='red',lty=2)
+    lines(CI[3,],col='red',lty=2)
+
+    Start_theta <- log(data.list_temp$Y)
+    ###
+    # poisson part
+    #
+    #
+    ###
+    MH_obj_GP <- MH_setup()
+    MH_obj_GP$sigma <- 0.01
+
+    n_theta <- length(Start_theta)
+    L <- toeplitz(c(-1,1, rep(0,n_theta-2)))
+    L[lower.tri(L,diag=F)] <- 0
[…]

### 2020-05-23 src/util.r

@@ -71,11 +71,11 @@ sample.deathsBB <- function(samples, deaths, alpha, beta, Reported, alpha.MCMC, 
         lik_i <- lik_i + prior(deaths[i], i)
       for(j in 1:samples){
         death_star <- sample((deaths[i]-alpha.MCMC[i]):(deaths[i]+alpha.MCMC[i]), 1)
-
         lik_star <- loglikDeathsGivenProbBB(death_star, alpha_i , beta_i, Reported_i)
         if(use.prior==T)
           lik_star <- lik_star + prior(death_star, i)
-
+        if(is.nan(lik_star))
+          next
         if(log(runif(1)) < lik_star-lik_i){
           lik_i = lik_star
           deaths[i] <- death_star

### 2020-05-23 src/Beta_GP_benchmark_build.R
leaving beta-bin for binomial

@@ -8,9 +8,9 @@ source(file.path("src", "util.r"))
 source(file.path("src", "MH.R"))
 source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
-source(file.path("src", "modelbenchmarkutil.R"))
+source(file.path("src", "modelbenchmarkutil_binomial.R"))
 nclust <- 6
-MCMC_sim <- 20000
+MCMC_sim <- 10000
 burnin_p = 0.3
 deaths_sim <- 5
 maxusage.day = 20 #must be less then N
@@ -35,14 +35,14 @@ cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
 doParallel::registerDoParallel(cl)
 foreach(j = start.predict.day:N_T)  %dopar% {
     result <- readRDS(file.path("data", "processed", "processed_data.rds"))
-  res_save<-benchmark_BetaGP_j(j,
+  res_save<-benchmark_BinomGP_j(j,
                                result,true.day,
                                maxusage.day,
                                unique.days,
                                MCMC_sim,
                                burnin_p,
                                deaths_sim = deaths_sim,
-                               prior = c(0,0))
+                               prior = c(1,0))
 
   save(res_save,
        file = file.path("data", "simulation_results", paste0("param_", result$dates_report[j], ".rds")))

### 2020-05-23 src/modelbenchmarkutil_binomial.R

@@ -1,0 +1,193 @@
+
+##
+#' running the benchmark for day j
+#'
+#' j - day j
+#' result -
+#' true.day  - (int) days not estimated
+#' maxusage.day - (int) how many reports days to use for each death
+#' unique.days  - (int) how many unique reports days to have as parameters
+#' MCMC_sim     - (int) how many simulations
+#' burnin_p     - ([0,1]) how many burnins to run in percantage of MCMC_sim
+#' prior        - (2 x 1) [1] 0: rw1, 1: rw2,
+#'                        [2] 0: no noise, 1: noise
+##
+benchmark_BinomGP_j <- function(j,
+                               result,
+                               true.day,
+                               maxusage.day,
+                               unique.days,
+                               MCMC_sim,
+                               burnin_p,
+                               deaths_sim,
+                               prior = c(0,0)){
+    require(Matrix)
+    N_T <- dim(result$detected)[1]
+    deaths_est_T <- apply(result$detected, 1, max, na.rm=T)
+    Reported <- result$detected[1:j,1:j]
+    dates_report <- result$dates_report[1:j]
+    N <- dim(Reported)[1]
+    deaths_est <- apply(Reported, 1, max, na.rm=T)
+    deaths_est[1:true.day] = deaths_est_T[1:true.day] #days with known deaths
+
+
+    data_ <- newDeaths(deaths_est,
+                       Reported,
+                       maxusage.day =maxusage.day)
+
+
+    ##
+    # building covariate matrix
+    ##
+    X <- setup_data(N, maxusage.day,  dates_report, unique.days)
+
+    ##
+    # seting up MCMC
+    ##
+    MH_obj <- MH_setup()
+    MH_obj$sigma <- 0.1
+    MH_obj$theta <- rep(0,dim(X)[2])
+
+
+    MH_obj$Lik <- loglikProb
+
+
+    ##
+    # seting up MCMC GP
+    ##
+
+    phi.samples <- 10
+    phi <- 100
+    alpha.phi <- 10
+    Start_theta <- log(deaths_est+1)
+    tau <- 0.1
+    MH_obj_GP <- MH_setup()
+    MH_obj_GP$sigma <- 0.1
+
+    MH_obj_GP$theta <- Start_theta
+    if(prior[1] == 0){
+        n_theta <- length(Start_theta)
[…]

### 2020-05-23 src/util.r

@@ -11,7 +11,9 @@ require(Matrix)
 #  alpha    - (N x 1) stepsizes
 #  true.dat  - (int) how many days after recording should one sample
 ##
-sample.deaths <- function(samples, deaths, P, Reported, alpha, true.day = 0){
+sample.deaths <- function(samples, deaths, P, Reported, alpha, true.day = 0,
+                          use.prior=FALSE,
+                          prior=NULL){
 
   N <- length(deaths)
   acc <- rep(0,N)
@@ -23,11 +25,18 @@ sample.deaths <- function(samples, deaths, P, Reported, alpha, true.day = 0){
       P_i         = P_i[index]
       Reported_i  = Reported_i[index]
       lik_i <- loglikDeathsGivenProb(deaths[i], P_i, Reported_i)
+      if(use.prior==T)
+        lik_i <- lik_i + prior(deaths[i], i)
       for(j in 1:samples){
         death_star <- sample((deaths[i]-alpha[i]):(deaths[i]+alpha[i]), 1)
-        lik_start <- loglikDeathsGivenProb(death_star,  P_i, Reported_i)
-        if(log(runif(1)) < lik_start-lik_i){
-          lik_i = lik_start
+        lik_star <- loglikDeathsGivenProb(death_star,  P_i, Reported_i)
+        if(use.prior==T)
+          lik_star <- lik_star + prior(death_star, i)
+        if(is.nan(lik_star))
+          next
+
+        if(log(runif(1)) < lik_star-lik_i){
+          lik_i = lik_star
           deaths[i] <- death_star
           acc[i] = acc[i] + 1
         }

### 2020-05-23 src/Beta_GP_benchmark.R
update

@@ -6,7 +6,7 @@ library(foreach)
 library(fst)
 library(data.table)
 source("src/util.r")
-nclust <- 1
+nclust <- 6
 maxusage.day <- 20
 path.to.files <- file.path("data")
 files <- list.files(paste(path.to.files,"/simulation_results",sep=""),

### 2020-05-23 src/plots.R
added 23/5

@@ -228,7 +228,7 @@ plot <- ggplot(data = plot_data, aes(x = dayofweek, y = V1, color = type, group 
          subtitle = "",
          caption = "",
          color = "Model",
-         x = "Last date included in the model",
+         x = "Weekday",
          y = "")
 
 ggsave(filename = file.path("output", "plots", "SCRPS_over_weekdays.pdf"),

### 2020-05-23 src/plots.R
duplicate targets added warning

@@ -79,13 +79,17 @@ day_plot <- function(DT, reported, plot.title) {
     colors <- c("#ECCBAE", "#046C9A", "gray50")
     colors <- setNames(colors, c(levels(DT$type), "Reported"))
 
+    if (any(DT[!is.na(target), uniqueN(target) != 1, by = date][, V1])) {
+        warning("Multiple unique target values for same day.")
+    }
+    hline <- DT[!is.na(target), .(unique(target)), by = .(date)]
+
     plot <- ggplot(data = DT,
                 aes(x = days_left,
                     y = predicted_deaths,
                     color = type,
                     group = type)) +
-        geom_hline(data = DT[!is.na(target), .(unique(target)), by = .(date)],
-                aes(yintercept = V1), color = "grey50") +
+        geom_hline(data = hline, aes(yintercept = V1), color = "grey50") +
         geom_line(data = reported,
                 aes(y = reported_dead, group = "Reported", color = "Reported"), linetype = "dashed") +
         geom_point(data = reported,
@@ -102,7 +106,7 @@ day_plot <- function(DT, reported, plot.title) {
         # scale_fill_manual(values = fill_colors, limits = label_order, drop = FALSE) +
         # scale_color_manual(values = wes_palette("Darjeeling2")) +
         scale_color_manual(values = colors) +
-        scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(0, 5))) +
+        scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(1, 5))) +
         labs(title = plot.title,
             subtitle = "",
             caption = "",

### 2020-05-23 src/functions.R
removed margins

@@ -81,9 +81,9 @@ set_default_theme <- function() {
     theme_ipsum(base_family = "EB Garamond") %+replace%
         theme(
             text = element_text(size = 12, color = "#333333", family = "EB Garamond"),
-            plot.title = element_text(size = rel(2), face = "plain", hjust = 0, margin = margin(0,0,5,0)),
-            plot.subtitle = element_text(size = rel(1), face = "plain", hjust = 0, margin = margin(0,0,0,0)),
-            plot.caption = element_text(size = rel(0.7), family = "EB Garamond", face = "italic", hjust = 1, vjust = 1, margin = margin(12,0,0,0)),
+            #plot.title = element_text(size = rel(2), face = "plain", hjust = 0, margin = margin(0,0,5,0)),
+            #plot.subtitle = element_text(size = rel(1), face = "plain", hjust = 0, margin = margin(0,0,0,0)),
+            #plot.caption = element_text(size = rel(0.7), family = "EB Garamond", face = "italic", hjust = 1, vjust = 1, margin = margin(12,0,0,0)),
 
             legend.text = element_text(size = rel(0.9), family = "EB Garamond", hjust = 0, margin = margin(0, 0, 0, 0)),
             legend.background = element_rect(fill = "grey95"),
@@ -99,6 +99,7 @@ set_default_theme <- function() {
             # Panels
             plot.background = element_rect(fill = "#FFFFFF", color = NA), # bg of the plot
             # plot.background = element_rect(fill = "#f5f5f5", color = NA), # bg of the plot
+            plot.margin = unit(c(0, 0, 0, 0), "pt"),
             panel.border = element_blank(),
             panel.grid.major = element_line(linetype = "dotted", color = "#CCCCCC", size = 0.3),
             panel.grid.minor = element_line(linetype = "dotted", color = "#CECECE", size = 0.2),

### 2020-05-23 src/plots.R

@@ -34,12 +34,12 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
     theme(axis.text.x = element_text(angle = 35, hjust = 1.3, vjust = 1.1)) +
     scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(0, 5))) +
     labs(#title = paste0("Reported deaths as of ", deaths_dt[, max(publication_date)], " and model prediction"),
-        subtitle = "",
-        caption = "",
-        fill = "",
-        linetype = "",
-        x = "Death date",
-        y = "Number of deaths")
+         #subtitle = "",
+         #caption = "",
+         fill = "",
+         linetype = "",
+         x = "Death date",
+         y = "Number of deaths")
 
 ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
@@ -108,8 +108,8 @@ day_plot <- function(DT, reported, plot.title) {
         scale_color_manual(values = colors) +
         scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(1, 5))) +
         labs(title = plot.title,
-            subtitle = "",
-            caption = "",
+             #subtitle = "",
+             #caption = "",
             color = "Model",
             x = "Days of lag to predict",
             y = "Number of deaths")
@@ -173,8 +173,8 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
     set_default_theme() +
     scale_color_manual(values = wes_palette("Darjeeling2")) +
     labs(#title = "Model metrics",
-         subtitle = "",
-         caption = "",
+         #subtitle = "",
+         #caption = "",
          color = "Model",
          x = "Days of lag to predict",
          y = "")
@@ -199,8 +199,8 @@ plot <- ggplot(data = plot_data, aes(x = state, y = V1, color = type, group = ty
     set_default_theme() +
     scale_color_manual(values = wes_palette("Darjeeling2")) +
     labs(#title = "Model metrics",
-         subtitle = "",
-         caption = "",
+         #subtitle = "",
+         #caption = "",
          color = "Model",
          x = "Last date included in the model",
          y = "SCRPS")
@@ -229,8 +229,8 @@ plot <- ggplot(data = plot_data, aes(x = dayofweek, y = V1, color = type, group 
[…]

### 2020-05-23 src/functions.R
need at least some margin

@@ -99,7 +99,7 @@ set_default_theme <- function() {
             # Panels
             plot.background = element_rect(fill = "#FFFFFF", color = NA), # bg of the plot
             # plot.background = element_rect(fill = "#f5f5f5", color = NA), # bg of the plot
-            plot.margin = unit(c(0, 0, 0, 0), "pt"),
+            plot.margin = unit(c(2, 2, 2, 2), "pt"),
             panel.border = element_blank(),
             panel.grid.major = element_line(linetype = "dotted", color = "#CCCCCC", size = 0.3),
             panel.grid.minor = element_line(linetype = "dotted", color = "#CECECE", size = 0.2),

### 2020-06-06 src/plots.R
eligible warning

@@ -81,7 +81,7 @@ day_plot <- function(DT, reported, plot.title) {
     colors <- setNames(colors, c(levels(DT$type), "Reported"))
 
     if (any(DT[!is.na(target), uniqueN(target) != 1, by = date][, V1])) {
-        warning("Multiple unique target values for same day.")
+        warning("Multiple unique target values for same day: ", plot.title)
     }
     hline <- DT[!is.na(target), .(unique(target)), by = .(date)]
 
@@ -143,7 +143,7 @@ for (i in seq_along(dates)) {
                      reported_dead[date == dates[i]],
                      plot.title = dates[i])
     ggsave(filename = file.path("output", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
-       plot = plot, device = cairo_pdf, width = w, height = w)
+           plot = plot, device = cairo_pdf, width = w, height = w)
 }
 #
 ## PLOT 2: Statistics ##

### 2020-06-06 output/paper/appendix.tex
Updates from Overleaf

@@
 \subsection{Notation}
-Before presenting the model we describe some notation used through out the article for a $m \times n$ matrix $r$ we use the following broadcasting notation $\mv{r}_{k,j:l}=[ r_{k,j}, r_{k,j+1}, \ldots, r_{k,l}]$.
+Before presenting the model we describe some notation used through out the appendix.
+For a $m \times n$ matrix $r$ we use the following broadcasting notation $\mv{r}_{k,j:l}=[ r_{k,j}, r_{k,j+1}, \ldots, r_{k,l}]$.
 Further $x | y \sim \pi(.)$ implies that the random variable $x$ if we conditioning on $y$ follows distribution $\pi(.)$.
@@
 the density of the observations given the parameters.
-Here the data consist the daily report of recored deaths for the past days.
+Here the data consist the daily report of recorded deaths for the past days.
 This can conveniently be represented upper triangular matrix, $\mv{r}$, where $r_{i,j}$ represents number of new reported deaths for day $i$ reported at day $j$.
@@
 
-Typically in removal sampling one let $p_{i,j}:=p$ however for this data this is clearly not realistic-- given delaying in the reporting.
+Typically in removal sampling one would set the probability of reporting uniform, i.e.
+$p_{i,j}:=p$.
+However for this data this is clearly not realistic given weekly patterns in reporting --very little reporting during the weekends.
 Instead we assume that we have $k$ different probabilities.
-Further to account for overdispertion we assume that each probability instead follows a Beta distribution.
-The Beta distibution has two parameters $\alpha$ and $\beta$.
+Further, to account for overdispertion, we assume that each probability rather being a fixed scalar is a random variable with a Beta distribution.
+The Beta distribution has two parameters $\alpha$ and $\beta$.
 This resulting the following distribution for the probabilities
@@
 $$
-Here, we let $H$ denote holidays and weekends and the parameters above are
+Here, if $j\in H$ then day $j$ is a holidays or weekends, and the parameters above are
 $$
@@
 This model is created to create a temporal smoothing between the reported deaths.
-For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution.
-
+For the hyperparameter $\sigma^2$ we impose a inverse Gamma distribution, this prior is suitable here because it guarantees that the process is not constant ($\sigma^2=0$) which we know is not the case.
 \subsection{Full model}
@@
 	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution.
-Here to we use an adaptive MALA \cite{Atchade2006_adaptive_version} to sample from these parameters.
+Here to we use an adaptive MALA (\cite{Atchade2006_adaptive_version}) to sample from these parameters.
 	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
@@
 \section{Model Benchmark}
-In this section, we present additional comparison of the model to a simple constant model. The benchmark model is the sum of average reporting lag for the preceding 14 days.
+In this section, we present additional comparison of the model to the benchmark. We first describe the benchmark model in detail.
+
+The benchmark model simply takes the sum of average historical reporting lags for the preceding 14 days.
+As before $r_{ij}$ is the number of deaths that happened on day $i$ and were recorded on day $j$.
+To predict the number of people that died on a given day, we first calculate lag averages:
+
+\begin{align}
+    \hat{r}_{i, i+L} = \frac{\sum^{14}_{k=i-14} r_{k - L, k}}{14},
+\end{align}
+
+where $\hat{r}_{i, i+L}$ is the average number of deaths reported with a lag of $L$ days, based on the 14 reports closest preceding day $i$.
+If we are looking at data released $2020-04-28$ and call this day 0, the latest death date that we have 10-day ($L=10$) reporting lag observation for is $r_{-10,0}$.
+The average for $Lag(0, 10)$ is therefore taken over the 14 days between $r_{-24,-14}$ and $r_{-10,0}$ (2020-04-04 and 2020-04-18).
+For this reason, some of the earlier predictions will not have data from $14$ days.
+The average is then taken over all available reports.
+
+In the comparisons we aim at predicting the total number of deaths that will have been reported within 14 days of the death date.
+To do so, we sum over the average lag that has yet to be reported.
+If we are predicting the number of people that have yet to be reported dead for day -3, we already know the true values for $r_{-3,-3}$, $r_{-3,-2}$, $r_{-3,-1}$, and $r_{-3,0}$ so we only need to predict $r_{-3,1}\ldots r_{-3,10}$.
+The prediction is then
+
+\begin{align}
+    Benchmark(i, j) = \sum_{l=i}^{j} r_{i,l}+ \sum_{l = j}^{14} \hat{r}_{i, l}.
+\end{align}
+
+As confidence interval we simply use a Normal assumption with standard deviations of the reporting lags, assuming independence, i.e. this is just the square root of the sum of $Var(\hat{r})$.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_states}
+    \caption{}
+    \label{fig:SCRPS_states}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_weekdays}
+    \caption{Model accuracy over time for four dates, compared to a benchmark}
+    \label{fig:SCRPS_weekdays}
+\end{figure}
 

### 2020-06-06 output/paper/paper.tex
Updates from Overleaf

@@
 
-The death counts suffer from the longest reporting day.
+The death counts suffer from the longest reporting delay.
 In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trend line 10 days before the latest date.
@@
 The removal method that has three major advantages over simply reporting moving averages:
+
 \begin{itemize}
-	\item it does not relay previous trend in the data,
-	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day,
+	\item it does not relay previous trend in the data;
+	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day;
 	\item the uncertainty in the estimate can be carried over to epidemiological models that uses the estimate as input, and hence give more realistic models.
@@
 This pattern of declining number of trapped animals allows one to draw inference of the underlying population size.
-Here we replace the animal population with the true number of deaths or cases in a given day.
+Here we replace the animal population with the true number of deaths, or cases, on a given day.
 Instead of traps we have the new reports of COVID--19 events.
@@
 
-In Figure 1 we apply the model to the latest statistics from Sweden.
+In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden.
 The graph shows reported and predicted deaths (with uncertainty intervals) as bars and a dashed line plots the 7-day (centered) moving average.
-
-In Figure 1 we illustrate the similarity and difference between the 7-day moving average and the new Bayesian prediction model with 95\% prediction intervals to the reported number of deaths from COVID-19 in Sweden.
-The model provides estimates of actual deaths considerably above the reported number of deaths and the uncertainty of the estimate.
+A version without predictions is used in the Public Health Agency´s daily press briefings.
+As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
 
@@
     \centering
-    \includegraphics{../plots/}
-    \caption{Caption}
-    \label{fig:my_label}
+    \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
+    \caption{Swedish Covid-19 deaths as of 2020-05-22 and model predictions}
+    \label{fig:latest_prediction}
 \end{figure}
 
-% TODO: FIGURE 1 HERE
-%Figure 1.
-The 7-day moving average (blue) and the prediction from the Bayesian model (black) with prediction intervals overlaid.
-The grey bars represent the number of reported deaths up to May the 12th, 2020, in COVID-19 in Sweden.
+\chapter{Model Performance}
 
-%In Figure 2 we use the method to predict the number of new deaths that will be reported in the most recent report and how their date of death will be distributed over time.
-It is clear from the model prediction that the delay is substantial and therefore interpretations of the daily reported number of deaths without these models to adjust it are difficult.
+Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to a benchmark prediction based on historical averages.
+The model and the benchmark are tasked with predicting the total number of individuals who have died at a given date and have been reported within 14 days of that date.
+As time progresses, more deaths are reported and the dashed grey line approaches the horizontal line.
+Meanwhile model uncertainty decreases.
 
-% TODO: FIGURE 2 here
-%Figure 2. The bars represent deaths in COVID-19 in Sweden and the prediction intervals represent the Bayesian predictions of the number of deaths for the specific day in the latest report.
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{../plots/lag_prediction_by_date}
+    \caption{Model accuracy over time for four dates, compared to benchmark}
+    \label{fig:four_dates}
+\end{figure}
+
+Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics.
+All three graphs are based on predictions of reported deaths within 14 days, and show how performance increases as more data has been reported.
+Each data point is the average of all dates where predictions can be evaluated.
+SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix).
+The central plot shows the width of the confidence intervals, and the rightmost one the proportion of CI´s that cover the true value.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=0.9\textwidth]{../plots/model_metrics}
+    \caption{Model accuracy over time for four dates, compared to a benchmark}
+    \label{fig:model_metrics}
+\end{figure}
 
@@
 Unfortunately we do not currently have access to the high resolution data required to do so.
-Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or mis classification of days.
+Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days.
 The number of such cases is very small, however, and its removal should not make much difference.
@@
 Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages.
+When the covariates to the reporting delay pattern is known, the model can incorporate them and provide a more accurate nowcasting of COVID-19 events.
 

### 2020-06-06 src/plots.R
small plot update

@@ -13,7 +13,7 @@ w <- 11 # plot width (inches)
 deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
 model_predict <- read_fst(file.path("data", "processed", "model_latest.fst"), as.data.table = TRUE)
 DT1 <- model_predict[date >= Sys.Date() - 28]
-DT2 <- deaths_dt[!is.na(N) & !is.na(date) & publication_date == max(publication_date)]
+DT2 <- deaths_dt[!is.na(N) & !is.na(date) & publication_date == max(model_predict[, date])]
 DT2[, avg := frollmean(N, 7, algo = "exact", align = "center")]
 DT2 <- DT2[date > "2020-03-15"]
 

### 2020-06-06 src/plots.R
fixed  moving avg to look more like FHM

@@ -14,7 +14,7 @@ deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.t
 model_predict <- read_fst(file.path("data", "processed", "model_latest.fst"), as.data.table = TRUE)
 DT1 <- model_predict[date >= Sys.Date() - 28]
 DT2 <- deaths_dt[!is.na(N) & !is.na(date) & publication_date == max(model_predict[, date])]
-DT2[, avg := frollmean(N, 7, algo = "exact", align = "center")]
+DT2[, avg := frollmean(N, 7, algo = "exact", align = "right")]
 DT2 <- DT2[date > "2020-03-15"]
 
 colors <- c("gray80", "#ECCBAE")
@@ -24,7 +24,8 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
     geom_bar(aes(fill = "Model prediction"), stat = "identity") +
     geom_bar(data = DT2,
              aes(y = N, fill = "Reported dead"), stat = "identity") +
-    geom_line(data = DT2[!is.na(avg)], aes(y = avg, linetype = "7-day moving average"), color = "grey50") +
+    geom_line(data = DT2[!is.na(avg) & date <= max(model_predict[, date]) - 10],
+              aes(y = avg, linetype = "7-day moving average"), color = "grey50") +
     geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                   width = 0.4, color = "#e0ac7e") +
     set_default_theme() +

### 2020-06-06 output/paper/paper.tex

@@
 But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend.
-Here we describe a statistical methodology for predicting the actual daily number occurring events a specific day, and its uncertainty, based on the daily reported event frequency.
-The methodology takes into account the observed distribution pattern of the delay and can be derived from the ``removal method'', a well-established method in the field of ecology.
+Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays.
+The methodology takes into account the observed distribution pattern of the lag.
+It is derived from the ``removal method'', a well-established estimation framework in the field of ecology.
 	\end{abstract}
@@
 Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}).
-The latest data are often not finalized, but change as new statistics are reported.
-In fact, since these updates are usually in the form of delayed reporting.
-Since the most recent days then have the least cases, such systematic under reporting gives the dangerous picture of an always improving situation.
+The latest data are often not finalized, but change as new information is reported.
+In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.
 
@@
 If the pandemic is indeed slowing, we should not wait for the data to be finalized before using it.
-Rather, we argue that actual case counts and deaths should be nowcasted to account for any reporting delay and ensure policymakers are using the most accurate numbers available.
+Rather, we argue that actual case counts and deaths should be nowcasted to account for reporting delay, thus allowing policymakers to use the latest numbers availiable without beinig misled by reporting bias.
 
@@
 
-In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and the degree of uncertainty, based on the daily reported event frequency and the observed distribution pattern of the reporting delay.
-The prediction model is building on methods developed in ecology, referred to as the ``removal methods'' or ``capture-retain'' models (\cite{Pollock1991_review_papers}).
+In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays.
+The prediction model is building on methods developed in ecology, referred to as the ``removal method'' or ``capture-retain'' models (\cite{Pollock1991_review_papers}).
+
+To help motivate why such forecasting is needed, we now turn to the case of Sweden. The model is flexible by design, however, and could easily be applied to other countries as well.
 
 \section{The current situation of COVID-19 in Sweden}
-Each day at 14:00 the Swedish Public Health Agency holds a press conference where new COVID-19 statistics are presented\footnote{The data is then published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/} where we download it every day.}.
-Deaths is one of the main indicators to follow for understanding the impact of the pandemic on public health in Sweden, but also the number of new admissions to critical care, to hospitals, and the new confirmed cases are reported.
-One of the reasons for following these indicators is to enable public health professionals and the public to observe the patterns of the evolving, flattened or suppressed epidemic (\cite{Anderson2020_how_will}).
-In relation to policy, it is of further interest to understand if growth rates change, which could indicate a potential response.
-However, at the daily presentation only a proportion of the number deaths for each of the most recent days is yet known, and this bias causes an artificial, downward, trend in the data.
+The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}.
+During a press conference, they present updates on the number of deaths, admissions to hosiptals and intensive care, as well as case counts.
+
+One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic (\cite{Anderson2020_how_will}).
+In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response.
+However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces the illusion of a downward trend.
 
 The death counts suffer from the longest reporting delay.
-In their daily presentation, the Swedish Public Health Agency warns for this by stopping the 7-day moving average trend line 10 days before the latest date.
-But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information from the most recent days creating a sense of a downward trend.
-In fact, this might be the reason why the number of deaths has been underestimated repeatedly.
-At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual level was closer to 100 (\cite{Ohman2020_antalet_virusdoda}).
+In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date.
+But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a sense of a downward trend.
+In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly.
+At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 (\cite{Ohman2020_antalet_virusdoda}).
 
@@
 \begin{itemize}
-	\item it does not relay previous trend in the data;
-	\item we can generate confidence bounds for what is the reasonable range of the uncertainty in the event frequency at a given day;
-	\item the uncertainty in the estimate can be carried over to epidemiological models that uses the estimate as input, and hence give more realistic models.
+	\item it does not relay any previous trend in the data,
+	\item we can generate confidence bounds for the uncertainty about daily true frequencies,
+	\item the uncertainty estimates can be carried over to epidemiological models to help create more realistic models.
 \end{itemize}
@@
 This pattern of declining number of trapped animals allows one to draw inference of the underlying population size.
-Here we replace the animal population with the true number of deaths, or cases, on a given day.
+Here we replace the animal population with the true number of deaths
+on a given day.
 Instead of traps we have the new reports of COVID--19 events.
 As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day.
-In fact, if we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.
+If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.
 
@@
 
+To get accurate estimates we apply two institution-specific corrections.
+First, we only count workdays as constituting reporting delay, as very few deaths are reported during weekends.
+Second, we apply a constant bias correction to account for the fact that Swedish deaths come from two distinct populations with different trends: deaths in hospitals, and in elderly care.
+
 In Figure~\ref{fig:latest_prediction} we apply the model to the latest statistics from Sweden.
-The graph shows reported and predicted deaths (with uncertainty intervals) as bars and a dashed line plots the 7-day (centered) moving average.
+The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average.
 A version without predictions is used in the Public Health Agency´s daily press briefings.
@@
     \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
-    \caption{Swedish Covid-19 deaths as of 2020-05-22 and model predictions}
+    \caption{Swedish Covid-19 deaths as of 2020-06-04 and model predictions}
     \label{fig:latest_prediction}

### 2020-06-11 src/Beta_GP_benchmark_refit3.R
def merge

@@ -217,6 +217,7 @@ model_benchmark_bias$days_left =  model_benchmark_bias$date -model_benchmark_bia
 
 names(model_benchmark) <- c('state','date','days_left','target','SCRPS','predicted_deaths','ci_lower','ci_upper')
 model_benchmark<-model_benchmark[,c('state','date','target','days_left','ci_upper','ci_lower','predicted_deaths','SCRPS')]
+model_benchmark$days_left =  model_benchmark$date -model_benchmark$state +14
 
 write_fst(model_benchmark, file.path("data", "processed", "model_benchmark.fst"))
 write_fst(model_benchmark_bias, file.path("data", "processed", "model_benchmark.fst"))

### 2020-06-11 src/MH.R

@@ -175,7 +175,6 @@ MH_LCN_Hessian <- function(MH_obj,...)
     MH_obj$R_old   = R
     MH_obj$accept  = 1
     MH_obj$grad_old <- res$grad
-    MH_obj$accept_count <- MH_obj$accept_count  +1
 
     return(MH_obj)
   }
@@ -225,7 +224,6 @@ MH_MALA_Hessian <- function(MH_obj,...)
     MH_obj$L_old   = t(R)
     MH_obj$accept  = 1
     MH_obj$grad_old <- res$grad
-    MH_obj$accept_count <- MH_obj$accept_count  +1
 
     return(MH_obj)
   }
@@ -282,10 +280,8 @@ MH_MALA <- function(MH_obj,...)
     MH_obj$lik_old = lik_star
     MH_obj$accept  = 1
     MH_obj$grad_old <- res$grad
-    MH_obj$accept_count <- MH_obj$accept_count  +1
     MH_obj$res <- res
 
-    return(MH_obj)
   }
   return(MH_obj)
 }
@@ -320,14 +316,20 @@ MH_SA <- function(MH_obj, option = 1)
       MH_obj$Sigma  <- (MH_obj$thetaM- MH_obj$theta)%*%t(MH_obj$thetaM- MH_obj$theta)
     }
     MH_obj$n_SA  <- MH_obj$n_SA + 1
+
     if(MH_obj$n_SA  == 20 * nrow(MH_obj$Sigma)){
+      #MH_obj$L <- t(chol(MH_obj$Sigma + 0.01*min(diag(MH_obj$Sigma)) * diag(dim(MH_obj$Sigma)[1])))
       MH_obj$L <- t(chol(MH_obj$Sigma))
       MH_obj$sigma <- MH_obj$sigma/sqrt(max(diag(MH_obj$Sigma)))
+
     }
-    if(MH_obj$n_SA > 20 * nrow(MH_obj$Sigma))
-      MH_obj$L <- t(chol(MH_obj$Sigma))
+    if(MH_obj$count %% 50 == 0){
+      if(MH_obj$n_SA > 20 * nrow(MH_obj$Sigma)){
+        #MH_obj$L <- t(chol(MH_obj$Sigma + 0.01*min(diag(MH_obj$Sigma)) * diag(dim(MH_obj$Sigma)[1])))
+        MH_obj$L <- t(chol(MH_obj$Sigma ))
+      }
     }
-  }
+  }}
   if( option == 2)
   {
       if(MH_obj$count %% 50  == 0){

### 2020-06-11 src/lagged/Beta_GP_lag_benchmark_build.R

@@ -1,0 +1,88 @@
+##
+#  building the benchmark prediction parameters for the full model
+#  need to run Beta_GP_benchmark.R
+##
+library(foreach)
+library(doParallel)
+library(tidyr)
+source(file.path("src", "util.r"))
+source(file.path("src", "MH.R"))
+source(file.path("src", "functions.R"))
+source(file.path("src", "GPutil.R"))
+source(file.path("src","lagged" ,"MLbeta.R"))
+nclust <- 6
+lag <- 1
+MCMC_sim <- 15000
+burnin_p = 0.3
+deaths_sim <- 5
+true.day <- 5
+maxusage.day <- 20
+start.predict.day = 20#16 # more then unique days
+
+predicition.list <-list()
+result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+
+cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
+doParallel::registerDoParallel(cl)
+foreach(j = start.predict.day:dim(result$detected)[1], .errorhandling='remove')  %dopar% {
+    library(rGIG)
+    cat('j=',j,", ", file=stdout())
+    result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+    res_save_post_lag<-benchmark_BetaGP_post_lag_j(j,
+                                                   result,
+                                                   lag,
+                                                   true.day,
+                                                   maxusage.day,
+                                                   MCMC_sim,
+                                                   burnin_p,
+                                                   deaths_sim,
+                                                   prior = 0,
+                                                   npars = 2)
+  res_save_lag<-benchmark_BetaGP_lag_j(j,
+                                      result = result,
+                                      lag   = lag,
+                                      MCMC_sim = MCMC_sim,
+                                      burnin_p = burnin_p,
+                                      deaths_sim = deaths_sim,
+                                      prior = 0,
+                                      npar = 2)
+  save(res_save_lag,res_save_post_lag,
[…]

### 2020-06-11 src/lagged/Beta_GP_lag_benchmark_refit.R

@@ -1,0 +1,232 @@
+###
+# benchmarking
+#
+###
+library( stringr)
+library(foreach)
+library(tidyr)
+library(fst)
+library(data.table)
+source("src/util.r")
+nclust <- 6
+report.day = 14 # how long after death we wish to predicit
+start.predict.day=20
+path.to.files <- file.path("data")
+#files <- list.files('/Users/jonaswallin/Dropbox/temp/simulation_results_beta',
+#                    pattern = "^param",
+#                    full.names = TRUE)
+files <- list.files(paste(path.to.files,"/simulation_results",sep=""),
+                             pattern = "^param",
+                             full.names = TRUE)
+
+index_files <- 1:(length(files)-1)
+result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+
+
+cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
+doParallel::registerDoParallel(cl)
+
+#for(i in index_files){
+sim.data <- foreach(i = index_files, .combine='rbind')  %dopar% {
+  library(Matrix)
+  library(tidyr)
+  library(fst)
+  library(data.table)
+  library(stringr)
+  load(files[i]) #res_save_lag
+
+  cat('i=',i,", ", file=stdout())
+  Date <- as.Date(result$dates_report[dim(res_save_lag$Death_est)[2]])
+  cat('file= ',format(Date) , '\n',file=stdout())
+
+  result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+
+  pred_time = which(result$dates_report==str_extract(files[i],'[0-9]{4}-[0-9]{2}-[0-9]{2}'))
+  cat('pred_time=',pred_time,"\n", file=stdout())
+  N <- dim(result$detected)[1]
+  lag <- res_save_lag$lag
+  report <- splitlag(result$detected,as.Date(result$dates_report) ,res_save_lag$lag)
+  report_dates <-c(result$dates_report[1:pred_time],
+                   result$dates_report[pred_time] + 1:report.day)
+  X_T <- setup_data_lag(pred_time +report.day, res_save_lag$npar, report_dates, res_save_lag$npar)
+
+  X_O <- setup_data_postlag2(pred_time +report.day,
+                             res_save_lag$lag,
+                             res_save_post_lag$npars,
[…]

### 2020-06-11 src/lagged/MLbeta.R

@@ -1,0 +1,518 @@
+
+source("src/util.r")
+##
+#'
+#'    lag    - (int)   lag days  (lag = 0 is only first non-holiday, lag = 1 two non holidays)
+#'    theta0 - (p x 1) inital guess of parameters
+##
+ML_betaBin <- function(result,
+                       lag,
+                       npar = 2,
+                       theta0 = NULL){
+
+    report <- splitlag(result$detected,as.Date(result$dates_report) ,lag)
+    report_lag <- report$Reported_T
+    deaths_est_T <- apply(report$Reported_T, 1, max, na.rm=T)
+
+    data_T <- newDeaths(deaths_est_T,
+                        report_lag)
+    N <- dim(data_T$death.remain)[2]
+    X <- setup_data_lag(N, npar, result$dates_report[1:N],npar)
+    IndexX <- matrix(FALSE, N, N)
+    for(i in 1:N){
+        if( sum(data_T$death.remain[i,] == 0 & is.na(data_T$death.remain[i,] )==F) >0){
+            IndexX[i,i:(which.max(data_T$death.remain[i,]==0)-2)] =T
+        }
+    }
+    X <- X[IndexX[upper.tri(data_T$death.remain,diag = T)],]
+    index <- upper.tri(data_T$death.remain,diag = T) & IndexX
+    y = data_T$report.new[index ]
+    n = data_T$death.remain[index ]
+    index = is.na(y)==F & is.na(n)==F
+    X <- as.matrix(X[index,])
+    y <- y[index]
+    n <- n[index]
+    p <- dim(X)[2]
+    if(is.null(theta0))
+        theta0 <- rep(0,2*p)
+    res <- optim(theta0, function(x){-log_bb(x, y, n, X)  + sum(exp(x[(p+1):(2*p)]))})
+    while(res$convergence>0)
+        res <- optim(res$par,function(x){-log_bb(x, y, n, X) + sum(exp(x[(p+1):(2*p)]))})
+    if(res$convergence>0){
+        print('warning non convergence')
+    }else{
+        print(paste('log lik = ',-res$value))
+    }
+    return(list(alpha_X = res$par[1:p], beta_X = res$par[(p+1):(2*p)]))
+}
+
+
+##
+#'
+#'    lag    - (int)   lag days  (lag = 0 is only first non-holiday, lag = 1 two non holidays)
+#'    theta0 - (p x 1) inital guess of parameters
+##
+ML_betaBin_post <- function(result,
+                       lag,
+                       maxusage.day,
[…]

### 2020-06-11 src/lagged/testing_lag.R

@@ -1,0 +1,62 @@
+library(rstan)
+
+
+source("src/util.r")
+source("src/MLbeta.R")
+sim <- 1000
+unique.days = 5
+maxusage.day=20
+result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+Reported_T = result$detected
+N_T <- dim(Reported_T)[1]
+fixed = N_T
+lag <- 3
+
+report <- splitlag(result$detected[1:fixed,1:fixed],as.Date(result$dates_report[1:fixed]), lag)
+deaths_O_est <- apply(report$Reported_O, 1, max, na.rm=T)
+
+data_O <-newDeaths(deaths_O_est,
+                  report$Reported_O,
+                  maxusage.day)
+X_O <- setup_data_postlag(fixed, 2, result$dates_report[1:fixed], 1)
+X_O <- X_O[,colSums(X_O)>0]
+index <- upper.tri(data_O$death.remain,diag = T)
+y_O = data_O$report.new[index]
+n_O = data_O$death.remain[index]
+index = is.na(y_O)==F & is.na(n_O)==F
+X_O <- as.matrix(X_O[index,])
+
+y_O <- y_O[index]
+n_O <- n_O[index]
+fit_O <- glm( cbind(y_O, n_O) ~ X_O-1,
+                 family = "binomial")
+
+res1 <- optim(rep(0, dim(X_O)[2]*2),
+             function(x){-loglikProbBB(x, data_O$death.remain, data_O$report.new, X_O)$loglik},
+             function(x){-loglikProbBB(x, data_O$death.remain, data_O$report.new, X_O)$grad},
+             method = 'CG')
+res <- optim(res1$par,
+             function(x){-loglikProbBB(x, data_O$death.remain, data_O$report.new, X_O)$loglik})
+
+
+
+###
+# XT
+#
+###
+X_T <- setup_data_postlag(fixed, maxusage.day, result$dates_report[1:fixed], 2)
+index <- upper.tri(data_T$death.remain,diag = T)
+y_T = data_T$report.new[index]
+n_T = data_T$death.remain[index]
+index = is.na(y)==F & is.na(n)==F
+X_T <- as.matrix(X_T[index,])
+X_T <- X_T[,colSums(X_T)>0]
+y_T <- y_T[index]
+n_T <- n_T[index]
+fit_T <- glm( cbind(y_T, n_T) ~ X_T,
+                 family = "binomial")
[…]

### 2020-06-11 src/lagged/time_series_experiment.R

@@ -1,0 +1,101 @@
+##
+# simple stan code for data
+#
+##
+library(data.table)
+library(rstan)
+source(file.path("src", "GPutil.R"))
+source(file.path("src", "MH.R"))
+source(file.path("src", "util.R"))
+sim <- 2000
+fixed <- 64
+lag <- 1
+rstan_options(auto_write = TRUE)
+result <- readRDS(file.path("data", "processed", "processed_data.rds"))
+report <- splitlag(result$detected[1:fixed,1:fixed],as.Date(result$dates_report[1:fixed]), lag)
+
+Reported = result$detected
+N <- dim(Reported)[1]
+
+deaths_est <- apply(report$Reported_O, 1, max, na.rm=T)
+
+death_est <- deaths_est[1:N]
+
+N <- length(death_est)
+data.list <- list(N=N,
+                  Y = death_est)
+
+plot(death_est,xlab='days',ylab='deaths')
+tau <- 1000
+for(i in 20){
+
+    data.list_temp <- list(N=i, Y = death_est[1:i])
+    stan_out = stan(file = "src/stan/RW1_negbin.stan",
+                    data= data.list_temp,
+                    iter = sim,
+                    warmup = ceiling(0.5*sim),
+                    chains =1)
+    Theta_stan <- extract(stan_out)$theta
+    #tau        <- extract(stan_out)$tau
+
+    CI <- apply(exp(Theta_stan),2,quantile,c(0.1,0.5,0.9))
+    lines(CI[2,],col='blue',lty=2)
+    lines(CI[1,],col='red',lty=2)
+    lines(CI[3,],col='red',lty=2)
+
+    Start_theta <- log(data.list_temp$Y)
+    if(0){
+    ###
+    # negbin part
+    #
+    #
+    ###
+    tau <- 1000
+    phi <- 100
+    alpha.phi <- 20
+    MH_obj_GP <- MH_setup()
+    MH_obj_GP$sigma <- 0.01
+    n_theta <- length(Start_theta)
+    L <- toeplitz(c(-1,1, rep(0,n_theta-2)))
+    L[lower.tri(L,diag=F)] <- 0
+    L <- L[-n_theta,]
+    L<-as(L, "sparseMatrix")
+    MH_obj_GP$theta <- Start_theta
+    MH_obj_GP$Lik <- function(x, N, L, phi) {
+        prior_list <- prior_GP(x, L)
+        lik_list   <- lik_negbin_theta(N, x, phi)
[…]

### 2020-06-11 src/old/modelbenchmarkutil_poisson.R

@@ -1,189 +1,0 @@
-
-##
-#' running the benchmark for day j
-#'
-#' j - day j
-#' result -
-#' true.day  - (int) days not estimated
-#' maxusage.day - (int) how many reports days to use for each death
-#' unique.days  - (int) how many unique reports days to have as parameters
-#' MCMC_sim     - (int) how many simulations
-#' burnin_p     - ([0,1]) how many burnins to run in percantage of MCMC_sim
-#' prior        - (2 x 1) [1] 0: rw1, 1: rw2,
-#'                        [2] 0: no noise, 1: noise
-##
-benchmark_BetaGP_j <- function(j,
-                               result,
-                               true.day,
-                               maxusage.day,
-                               unique.days,
-                               MCMC_sim,
-                               burnin_p,
-                               deaths_sim,
-                               prior = c(0,0)){
-    require(Matrix)
-    N_T <- dim(result$detected)[1]
-    deaths_est_T <- apply(result$detected, 1, max, na.rm=T)
-    Reported <- result$detected[1:j,1:j]
-    dates_report <- result$dates_report[1:j]
-    N <- dim(Reported)[1]
-    deaths_est <- apply(Reported, 1, max, na.rm=T)
-    deaths_est[1:true.day] = deaths_est_T[1:(1+true.day-1)] #days with known deaths
-
-
-    data_ <- newDeaths(deaths_est,
-                       Reported,
-                       maxusage.day =maxusage.day)
-
-
-    ##
-    # building covariate matrix
-    ##
-    X <- setup_data(N, maxusage.day,  dates_report, unique.days)
-
-    ##
-    # seting up MCMC
-    ##
-    MH_obj <- MH_setup()
-    MH_obj$sigma <- 0.1
-    MH_obj$theta <- rep(0,2*dim(X)[2])
-
-
-    MH_obj$Lik <- loglikProbBB
-
-
-    ##
-    # seting up MCMC GP
-    ##
-    Start_theta <- log(deaths_est+1)
-    tau <- 0.1
-    MH_obj_GP <- MH_setup()
-    MH_obj_GP$sigma <- 0.1
-
-    MH_obj_GP$theta <- Start_theta
-    if(prior[1] == 0){
-        n_theta <- length(Start_theta)
-        L <- toeplitz(c(-1,1, rep(0,n_theta-2)))
[…]

### 2020-06-11 src/plots.R

@@ -47,7 +47,7 @@ ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
 # Load data
 benchmark <- read_fst(file.path("data", "processed", "constant_benchmark.fst"), as.data.table = TRUE)
 model <- read_fst(file.path("data", "processed", "model_benchmark.fst"), as.data.table = TRUE)
-benchmark <- benchmark[state >="2020-04-11"]
+benchmark <- benchmark[state >="2020-04-21"]
 
 # Fix data tables so they look the same
 reported_dead <- benchmark[, .(state, date, days_left = as.integer(days_left), reported_dead)]
@@ -127,7 +127,7 @@ plot_data[, type := factor(type)]
 
 # Figure 1 - Pick 4 dates at random to plot
 set.seed(1234)
-example_dates <- sample(seq(as.Date("2020-04-15"), model[days_left == 13, max(date)], 1), 4)
+example_dates <- sample(seq(as.Date("2020-04-21"), model[days_left == 13, max(date)], 1), 4)
 plot <- day_plot(plot_data[date %in% example_dates],
                  reported_dead[date %in% example_dates],
                  "")

### 2020-06-11 src/util.r

@@ -116,6 +116,65 @@ sample.deathsBB <- function(samples, deaths, alpha, beta, Reported, alpha.MCMC, 
 #  true.dat  - (int) how many days after recording should one sample
 #  prior     - (function) should  return log of prior density (use N and i)
 ##
+sample.deathsBB_bi <- function(samples, deaths, alpha, beta, Reported, alpha.MCMC, true.day = 0 ,
+                               p_null,
+                               pi_null,
+                            use.prior=FALSE,
+                            prior=NULL){
+
+  N <- length(deaths)
+  acc <- rep(0,N)
+  for(i in 1:N){
+    if(i > true.day){
+      alpha_i     = alpha[i,i:N]
+      beta_i      = beta[i,i:N]
+      Reported_i  = Reported[i,i:N]
+      index = is.na(Reported_i)==F
+      alpha_i  = alpha_i[index]
+      beta_i  = beta_i[index]
+      Reported_i  = Reported_i[index]
+      if(length(Reported_i)>0){
+        lik_i <- loglikDeathsGivenProbBB_bi(deaths[i],alpha_i , beta_i, pi_null,p_null, Reported_i)
+      }else{
+        lik_i <- 0
+      }
+      if(use.prior==T)
+        lik_i <- lik_i + prior(deaths[i], i)
+      for(j in 1:samples){
+        death_star <- sample((deaths[i]-alpha.MCMC[i]):(deaths[i]+alpha.MCMC[i]), 1)
+        if(length(Reported_i)>0){
+          lik_star <- loglikDeathsGivenProbBB_bi(death_star, alpha_i , beta_i, pi_null,p_null, Reported_i)
+        }else{
+          lik_star <- 0
+        }
+        if(use.prior==T)
+          lik_star <- lik_star + prior(death_star, i)
+        if(is.nan(lik_star))
+          next
+        if(log(runif(1)) < lik_star-lik_i){
+          lik_i = lik_star
+          deaths[i] <- death_star
+          acc[i] = acc[i] + 1
+        }
+      }
+    }
+  }
+  return(list(deaths=deaths, acc = acc))
+}
+
+
+##
[…]

### 2020-06-11 output/paper/paper.tex

@@
 As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
+Not how the model predicts additional deaths above the moving average line.
 
@@
 
-\chapter{Model Performance}
+\section{Model Performance}
+To say judge whether or not the model is accurate we need to compare it to a benchmark.
+The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week.
+Instead, we create a benchmark prediction by adding the mean historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
+We use the sum of standard deviations as confidence intervals, assuming that lags are independent across days.
+The exact calculation is described in the appendix.}.
 
-Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to a benchmark prediction based on historical averages.
+Figure~\ref{fig:four_dates} depicts four randomly chosen dates where the model is compared to the benchmark.
 The model and the benchmark are tasked with predicting the total number of individuals who have died at a given date and have been reported within 14 days of that date.
@@
 
+Benchmark and model point estimates are similarly close to the truth. The model produces tighter confidence intervals however, except during the very first day when only a single report is available.
+
 \begin{figure}
@@
 \chapter{Implications and limitations}
-The model proposed here has much better ability to estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden.
+The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden.
 To generate accurate estimates of the actual event frequencies based on these reports is highly relevant and can have large implications for interpretations of the trends and evolution of disease outbreaks.
 In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data.
-The method and algorithm proposed overcome major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic.
-It also provides valuable measures of uncertainty around these estimates, showing readers how large the range of possible outcomes can be.
+The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic.
+It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.
 
@@
 Nevertheless, the method also has its limitations.
-As presented, the model assumes that all deaths are reported in the same manner, given there exists many regions in Sweden this is unlikely to be the case.
-For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure compared to Stockholm.
+As presented, the model assumes that all deaths are reported in the same manner.
+Given there exists many regions in Sweden this is unlikely to be the case.
+For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm.
 Building a model for each region separately would most likely give better results and make the assumptions more reasonable.
 Unfortunately we do not currently have access to the high resolution data required to do so.
+
+Moreover, deaths are reported from two distinct populations that seem to follow different trends.
+At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths.
+But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes.
+
 Another limitation is that the model assumes that the number of new reported deaths for a given day cannot be negative, which is not actually true, due to miscount or misclassification of days.
@@
 The central assumption of the model is that the proportions deaths reported each day is fixed (up to the known covariates).
-If actual reporting standards change over time and its not explicitly modelled by a covariate, the model will not be able to account for this.
+If actual reporting standards change over time, the model will not be able to account for this.
 But reporting likely becomes faster as the crisis infrastructure improves.
@@
 Of course, there might be unknown variables that we have failed to incorporate, but at the least the model is an improvement from the estimates using moving averages.
-When the covariates to the reporting delay pattern is known, the model can incorporate them and provide a more accurate nowcasting of COVID-19 events.
+When the covariates to the reporting delay pattern are known, the model can incorporate them and provide more accurate predictions.
 
 \chapter{Conclusion}
+In this paper, we provide a simple method to accurately nowcast daily Covid-19 statistics that are reported with delay.
+By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends.
+Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.
 

### 2020-06-11 src/data_processing.R
recalculated deaths_dt to never use negative numbers

@@ -71,11 +71,19 @@ deaths_dt[!is.na(date) & publication_date > "2020-04-02", workdays_lag := n_work
 
 deaths_dt[date == "2020-04-02" & publication_date == "2020-04-02", `:=`(workdays_lag = 0, days_lag = 0)]
 
-deaths_dt[!is.na(date), paste0("n_m", 1) := shift(N, n = 1, type = "lag", fill = 0L), by = date]
+# Change reports to ensure no negatives
+deaths_dt[!is.na(date), n_m1 := shift(N, n = 1, type = "lag", fill = 0L), by = date]
+deaths_dt[!is.na(date), n_p1 := shift(N, n = 1, type = "lead", fill = NA_integer_), by = date]
+while (deaths_dt[N < n_m1, .N] > 0) {
+    # cat(deaths_dt[N - n_m1 < 0, .N], ", ")
+    deaths_dt[N < n_m1, N := ceiling((N + n_m1) / 2)]
+    deaths_dt[N > n_p1, N := ceiling((N + n_p1) / 2)]
+    deaths_dt[!is.na(date), n_m1 := shift(N, n = 1, type = "lag", fill = 0L), by = date]
+    deaths_dt[!is.na(date), n_p1 := shift(N, n = 1, type = "lead", fill = NA_integer_), by = date]
+}
+
 deaths_dt[!is.na(date), n_diff := N - n_m1]
-# deaths_dt[!is.na(date) & n_m1 > 0 & !is.na(n_m1), n_diff_pct := N/n_m1 - 1]
-# deaths_dt[!is.na(date) & n_m1 == 0 & N == 0, n_diff_pct := 0]
-deaths_dt[, n_m1 := NULL]
+deaths_dt[, c("n_m1", "n_p1") := NULL]
 
 setkey(deaths_dt, date, publication_date)
 write_fst(deaths_dt, file.path("data", "processed", "deaths_dt.fst"))

### 2020-06-11 src/extra_code_adam.r

@@ -1,0 +1,36 @@
+deaths_dt
+
+
+deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
+
+# Include only up to 14 days
+n_days <- 14
+DT <- deaths_dt[days_lag <= n_days]
+DT <- DT[date > "2020-04-02"]
+DT <- DT[date <= max(publication_date) - n_days]
+# DT[, N_14d := N[days_lag == 14], by = date]
+
+# Ignore negative reports
+DT[n_diff < 0, n_diff := 0]
+DT <- DT[!is.na(workdays_lag)]
+
+lag_means <- function(lags, N, groups = seq(0, 1, 0.25)) {
+    means <- data.table(cut(1:sum(N), quantile(1:sum(N), probs = groups), include = TRUE), rep(lags, N))[, .(l=max(V2)), by = V1][, l]
+    out <- data.table(t(means))
+    setnames(out, paste0("p", gsub("\\.", "", groups[2:length(groups)])))
+
+    return(out)
+}
+
+plot_data <- melt(DT[, lag_means(as.integer(workdays_lag), n_diff, groups = seq(0, 1, 0.25)), by = date], id.vars = "date")
+
+g <- ggplot(data = plot_data, aes(x = date, y = value, group = variable, color = variable)) + geom_line()
+g
+
+model[date == "2020-05-20", ]
+
+g <- ggplot(data = deaths_dt[date > "2020-04-02", sum(n_diff), by = .(date, workdays_lag <= 2)], aes(x = date, y = V1, color = workdays_lag, group = workdays_lag)) + geom_point() + geom_smooth()
+g
+
+deaths_dt[date > "2020-04-02", sum(n_diff), by = .(date, workdays_lag <= 1)][date <= "2020-04-15"]
+deaths_dt[date == "2020-04-12"]

### 2020-06-11 output/paper/paper.tex

@@
 
+\usepackage{amsmath}
+\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}
+
 % TITLE
@@
 	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
-	Joacim Rocklöv \and
-	Jonas Wallin
+	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Section of Sustainable Health, Umeå University}\thanks{Heidelberg Institute of Global Health, University of Heidelberg} \and
+	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
 }
@@
 In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays.
-The prediction model is building on methods developed in ecology, referred to as the ``removal method'' or ``capture-retain'' models (\cite{Pollock1991_review_papers}).
+The prediction model is building on methods developed in ecology, referred to as the ``removal method'' (\cite{Pollock1991_review_papers}).
 
@@
 The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week.
-Instead, we create a benchmark prediction by adding the mean historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
-We use the sum of standard deviations as confidence intervals, assuming that lags are independent across days.
+Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
+We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days.
 The exact calculation is described in the appendix.}.
@@
 SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix).
-The central plot shows the width of the confidence intervals, and the rightmost one the proportion of CI´s that cover the true value.
+The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.
 
@@
 But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes.
+However, wew have noted a clear decline in proportions of deaths reported the two first working days.
+In early april .\% of deaths where reported within the first two working days wheras in late may only \% where reported during the same time interval.
+We adress this by assuming that the two deaths from the two first working days comes from a different population then the remainder of days.
 

### 2020-06-11 output/paper/paper.tex
Updates from Overleaf

@@
 \usepackage{tabularx}
+\usepackage{amsthm}
 \newcolumntype{L}{>{\raggedright\arraybackslash}X}
-
-\usepackage{amsmath}
-\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}
+\newcommand{\mv}[1]{{\boldsymbol{#1}}}
 
@@
 	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
-	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Section of Sustainable Health, Umeå University}\thanks{Heidelberg Institute of Global Health, University of Heidelberg} \and
+	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Umeå University,\href{mailto:joacim.rocklov@umu.se}{joacim.rocklov@umu.se }}\thanks{Heidelberg Institute of Global Health, University of Heidelberg}  \and
 	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
@@
 
-	\noindent Keywords:
+	%\noindent Keywords:
 
@@
 	\item it does not relay any previous trend in the data,
-	\item we can generate confidence bounds for the uncertainty about daily true frequencies,
+	\item we can generate prediction intervals for the uncertainty about daily true frequencies,
 	\item the uncertainty estimates can be carried over to epidemiological models to help create more realistic models.
@@
 We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19.
-We then calculate the posterior distribution, prediction median and 95\% uncertainty intervals of the expected deaths from the reported deaths on each specific day.
+We then calculate the posterior distribution, prediction median and 95\% prediction intervals of the expected deaths from the reported deaths on each specific day.
 The method and algorithm is thoroughly described in the Supplementary Information.
@@
 Each data point is the average of all dates where predictions can be evaluated.
-SCRPS is a measure of accuracy that rewards precision, similar to a Brier score (see definition in Appendix).
+SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \cite{bolin2019scale}.
 The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.
 
-Benchmark and model point estimates are similarly close to the truth. The model produces tighter confidence intervals however, except during the very first day when only a single report is available.
+Benchmark and model point estimates are similarly close to the truth.
+The model produces tighter prediction intervals however, they are too tight between day 8-5 lags to reporting, see Figure \ref{fig:model_metrics}, this is likely due to that Swedish Public Health Agency does a database search during only certain days to match people reported dead to Covid infected.
+This type of reporting is currently not implemented in the model hence the poorer performance during these lags.
 
@@
 But statistics offer only aggregate numbers, prohibiting us from modeling two distinct processes.
-However, wew have noted a clear decline in proportions of deaths reported the two first working days.
-In early april .\% of deaths where reported within the first two working days wheras in late may only \% where reported during the same time interval.
-We adress this by assuming that the two deaths from the two first working days comes from a different population then the remainder of days.
+However, we have noted a clear decline in proportions of deaths reported the two first working days.
+For example the number of deaths occurring at the second of April $\approx 30\%$ of deaths where reported within the first two working days whereas for the eighteens of May only $\approx 10\%$ where reported during the two first working days.
+We address this by assuming that the deaths reported during the two first working days comes from a different population then the remainder of days.
 
@@
 \chapter{Conclusion}
-In this paper, we provide a simple method to accurately nowcast daily Covid-19 statistics that are reported with delay.
+In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay.
 By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends.

### 2020-06-11 output/paper/appendix.tex
Updates from Overleaf

@@
-% \documentclass[11pt, a4paper]{article}
-
-% \usepackage{upref,amsmath,amssymb,amsthm}
-% \usepackage{amsmath}
-% \usepackage{graphicx}
-% %\usepackage{caption}
-% \usepackage[normalem]{ulem}
-% %\documentclass[12pt,a4paper]{article}
-% %\usepackage{lipsum}
-% \usepackage{amsmath,amssymb,upref,amsthm,eucal}
-% %\usepackage{pstricks-add}
-% %\usepackage{pst-node}
-% %\usepackage[T1]{fontenc}
-% \usepackage{graphics}
-% \usepackage{mathrsfs}
-% \usepackage{natbib}
-
-% %%%ADDED BY KRYS
-% \usepackage{marginnote} %Package helping with margin notes within equations
-%                         %This package works well with ''pdflatex'' but not so well with standard ''latex''
-%                         %It is enough to comment the second line below to get rid of all margin-notes
-
-% %The following was added by K.P.
-% \usepackage[usenames, dvipsnames]{xcolor}
-% \definecolor{darkblue}{HTML}{2C326E}
-% %%%
-% %%%
-% \usepackage{marginnote} %Package helping with margin notes
-% %To change the width of paragraph on the margin
-% \setlength{\marginparwidth}{2.2cm}
-% \newcommand\mn[1]{
-% \marginnote{\tiny #1}
-% }
-
-% \newcommand\omn[1]{
-% \reversemarginpar
-% \mn{ #1}
-% \normalmarginpar
-% }
-% %% Label's name on the margin.
-% %% To remove label's names from margin simply comment the third line.
-% %% It is important not to use any 'Latex' meaningful sequences of characters in labeling.
-% %% For example `_` would lead to subscripting in the shown label.
-% \newcommand\labmarg[1]{
-% \label{#1}\mbox{} %\mbox{} is a trick to make it work in subsection environment
-% \mn{#1}
-% }
-
-% %\theoremstyle{plain}
-% \newtheorem{theorem}{Theorem}
-% \newtheorem{lemma}[theorem]{Lemma}
-% \newtheorem{remark}[theorem]{Remark}
-% \newtheorem{proposition}[theorem]{Proposition}
-% %\newtheorem{example}{example}
-
-% %\theoremstyle{definition}
-% \newtheorem{definition}[theorem]{Definition}%[section]
-% %\newtheorem{definition}{Definition}[subsection]
-% \theoremstyle{remark}
-% \newtheorem{example}{Example}
-
-% \DeclareMathOperator{\ran}{Ran}
-% \DeclareMathOperator{\cov}{Cov}
-% \DeclareMathOperator{\vect}{vec}
-% \DeclareMathOperator{\tr}{tr}
-% \DeclareMathOperator{\var}{Var}
-% \DeclareMathOperator{\spant}{Span}
-% \newcommand{\XB}{\mathbf{X}}
-% \newcommand{\YB}{\mathbf{Y}}
-% \newcommand{\UB}{\mathbf{U}}
-% \newcommand{\VB}{\mathbf{V}}
-% \newcommand{\PB}{\mathbf{P}}
-% \newcommand{\PH}{\mathbf{h}}
-% \newcommand{\obsy}{\mathscr{Y}}
-% \newcommand{\eps}{{\epsilon}}
-% \newcommand{\calH}{{\cal H}}
-% %%
-% % JW added
-% %%
-% \newcommand{\jw}[1]{\textcolor{red}{#1}}
-\newcommand{\mv}[1]{{\boldsymbol{\mathrm{#1}}}}
-% \usepackage{cleveref}
-% \newcommand{\Ordo}{\mathcal{O}}
-% \newcommand{\E}{\mathbb{E}}
-% \usepackage{algorithm}
-% \usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
-% \algnewcommand{\ffor}[1]{\State\algorithmicfor\ #1\ \algorithmicdo}
-% \algnewcommand{\Endffor}{\unskip\ \algorithmicend\ \algorithmicfor}
-
-% \newtheorem{thm}{Theorem}[section]
-% \newtheorem{alg}[thm]{Algorithm}
-
-% \usepackage{array,multirow,graphicx}
-% \newtheorem{claim}[theorem]{Claim}
-% \begin{document}
-
-% \title{Supp}
-
-%  \maketitle
-
 \section{Model}
-
 \subsection{Notation}
@@
 		$\mv{\beta}^H$ & $2 \times 1$ & parameter for the probability, $\mv{p}$ for holiday adjustment. \\
-		$\mv{\lambda}$ &  $T \times 1$ &  $\lambda_i$  is the intensity of the expected number of deaths at day $i$. \\
-		$\sigma^2$ & $1\times 1$ & Variation of the random walk prior for the log intensity.
-
+		$\mv{\mu}$ &  $T \times 1$ &  $\mu_i$  is the intensity of the expected number of deaths at day $i$. \\
+		$\sigma^2$ & $1\times 1$ & Variation of the random walk prior for the log intensity. \\
+		$\phi$ & $1\times 1$ & overdispersion parameter for negative binomial distribution. \\
+		$p_0$ & $1\times 1$ & probability of reporting for a low reporting event. \\
+		$pi$ & $1\times 1$ & probability of a low reporting event.
 	\end{tabularx}
@@
 These extra parameters are created to account for the under-reporting that occurs during weekend and holidays.
+Finally we add an extra mixture component that allows for very low reporting.
 
@@
 For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model.
-However, here we just assume a log-Gaussian Cox processes \cite{Moller1998_log_gaussian} where the Gaussian processes has a intrinsic random walk distribution \cite{Rue2005_gaussian_markov} i.e.
+However, here we just assume a log-Gaussian Cox processes \cite{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion.
+The latent Gaussian processes has a intrinsic random walk distribution \cite{Rue2005_gaussian_markov} i.e.
 \begin{align*}
-\log(\lambda_i) - \log(\lambda_{i-1}) &\sim N(0,\sigma^2),\\
-d_i| \lambda_i  &\sim Po(\lambda_i).
+\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
+d_i| \lambda_i  &\sim NegBin(\mu_i, \phi).
 \end{align*}
@@
 \begin{align*}
-\sigma^2 &\sim \Gamma^{-1}(0.01,0.01) \\
+\sigma^2 &\sim \Gamma(1,0.01) \\
+\phi &\sim \Gamma(1,0.01) \\
 \alpha_k &\sim U[0,\infty] \\
@@
 \beta_k^H &\sim U[0,\infty] \\
-\log(\lambda_i) - \log(\lambda_{i-1}) &\sim N(0,\sigma^2)\\
-d_i| \lambda_i  &\sim Po(\lambda_i) \\
+\log(\mu_i) - \log(\mu{i-1}) &\sim N(0,\sigma^2)\\
+d_i| \lambda_i  &\sim NegBin(\mu_i,\phi) \\
 p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
-r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
+r_{i,j}|d_i,\mv{r}_{1,1:j},p &\sim \pi Bin(d_i - \sum_{k=1}^{j-1}r_{i,k},p_0)+ (1-\pi) Bin(d_i - \sum_{k=1}^{j-1}r_{i,k}, p_{i,j}),
 \end{align*}
@@
 	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
-	\item Finally we sample $\sigma^2|\mv{d}$ directly since this distribution is explicit.
+	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
 \end{itemize}
@@
     \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_states}
-    \caption{}
+    \caption{Average SCRPS as the pandemic progresses.}
     \label{fig:SCRPS_states}
@@
     \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_weekdays}
-    \caption{Model accuracy over time for four dates, compared to a benchmark}
+    \caption{Average SCRPS per weekday.}
     \label{fig:SCRPS_weekdays}
 \end{figure}
-
-% TODO: Add figures
-
-% \bibliographystyle{chicago}
-% \bibliography{researchplan}
-% \end{document}

### 2020-06-11 output/paper/paper.tex
test

@@
-% !TEX TS-program = pdflatex
-% !TEX encoding = UTF-8 Unicode
-% !BIB TS-program = biber
-% !TeX root = ./paper.tex
-\documentclass[a4paper,11pt,article,oneside,openany,american]{memoir}
-\usepackage[memoir,bibliography]{setupscript}
-\addbibresource{bibliography.bib}
-
-% Customization
+\documentclass[a4paper,11pth,american]{article}
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+\usepackage{babel}
+\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
+\usepackage[babel]{csquotes}
+\usepackage{babel}
+\usepackage{natbib}
+\usepackage{amsmath,amssymb,amsfonts,amsthm}
+\usepackage{booktabs}
+\usepackage{graphicx}
+\DeclareGraphicsExtensions{.pdf,.png,.jpg,.eps}
 \usepackage{array}
@@
 \usepackage{tabularx}
-\usepackage{amsthm}
+\usepackage{hyperref}
 \newcolumntype{L}{>{\raggedright\arraybackslash}X}
@@
 \begin{document}
-\frontmatter%
-\begin{titlingpage}
     \maketitle
@@
 	\end{abstract}
-
-	%\noindent Keywords:
-
-\end{titlingpage}
-%\tableofcontents*
 %%%%%%%%%%%%%%%%%%%
 % MAIN TEXT SECTION
-\mainmatter%
-
-\chapter{Pandemic response demands timely data}
+\section{Pandemic response demands timely data}
 The new corona virus pandemic is affecting societies all around the world.
@@
 
-\section{The current situation of COVID-19 in Sweden}
+\subsection{The current situation of COVID-19 in Sweden}
 The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}.
@@
 
-\chapter{The removal method}
+\section{The removal method}
 We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty.
@@
 
-\chapter{Applying the model to COVID-19 in Sweden}
+\section{Applying the model to COVID-19 in Sweden}
 We propose a Bayesian version of the removal model that assumes an overdispersed binomial distribution for the daily observations of deaths in Sweden in COVID-19.
@@
 
-\section{Model Performance}
-To say judge whether or not the model is accurate we need to compare it to a benchmark.
-The moving average of reported deaths is not useful, since it becomes strongly biased for deaths that occurred within the last week.
+\subsection{Model Performance}
+To judge whether or not the model is accurate we need to compare it to a benchmark.
+The moving average of reported deaths is not useful, since it is biased for deaths that occurred within the last week.
 Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
@@
 Benchmark and model point estimates are similarly close to the truth.
-The model produces tighter prediction intervals however, they are too tight between day 8-5 lags to reporting, see Figure \ref{fig:model_metrics}, this is likely due to that Swedish Public Health Agency does a database search during only certain days to match people reported dead to Covid infected.
-This type of reporting is currently not implemented in the model hence the poorer performance during these lags.
+The model produces tighter prediction intervals.
+For 8-5 days of reporting lag (see Figure \ref{fig:model_metrics}), the intervals are too tight.
+This is likely because the Public Health Agency queries the Swedish death registry for Covid-19 deaths only once or twice a week.
+Since we do not know the process, it has not been explicitly modeled.
 
@@
 
-\chapter{Implications and limitations}
+\section{Implications and limitations}
 The model proposed here can estimate the trends in surveillance data with reporting delays, such as the daily COVID-19 reports in Sweden.
@@
 
-\chapter{Conclusion}
+\section{Conclusion}
 In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay.
@@
 % APPENDIX
-\printbibliography%
-\backmatter%
-\appendix%
-\chapter{Appendix}
+\bibliographystyle{chicago}
+\bibliography{bibliography}
+\appendix
+\section{Appendix}
 \input{appendix.tex}%

### 2020-06-11 output/paper/appendix.tex
tex

@@
 For the deaths, $\mv{d}$, one could imagine several different prior ideally some sort of epidemiological model.
-However, here we just assume a log-Gaussian Cox processes \cite{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion.
-The latent Gaussian processes has a intrinsic random walk distribution \cite{Rue2005_gaussian_markov} i.e.
+However, here we just assume a log-Gaussian Cox processes \citep{Moller1998_log_gaussian}, but instead of Poisson distribution we use a negative binomial to handle possible over dispersion.
+The latent Gaussian processes has a intrinsic random walk distribution \citep{Rue2005_gaussian_markov} i.e.
 \begin{align*}
@@
 As the main goal to generate inference of the number of death $\mv{d}$ is through the posterior distribution of number of deaths $\mv{d}$ given the observations $\mv{r}$.
-In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \cite{Brooks2011_handbook_markov}.
+In order to generate samples from this distribution we use a Markov Chain Monte Carlo method \citep{Brooks2011_handbook_markov}.
 In more detail we use a blocked Gibbs sampler, which generates samples in the following sequence:
@@
 	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution.
-Here to we use an adaptive MALA (\cite{Atchade2006_adaptive_version}) to sample from these parameters.
+Here to we use an adaptive MALA \citep{Atchade2006_adaptive_version} to sample from these parameters.
 	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.

### 2020-06-11 output/paper/paper.tex
tex

@@
-\documentclass[a4paper,11pth,american]{article}
+\documentclass[a4paper,11pth]{article}
 \usepackage[T1]{fontenc}
 \usepackage[utf8]{inputenc}
-\usepackage{babel}
-\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
-\usepackage[babel]{csquotes}
-\usepackage{babel}
+%\usepackage{babel}
+%\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
+%\usepackage[babel]{csquotes}
+%\usepackage{babel}
 \usepackage{natbib}
@@
 The new corona virus pandemic is affecting societies all around the world.
-As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions (\cite{No_author_2020_coronavirus_three}).
-Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers (\cite{Jajosky2004_evaluation_reporting}).
+As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}.
+Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}.
 The latest data are often not finalized, but change as new information is reported.
@@
 In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays.
-The prediction model is building on methods developed in ecology, referred to as the ``removal method'' (\cite{Pollock1991_review_papers}).
+The prediction model is building on methods developed in ecology, referred to as the ``removal method'' \citep{Pollock1991_review_papers}.
 
@@
 The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}.
-During a press conference, they present updates on the number of deaths, admissions to hosiptals and intensive care, as well as case counts.
+During a press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.
 
-One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic (\cite{Anderson2020_how_will}).
+One of the reasons for following these indicators is to enable public health professionals and the public to observe the evolving patterns of the epidemic \citep{Anderson2020_how_will}.
 In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response.
@@
 In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly.
-At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 (\cite{Ohman2020_antalet_virusdoda}).
+At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.
 
 \section{The removal method}
-We propose to use the removal method, developed in animal management (\cite{Pollock1991_review_papers}), to present an estimate of the actual frequencies at a given day and their uncertainty.
-The method has a long history dating back at least to the 1930s (\cite{Leslie1939_attempt_determine}).
-However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today (\cite{Matechou2016_open_models}).
+We propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty.
+The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}.
+However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}.
 It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.
@@
 
-A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals (\cite{Pollock1991_review_papers}).
+A classic example where the method proposed to solve this problem has been used is in estimating statistics of trapping a closed population of animals \citep{Pollock1991_review_papers}.
 Each day the trapped animals are collected, and kept, and if there is no immigration the number of trapped animals the following days will, on average, decline.
@@
 The graph shows reported and predicted deaths (with uncertainty intervals) as bars, and a dashed line plots the 7-day (centered) moving average.
-A version without predictions is used in the Public Health Agency´s daily press briefings.
+A version without predictions is used in the Public Health Agency\'s daily press briefings.
 As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
@@
 Each data point is the average of all dates where predictions can be evaluated.
-SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \cite{bolin2019scale}.
-The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI´s that cover the true value.
+SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}.
+The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.
 

### 2020-06-11 src/MLbeta.R

@@ -1,474 +1,0 @@
-
-source("src/util.r")
-##
-#'
-#'    lag    - (int)   lag days  (lag = 0 is only first non-holiday, lag = 1 two non holidays)
-#'    theta0 - (p x 1) inital guess of parameters
-##
-ML_betaBin <- function(result,
-                       lag,
-                       npar = 2,
-                       theta0 = NULL){
-
-    report <- splitlag(result$detected,as.Date(result$dates_report) ,lag)
-    report_lag <- report$Reported_T
-    deaths_est_T <- apply(report$Reported_T, 1, max, na.rm=T)
-
-    data_T <- newDeaths(deaths_est_T,
-                        report_lag)
-    N <- dim(data_T$death.remain)[2]
-    X <- setup_data_lag(N, npar, result$dates_report[1:N],npar)
-    IndexX <- matrix(FALSE, N, N)
-    for(i in 1:N){
-        if( sum(data_T$death.remain[i,] == 0 & is.na(data_T$death.remain[i,] )==F) >0){
-            IndexX[i,i:(which.max(data_T$death.remain[i,]==0)-2)] =T
-        }
-    }
-    X <- X[IndexX[upper.tri(data_T$death.remain,diag = T)],]
-    index <- upper.tri(data_T$death.remain,diag = T) & IndexX
-    y = data_T$report.new[index ]
-    n = data_T$death.remain[index ]
-    index = is.na(y)==F & is.na(n)==F
-    X <- as.matrix(X[index,])
-    y <- y[index]
-    n <- n[index]
-    p <- dim(X)[2]
-    if(is.null(theta0))
-        theta0 <- rep(0,2*p)
-    res <- optim(theta0, function(x){-log_bb(x, y, n, X)  + sum(exp(x[(p+1):(2*p)]))})
-    while(res$convergence>0)
-        res <- optim(res$par,function(x){-log_bb(x, y, n, X) + sum(exp(x[(p+1):(2*p)]))})
-    if(res$convergence>0){
-        print('warning non convergence')
-    }else{
-        print(paste('log lik = ',-res$value))
-    }
-    return(list(alpha_X = res$par[1:p], beta_X = res$par[(p+1):(2*p)]))
-}
-
-
-##
-#'
-#'    lag    - (int)   lag days  (lag = 0 is only first non-holiday, lag = 1 two non holidays)
-#'    theta0 - (p x 1) inital guess of parameters
-##
-ML_betaBin_post <- function(result,
-                       lag,
-                       maxusage.day,
-                       npars,
[…]

### 2020-06-11 src/lagged/Beta_GP_lag_benchmark_refit.R

@@ -216,6 +216,7 @@ model_benchmark_bias$days_left =  model_benchmark_bias$date -model_benchmark_bia
 
 names(model_benchmark) <- c('state','date','days_left','target','SCRPS','predicted_deaths','ci_lower','ci_upper')
 model_benchmark<-model_benchmark[,c('state','date','target','days_left','ci_upper','ci_lower','predicted_deaths','SCRPS')]
+model_benchmark$days_left =  model_benchmark$date -model_benchmark$state +14
 
 write_fst(model_benchmark, file.path("data", "processed", "model_benchmark.fst"))
 write_fst(model_benchmark_bias, file.path("data", "processed", "model_benchmark_bias.fst"))

### 2020-06-11 src/lagged/MLbeta.R

@@ -328,12 +328,14 @@ benchmark_BetaGP_post_lag_j <- function(j,
     N_j <- j
 
     X_j <- setup_data_postlag2(j, lag, npars, result_j$dates_report)
+    X_M <- setup_data_postlag2(j, lag, npars, result_j$dates_report)
     deaths_est <- apply(report_j$Reported_O, 1, function(x) { if(all(is.na(x))){return(0)};
                                                             max(x,na.rm=T)})
     X_mu <- X_j
+    #X_mixed <- setup_data_mixed_effect(N, 2, result$dates_report[1:N])
 
     p  <- dim(X_mu)[2]
-    X_M  <- cbind(rep(1,dim(X_mu)[1]),rowSums(X_mu[,4:5])>0)
+    #X_M  <- cbind(rep(1,dim(X_mu)[1]),rowSums(X_mu[,4:5])>0)
     p1 <- dim(X_M)[2]
 
 

### 2020-06-11 src/util.r

@@ -781,6 +781,72 @@ loglik <- function(beta,death.remain, report.new, X, sigma){
 ##
 #####
 
+##
+# likelihood observations given deaths and probabilites
+# model is:
+# newreport[upper.tri] \sim   BB(deaths[upper.tri], exp(X\beta_1), exp(X\beta_2))
+#
+#' @return obj - list
+#'               $loglik  - logliklihood
+#'               $grad    - gradient
+##
+loglikProbBB_mixedeffect_bi <- function(beta, death.remain, report.new, X_mu, X_M, X_mixed, sigma_mixed){
+
+  p <- dim(X_mu)[2]
+  p1 <- dim(X_M)[2]
+  p2 <- dim(X_mixed)[2]
+  beta_mu    <- beta[1:p]
+  beta_M     <- beta[p + (1:p1)]
+  beta_mixed <- beta[p + p1 + (1:p2)]
+  N <- dim(report.new)[1]
+
+  index <- upper.tri(death.remain,diag = T)
+  y = report.new[index]
+  n = death.remain[index]
+  index = is.na(y)==F & is.na(n)==F
+  X_mu    <- X_mu[index,]
+  X_mixed <- X_mixed[index,]
+  X_M     <- X_M[index,]
+  mu      <- 1/(1+exp(-X_mu%*%beta_mu - X_mixed%*%beta_mixed))
+  M       <- exp(X_M%*%beta_M)
+
+  alpha <- M * mu
+  beta  <- M * (1-mu)
+  y <- y[index]
+  n <- n[index]
+
+  if(max(beta)>120)
+    return(list(loglik = -Inf, grad = 0))
+
+
+  lik <- sum(dBB(y, size = n, alpha = alpha, beta = beta, log.p=T))
+  if(is.na(lik))
+    return(list(loglik = -Inf, grad = 0))
+
+  ##
+  #
+  # mixed effect
+  ##
+  lik <- lik - (0.5/sigma_mixed^2) * sum(beta_mixed^2)
+  ##
+  # prior
+  ##
+  #lik <- lik - sum(exp(beta_M))
+  grad_lik <- grad_dBB(y, size = n, alpha = alpha, beta = beta)
+
+  grad_mu <- (M*mu * (1-mu) * grad_lik$grad_alpha)
+  grad_mu <- (grad_mu - (M * mu * (1-mu) * grad_lik$grad_beta))
[…]

### 2020-06-16 src/plots.R
made lines slightly darker

@@ -8,6 +8,7 @@ source("src/functions.R")
 
 #
 w <- 11 # plot width (inches)
+my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000") # wes_palette("Darjeeling2") # (replaced #ECCBAE)
 
 # Plot 1 = Predictions and current stats
 deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
@@ -78,7 +79,7 @@ model[is.na(target), predicted_deaths := model[!is.na(target), unique(target), b
 
 # Plot a specific day
 day_plot <- function(DT, reported, plot.title) {
-    colors <- c("#ECCBAE", "#046C9A", "gray50")
+    colors <- c("#d1ae90", "#046C9A", "gray50")
     colors <- setNames(colors, c(levels(DT$type), "Reported"))
 
     if (any(DT[!is.na(target), uniqueN(target) != 1, by = date][, V1])) {
@@ -106,7 +107,7 @@ day_plot <- function(DT, reported, plot.title) {
         # Theming
         set_default_theme() +
         # scale_fill_manual(values = fill_colors, limits = label_order, drop = FALSE) +
-        # scale_color_manual(values = wes_palette("Darjeeling2")) +
+        # scale_color_manual(values = my_palette) +
         scale_color_manual(values = colors) +
         scale_y_continuous(minor_breaks = seq(0,200,10), breaks = seq(0,200,40), expand = expansion(add = c(1, 5))) +
         labs(title = plot.title,
@@ -174,7 +175,7 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
     geom_point(aes(y = value)) +
     facet_wrap(~variable, scales = "free_y") +
     set_default_theme() +
-    scale_color_manual(values = wes_palette("Darjeeling2")) +
+    scale_color_manual(values = my_palette) +
     labs(#title = "Model metrics",
          #subtitle = "",
          #caption = "",
@@ -200,7 +201,7 @@ plot_data <- rbindlist(list(
 plot <- ggplot(data = plot_data, aes(x = state, y = V1, color = type, group = type)) +
     geom_line() + geom_point() +
     set_default_theme() +
-    scale_color_manual(values = wes_palette("Darjeeling2")) +
+    scale_color_manual(values = my_palette) +
     labs(#title = "Model metrics",
          #subtitle = "",
[…]

### 2020-06-16 output/paper/paper.tex
Updates from Overleaf

@@
     \begin{abstract}
-        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world.
+        The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world.
 The availability of unbiased timely statistics of trends in disease events are a key to effective responses.
@@
 \section{Pandemic response demands timely data}
-The new corona virus pandemic is affecting societies all around the world.
+The SARS-CoV-2 pandemic is affecting societies all around the world.
 As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}.
@@
 As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
-Not how the model predicts additional deaths above the moving average line.
+Note how the model predicts additional deaths above the moving average line.
 
@@
 In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data.
-The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic.
+The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the SARS-CoV-2 virus pandemic.
 It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.
@@
 Given there exists many regions in Sweden this is unlikely to be the case.
-For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm.
+For example, it is easy to see that the Swedish region Västra Götland follows a differewnt reporting structure than Stockholm.
 Building a model for each region separately would most likely give better results and make the assumptions more reasonable.
@@
 
-Moreover, deaths are reported from two distinct populations that seem to follow different trends.
+Moreover, deaths seem to be reported from two distinct populations that follow different trends.
 At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths.
@@
 However, we have noted a clear decline in proportions of deaths reported the two first working days.
-For example the number of deaths occurring at the second of April $\approx 30\%$ of deaths where reported within the first two working days whereas for the eighteens of May only $\approx 10\%$ where reported during the two first working days.
-We address this by assuming that the deaths reported during the two first working days comes from a different population then the remainder of days.
+For example, out of all deaths that occurred on the second of April, $\approx 30\%$ were reported within the first two workdays, whereas for the eighteenth of May only $\approx 10\%$ were reported that fast.
+We address this by assuming that the deaths reported during the two first working days comes from a different population than the remainder of days.
 

### 2020-09-15 src/data_processing.R
updated data_processing to correct typo

@@ -31,7 +31,7 @@ load_fhm_deaths <- function(f) {
 
     setnames(DT, c("date", "N"))
 
-    DT[date == "Uppgift saknas" | date == "uppgift saknas", date := NA]
+    DT[date == "Uppgift saknas" | date %in%c("uppgift saknas","Uppgift saknaa"), date := NA]
 
     if (can_be_numeric(DT[, date])) {
         DT[, date := as.Date(as.numeric(date), origin = "1899-12-30")]

### 2020-09-16 src/Beta_GP_benchmark_build.R
cleaned up the files and created a main file

@@ -1,41 +1,0 @@
-##
-#  building the benchmark prediction parameters for the full model
-#  need to run Beta_GP_benchmark.R
-##
-library(foreach)
-library(tidyr)
-library(doParallel)
-source(file.path("src", "util.r"))
-source(file.path("src", "MH.R"))
-source(file.path("src", "functions.R"))
-source(file.path("src", "GPutil.R"))
-source(file.path("src", "modelbenchmarkutil_poisson.R"))
-nclust <- 4
-MCMC_sim <- 15000
-burnin_p = 0.5
-deaths_sim <- 5
-maxusage.day = 20 #must be less then N
-unique.days  = 2
-true.day = 5
-start.predict.day = 11#16 # more then unique days
-
-result <- readRDS(file.path("data", "processed", "processed_data.rds"))
-N_T <- dim(result$detected)[1]
-cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
-doParallel::registerDoParallel(cl)
-foreach(j = start.predict.day:N_T)  %dopar% {
-    cat('j=',j,", ", file=stdout())
-    result <- readRDS(file.path("data", "processed", "processed_data.rds"))
-  res_save<-benchmark_BetaGP_j(j,
-                               result,true.day,
-                               maxusage.day,
-                               unique.days,
-                               MCMC_sim,
-                               burnin_p,
-                               deaths_sim = deaths_sim)
-
-  save(res_save,
-       file = file.path("data", "simulation_results_model1", paste0("param_", result$dates_report[j], ".rds")))
-}
-
-parallel::stopCluster(cl)

### 2020-09-16 src/Beta_GP_lag_benchmark_build.R

@@ -1,15 +1,24 @@
 ##
-#  building the benchmark prediction parameters for the full model
+#  builds and stores the parameters
+#  (theta    - parameter of the detection probabilites
+#  Death_est - estimated through MCMC)
+#  needed for the benchmark
 #  need to run Beta_GP_benchmark.R
+#
+#  load file:
+#  data/processed/processed_data.rds (buildData.R)
+
+#  generates the files:
+#  data/simulation_results/param_lagged_YYYY-MM-DD.rdx
 ##
 library(foreach)
 library(doParallel)
 library(tidyr)
-source(file.path("src", "util.r"))
-source(file.path("src", "MH.R"))
-source(file.path("src", "functions.R"))
-source(file.path("src", "GPutil.R"))
-source(file.path("src","lagged" ,"MLbeta.R"))
+source(file.path("src", "util","util.r"))
+source(file.path("src", "util","MH.R"))
+source(file.path("src", "util","functions.R"))
+source(file.path("src", "util","GPutil.R"))
+source(file.path("src","util","MLbeta.R"))
 nclust <- 6
 lag <- 1
 MCMC_sim <- 15000

### 2020-09-16 src/Beta_GP_lag_benchmark_refit.R

@@ -1,20 +1,30 @@
 ###
 # benchmarking
+# using parameters(theta     - parameters of the detection probabilites
+#                  Death_est - true number of deaths)
+# from Beta_GP_lag_benchmark_build.R to generate
+# preidcition of the reporting
 #
+# load files:
+#  data/processed/processed_data.rds   (buildData.R)
+#  data/simulation_results/param_lagged_YYYY-MM-DD.rdx (Beta_GP_lag_benchmark_build.R)
+#
+# generates files:
+# data/processed/model_latest.fst
+# data/processed/model_lag_refit.fst
+# data/processed/model_benchmark.fst
+# data/processed/model_benchmark_bias.fst
 ###
 library( stringr)
 library(foreach)
 library(tidyr)
 library(fst)
 library(data.table)
-source("src/util.r")
+source(file.path("src","util","util.r"))
 nclust <- 6
 report.day = 14 # how long after death we wish to predicit
 start.predict.day=20
 path.to.files <- file.path("data")
-#files <- list.files('/Users/jonaswallin/Dropbox/temp/simulation_results_beta',
-#                    pattern = "^param",
-#                    full.names = TRUE)
 files <- list.files(paste(path.to.files,"/simulation_results",sep=""),
                              pattern = "^param",
                              full.names = TRUE)

### 2020-09-16 src/buildData.R

@@ -1,29 +1,18 @@
 ###
-# load data
+#  Building the data file for the model
+#  main thing done is averaing over overcount of deaths and creates and average of them see line:
+# 36-48
+# load files:
+# data/processed/deaths_dt.fst (data_processing.R)
+#
+# generates the files:
+# data/processed/processed_data.rds
+#
 ##
 library(Matrix)
 library(stringr)
 library(dplyr)
 library(fst)
-
-# source("src/functions.R")
-
-# Sys.setlocale("LC_TIME", 'en_US.UTF-8')
-# #download_latest_fhm()
-# path.to.files <- file.path("data")
-# fhm_files = list_fhm_files(folder = paste(path.to.files,"/FHM/",sep=""))
-# death_dts <- c()
-# for(i in 1:length(fhm_files)){
-#   date_in <- as.Date(fhm_files[i]%>% str_match_all("[0-9]{4}-[0-9]{2}-[0-9]{2}") %>%unlist)
-#   if(date_in > as.Date('2020-04-01'))
-#     death_dts = rbind(death_dts,load_fhm(fhm_files[i]))
-# }
-#
-# # res <- death_dts %>% dplyr::group_by(date, publication_date) %>% dplyr::spread(publication_date, value=N)
-# res <- death_dts %>% dplyr::group_by(date, publication_date)
-# death_dt <- data.table(death_dts)
-# setkey(death_dt, publication_date, date)
-# death_dt <-death_dt[!is.na(date) & publication_date > "2020-04-02" & date > "2020-04-02"]
 
 deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
 deaths_dt <- deaths_dt[date > "2020-04-01", .(date, publication_date, N)]

### 2020-09-16 src/constant_benchmark.R

@@ -1,8 +1,18 @@
+##
+# builds the constant benchmark model
+#
+# load files:
+# data/processed/deaths_dt.fst (data_processing.R)
+#
+# generates the files:
+# data/processed/constant_benchmark.fst
+##
+
 library(data.table)
 library(fst)
 library(tidyr)
 library(ggplot2)
-source("src/util.scores.R")
+source(file.path("src","util","util.scores.R"))
 
 alpha <- 0.1
 

### 2020-09-16 src/data_processing.R

@@ -1,3 +1,14 @@
+##
+# building base data file from FHM data files
+#
+# reads files:
+#  data/FHM/Folkhalsomyndigheten_Covid19_YYYY-MM-DD.xlsx
+# (external files avialable at <https://github.com/adamaltmejd/covid>)
+#
+# generates the files:
+# /data/processed/deaths_dt.fst
+##
+
 library(data.table)
 library(readxl)
 library(stringr)

### 2020-09-16 src/main.R

@@ -1,0 +1,11 @@
+##
+# run file to create and reproduce the figures in the article
+# make sure that correct FHM files exists at data/FHM/
+# they are aviable from <https://github.com/adamaltmejd/covid>
+##
+source("src/data_processing.R")
+source("src/buildData.R")
+source("src/constant_benchmark.R")
+source("src/Beta_GP_lag_benchmark_build.R")
+source("src/Beta_GP_lag_benchmark_refit.R")
+source("src/plots.R")

### 2020-09-16 src/old/Beta_GP_benchmark_build.R

@@ -3,6 +3,7 @@
 #  need to run Beta_GP_benchmark.R
 ##
 library(foreach)
+library(tidyr)
 library(doParallel)
 source(file.path("src", "util.r"))
 source(file.path("src", "MH.R"))
@@ -10,27 +11,16 @@ source(file.path("src", "functions.R"))
 source(file.path("src", "GPutil.R"))
 source(file.path("src", "modelbenchmarkutil_poisson.R"))
 nclust <- 4
-MCMC_sim <- 20000
+MCMC_sim <- 15000
 burnin_p = 0.5
 deaths_sim <- 5
 maxusage.day = 20 #must be less then N
-unique.days  = 6
+unique.days  = 2
 true.day = 5
-start.predict.day = 40#16 # more then unique days
+start.predict.day = 11#16 # more then unique days
 
 result <- readRDS(file.path("data", "processed", "processed_data.rds"))
-Reported_T = result$detected
-N_T <- dim(Reported_T)[1]
-
-deaths_est_T <- apply(Reported_T, 1, max, na.rm=T)
-
-data_T <- newDeaths(deaths_est_T,
-                   Reported_T,
-                   maxusage.day =maxusage.day)
-X_T <- setup_data(N_T, maxusage.day, result$dates_report, unique.days)
-predicition.list <-list()
-
-
+N_T <- dim(result$detected)[1]
 cl <- parallel::makeCluster(nclust,setup_strategy = "sequential", outfile="")
 doParallel::registerDoParallel(cl)
 foreach(j = start.predict.day:N_T)  %dopar% {
@@ -45,7 +35,7 @@ foreach(j = start.predict.day:N_T)  %dopar% {
                                deaths_sim = deaths_sim)
 
   save(res_save,
-       file = file.path("data", "simulation_results_old", paste0("param_", result$dates_report[j], ".rds")))
+       file = file.path("data", "simulation_results_model1", paste0("param_", result$dates_report[j], ".rds")))
 }
 
 parallel::stopCluster(cl)

### 2020-09-16 src/plots.R

@@ -1,10 +1,26 @@
+##
+# generate the figures for the article x
+#
+# read files:
+# data/processed/deaths_dt.fst           (data_processing.R)
+# data/processed/model_latest.fst        (Beta_GP_lag_benchmark_build.R)
+# data/processed/constant_benchmark.fst  (constant_benchmark.R)
+# data/processed/model_benchmark.fst.fst (Beta_GP_lag_benchmark_build.R)
+#
+# generates files:
+# output/plots/SCRPS_over_weekdays.pdf
+# output/plots/daily/prediction_YYYY-MM-DD.pdf
+# output/plots/SCRPS_over_states.pdf
+# output/plots/model_metrics.pdf
+##
+
 library(data.table)
 library(fst)
 library(forcats)
 library(ggplot2)
 library(hrbrthemes)
 library(wesanderson)
-source("src/functions.R")
+source(file.path("src", "util","functions.R"))
 
 #
 w <- 11 # plot width (inches)

### 2020-09-16 src/util/MLbeta.R

@@ -1,5 +1,5 @@
 
-source("src/util.r")
+source(file.path("src", "util","util.r"))
 ##
 #'
 #'    lag    - (int)   lag days  (lag = 0 is only first non-holiday, lag = 1 two non holidays)

### 2020-09-16 output/paper/paper.tex

@@
 \author{
-	Adam Altmejd\thanks{SOFI, Stockholm University and Swedish House of Finance, \href{mailto:adam@altmejd.se}{adam@altmejd.se}} \and
-	Joacim Rocklöv\thanks{Department of Public Health and Clinical Medicine, Umeå University,\href{mailto:joacim.rocklov@umu.se}{joacim.rocklov@umu.se }}\thanks{Heidelberg Institute of Global Health, University of Heidelberg}  \and
-	Jonas Wallin\thanks{Department of statistics, Lund university, \href{mailto:jonas.wallin81@gmail.com}{jonas.wallin81@gmail.com}}
+.
 }
@@
         The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world.
-The availability of unbiased timely statistics of trends in disease events are a key to effective responses.
-But due to reporting delays, the most recently reported numbers are frequently underestimating of the total number of infections, hospitalizations and deaths creating an illusion of a downward trend.
-Here we describe a statistical methodology for predicting true daily quantities and their uncertainty, estimated using historical reporting delays.
+The availability of unbiased timely surveillance statistics, such as the trends in disease frequencies, are a key to evaluating and planning responses.
+But due to reporting delays, the most recently reported numbers are often underestimating of the total number of new infections, hospitalizations and deaths.
+This sometimes creates an illusion of a downward trend.
+Here we describe a statistical methodology for predicting the expected actual daily surveillance frequencies and their uncertainty, considering historical reporting delays.
 The methodology takes into account the observed distribution pattern of the lag.
@@
 The SARS-CoV-2 pandemic is affecting societies all around the world.
-As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}.
-Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers \citep{Jajosky2004_evaluation_reporting}.
-The latest data are often not finalized, but change as new information is reported.
+As countries are challenged to control and fight back, they are in need of timely, unbiased, surveillance data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}.
+Official statistics are usually reported with long delay after thorough verification, but in the midst of a deadly pandemic, real time data is of critical importance for policymakers to evaluate and plan rapid responses \citep{Jajosky2004_evaluation_reporting}.
+But as a general rule, the latest data are often not complete, but change as new information is reported.
 In fact, reporting delays make the most recent days have the least cases accounted for, producing a dangerous illusion of an always improving outlook.
 
-Still, these unfinished statistics offer crucial information.
+Still, these unfinished surveillance statistics offer crucial information.
 If the pandemic is indeed slowing, we should not wait for the data to be finalized before using it.
-Rather, we argue that actual case counts and deaths should be nowcasted to account for reporting delay, thus allowing policymakers to use the latest numbers availiable without beinig misled by reporting bias.
+Rather, we argue that actual case counts and deaths should be nowcasted to account for reporting delays, thus allowing policymakers to use the latest numbers available surveillance without being misled by the obvious reporting bias.
 
 Such predictions provide an additional feature that is perhaps even more important.
-They explicitly model the uncertainty about these unknown quantities, ensuring that all users of these data have the same view of the current state of the epidemic.
+They explicitly model the uncertainty in these unknown quantities, ensuring that all users of these data have the same view of the current state of the epidemic, including the uncertainty.
 
-In this paper we describe a statistical methodology for nowcasting the epidemic statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays.
+In this paper we describe a statistical methodology for nowcasting the surveillance statistics, such as hospitalizations or deaths, and their degrees of uncertainty, based on the daily reported event frequency and the observed distribution pattern of reporting delays.
 The prediction model is building on methods developed in ecology, referred to as the ``removal method'' \citep{Pollock1991_review_papers}.
 
-To help motivate why such forecasting is needed, we now turn to the case of Sweden. The model is flexible by design, however, and could easily be applied to other countries as well.
+To help motivate why such forecasting is needed, we now turn to the case of Sweden, and illustrate how this method and can provide more reliable and relevant data to decision-makers in the midst of a pandemic.
+We stress, however, that the model is flexible by design and that it could easily be applied to other countries and surveillance data as well.
 
@@
 The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}.
-During a press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.
+During frequently re-occurring press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.
 
@@
 In relation to policy, it is of specific interest to understand if the growth rates changes, which could indicate the need for a policy response.
-However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces the illusion of a downward trend.
+However, in each daily report only a proportion of the number of recent deaths is yet known, and this bias produces makes the recent data presented less useful.
 
-The death counts suffer from the longest reporting delay.
+The death counts strongly suffer from the longer reporting delays.
 In their daily press conference, the Swedish Public Health Agency warns for this by stopping the reported 7-day moving average trend line 10 days before the latest date.
-But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a sense of a downward trend.
-In fact, this might be the reason why the number of daily deaths have been underestimated repeatedly.
-At the peak, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.
+But not only are deaths often reported far further back than 10 days, a bar plot still shows the latest information, creating a false impression of sense of a downward trend.
+In fact, in the Swedish situation, this might be the reason why the number of daily deaths have been underestimated repeatedly.
+As an example illustrating this, at the peak of the Swedish outbreak in the spring 2020, deaths were initially believed to level out at around 60 per day, but after all cases had been reported more than two weeks later, the actual number was close to 120 \citep{Ohman2020_antalet_virusdoda}.
 
 \section{The removal method}
-We propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty.
+To correct for the reporting delays we propose to use the removal method, developed in animal management \citep{Pollock1991_review_papers}, to present an estimate of the actual frequencies at a given day and their uncertainty.
 The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}.
@@
 
-The removal method that has three major advantages over simply reporting moving averages:
+The removal method has three major advantages over the simplistic approach to just reporting moving averages:
 
@@
 This pattern of declining number of trapped animals allows one to draw inference of the underlying population size.
-Here we replace the animal population with the true number of deaths
-on a given day.
-Instead of traps we have the new reports of COVID--19 events.
-As the number of new reported deaths for a given day declines, we can draw inference on how many actually died that day.
-If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number.
+Here we replace the animal population with the true number of deaths on a given day.
+For COVID-19, instead of traps we have the new reports of COVID--19 events.
+As the number of non-yet reported deaths for a given day declines according to some patterns over time, we can use this patterns to draw inference on how many actually died a specific day.
+If we assume that the reporting structure is constant over time we can after a while quickly get good estimate of the actual number, not the reported one.
 
@@
 To account for this, we allow the estimated proportions of daily reported cases to follow a probability distribution taking into consideration what type of day it is.
+
+We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \cite{lawless1994adjustments}.
+In \cite{lawless1994adjustments} the author developed the Occurred But Not Reported method (OBNR), which can give a frequentist point estimate of the number of deaths, and under a Poisson distributional assumption, generate prediction intervals.
+A Bayesian version of this method for nowcasting was developed in \cite{hohle2014bayesian}.
 
@@
 As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
-Note how the model predicts additional deaths above the moving average line.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
-    \caption{Swedish Covid-19 deaths as of 2020-06-04 and model predictions}
-    \label{fig:latest_prediction}
-\end{figure}
+We note specifically how the model predicts additional deaths above the moving average line.
 
@@
 The moving average of reported deaths is not useful, since it is biased for deaths that occurred within the last week.
-Instead, we create a benchmark prediction by a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
+Instead, we create a benchmark prediction by using a Normal distribution where the mean and standard deviation is taken from the historical lags from the last two weeks to the reported numbers\footnote{For a death date two days ago we add the mean of deaths reported after 3 days, 4 days, etc.
 We use the sum of standard deviations to generate the prediction intervals, assuming that lags are independent across days.
@@
 
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\textwidth]{../plots/lag_prediction_by_date}
-    \caption{Model accuracy over time for four dates, compared to benchmark}
-    \label{fig:four_dates}
-\end{figure}
-
 Figure~\ref{fig:model_metrics} shows model performance compared to the benchmark for three difference performance metrics.
@@
 Each data point is the average of all dates where predictions can be evaluated.
-SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}.
+In forecast evaluation, a function that guarantees, asymptotically, that the forecasting distribution (if available) will be chosen, is known as a proper scoring rule, see \cite{gneiting2007strictly}.
+In \cite{hohle2014bayesian} validation was done using the proper scoring rules logarithmic score and ranked probability score.
+These two scores requires access to either the PMF or CDF of the predictive distribution.
+Since we don't have easy access to these functions we instead use the  SCRPS (Scaled continuous proability ranking score, see definition in Appendix) \citep{bolin2019scale}.
 The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.
@@
 Since we do not know the process, it has not been explicitly modeled.
-
-\begin{figure}
-    \centering
-    \includegraphics[width=0.9\textwidth]{../plots/model_metrics}
-    \caption{Model accuracy over time for four dates, compared to a benchmark}
-    \label{fig:model_metrics}
-\end{figure}
 
@@
 Whenever case statistics are collected from multiple sources and attributed to its actual event date in the middle of a public health emergency, similar reporting delays to the ones in Sweden will necessarily occur.
-The method described thus has implications and value beyond Sweden, for any situation where nowcasts of disease event frequencies are of relevance to public health.
+The method described thus has implications and value beyond Sweden, for any situation where nowcasts of surveillance data are of relevance to public health.
 
-Nevertheless, the method also has its limitations.
+In \cite{schneble2020nowcasting} nowcasting of number of Covid-19 deaths in Germany was preformed.
+For nowcasting they applied the OBNR method, using penalized smoothing splines.
+The accuracy of their nowcasting model is, however, not possible to compare to our model as the study by Schneble et al.
+didn\'t validate the nowcasting on data withheld from the model fitting.
+
+Our method has its limitations.
 As presented, the model assumes that all deaths are reported in the same manner.
 Given there exists many regions in Sweden this is unlikely to be the case.
-For example, it is easy to see that the Swedish region Västra Götland follows a differewnt reporting structure than Stockholm.
+For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm.
 Building a model for each region separately would most likely give better results and make the assumptions more reasonable.
@@
 \section{Conclusion}
-In this paper, we provide a method to accurately nowcast daily Covid-19 statistics that are reported with delay.
-By systematically modelling the delay, policy makers can avoid dangerous illusory downward trends.
-Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during this pandemic.
+In this paper, we provide a method to accurately adjust for reporting delays.
+This is illustrated by nowcasting daily Covid-19 deaths in Sweden, which are reported with strong delay patterns.
+By systematically accounting for the delay by nowcasting, better timing in the evaluation and planning of responses and policy can be achieved, which is of great importance for in pandemic situations.
+Our model also gives precise uncertainty intervals, making sure users of these statistics are aware of the fast-paced changes that are possible during pandemics.
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% APPENDIX
+\section{Additional Information}
+The authors declare no conflict of interest.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%'
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newpage
+
+% Author contributions
 \bibliographystyle{chicago}
 \bibliography{bibliography}
+\input{contributions.tex}%
+
+\newpage
+
+\input{Figure1.tex}%
+\newpage
+\input{Figure2.tex}%
+\newpage
+
+\input{Figure3.tex}%
+
+\newpage
 \appendix

### 2021-03-03 output/paper/appendix.tex
Updates from Overleaf

@@
+\renewcommand{\thefigure}{A\arabic{figure}}
+\setcounter{figure}{0}
+
 \section{Model}
@@
 \log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2),\\
-d_i| \lambda_i  &\sim NegBin(\mu_i, \phi).
+d_i| \phi, \mu_i  &\sim NegBin(\mu_i, \phi).
 \end{align*}
@@
 \beta_k^H &\sim U[0,\infty] \\
-\log(\mu_i) - \log(\mu{i-1}) &\sim N(0,\sigma^2)\\
-d_i| \lambda_i  &\sim NegBin(\mu_i,\phi) \\
+\log(\mu_i) - \log(\mu_{i-1}) &\sim N(0,\sigma^2)\\
+d_i| \phi,\mu_i  &\sim NegBin(\mu_i,\phi) \\
 p_{i,j}|  \mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H &\sim Beta(\alpha^H_j\alpha_{min(j-i,k)},\beta^H_j\beta_{min(j-i,k)}) \\
@@
 \begin{itemize}
-	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$  follows a Beta-Binomial distribution.
+	\item  We sample $\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H|\mv{d}, \mv{r}$ using the fact that one can integrate out $p$ in the model, and then  $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\mu},\phi$  follows a Beta-Binomial distribution.
 Here to we use an adaptive MALA \citep{Atchade2006_adaptive_version} to sample from these parameters.
-	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\lambda}$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
-	\item To sample $\mv{\lambda} | \mv{d},\sigma^2$ we again use an adaptive MALA.
+	\item  To sample $\mv{d}|\mv{\alpha},\mv{\beta}, \mv{\alpha}^H,\mv{\beta}^H,\mv{r},\mv{\mu},\phi$, that each death, $d_i$ is conditionally independent, and we just use a Metropolis Hastings random walk to sample each one.
+	\item To sample $\mv{\mu} | \mv{d},\sigma^2,\phi$ we again use an adaptive MALA.
 	\item Finally We sample $\sigma^2|\mv{d}$,and  $p_0,\pi$ directly since this distribution is explicit, and $\phi$ using a MH-RW.
@@
     \includegraphics[width=0.9\textwidth]{../plots/SCRPS_over_states}
-    \caption{Average SCRPS as the pandemic progresses.}
+    \caption{Average SCRPS as the pandemic progresses.w}
     \label{fig:SCRPS_states}

### 2021-03-03 output/paper/paper.tex
Updates from Overleaf

@@
 \usepackage[utf8]{inputenc}
-%\usepackage{babel}
-%\usepackage[protrusion=true,expansion=true,final,babel]{microtype}
-%\usepackage[babel]{csquotes}
-%\usepackage{babel}
-\usepackage{natbib}
+\usepackage[square,numbers]{natbib}
+%\usepackage[style=nature]{biblatex}
 \usepackage{amsmath,amssymb,amsfonts,amsthm}
@@
 \subsection{The current situation of COVID-19 in Sweden}
-The Swedish Public Health Agency updates the COVID-19 statistics daily\footnote{The data is published on \url{https://www.folkhalsomyndigheten.se/smittskydd-beredskap/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/}.}.
+The Swedish Public Health Agency updates the COVID-19 statistics daily (the data is published on https://www.folkhalsomyndigheten.se/smittskydd-beredskap
+/utbrott/aktuella-utbrott/covid-19/bekraftade-fall-i-sverige/)..
 During frequently re-occurring press conference, they present updates on the number of deaths, admissions to hospitals and intensive care, as well as case counts.
@@
 The method has a long history dating back at least to the 1930s \citep{Leslie1939_attempt_determine}.
-However, the first refined mathematical treatment of the method is credited to \citet{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}.
+However, the first refined mathematical treatment of the method is credited to \cite{Moran1951_mathematical_theory}, more modern derivatives exits today \citep{Matechou2016_open_models}.
 It is a commonly applied method today when analyzing age cohorts in fishery and wildlife management.
@@
 
-We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \cite{lawless1994adjustments}.
+We note that a similar method has been proposed for delay adjusted reporting, but without using the removal method \citep{lawless1994adjustments}.
 In \cite{lawless1994adjustments} the author developed the Occurred But Not Reported method (OBNR), which can give a frequentist point estimate of the number of deaths, and under a Poisson distributional assumption, generate prediction intervals.
@@
 % Author contributions
-\bibliographystyle{chicago}
+\bibliographystyle{unsrt}
 \bibliography{bibliography}
 \input{contributions.tex}%
+
+\newpage
+\section{Figure legends}
+Figure 1. Swedish Covid-19 deaths as of 2020-06-04 and model predictions.
+
+Figure 2. Model accuracy over time for four dates, compared to a benchmark.
+
+Figure 3. Model accuracy over time for four dates, compared to benchmark.
 

### 2021-03-04 src/plots.R
renv update

@@ -58,7 +58,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
          x = "Death date",
          y = "Number of deaths")
 
-ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "latest_prediction.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
 
 
@@ -150,7 +150,7 @@ plot <- day_plot(plot_data[date %in% example_dates],
                  reported_dead[date %in% example_dates],
                  "")
                  #"Predicting the number of deaths in a given day reported within 14 days.")
-ggsave(filename = file.path("output", "plots", "lag_prediction_by_date.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "lag_prediction_by_date.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w)
 
 # For verification, plot each date as well
@@ -160,7 +160,7 @@ for (i in seq_along(dates)) {
     plot <- day_plot(plot_data[date == dates[i]],
                      reported_dead[date == dates[i]],
                      plot.title = dates[i])
-    ggsave(filename = file.path("output", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
+    ggsave(filename = file.path("output", "paper", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
            plot = plot, device = cairo_pdf, width = w, height = w)
 }
 #
@@ -199,7 +199,7 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
          x = "Days of lag to predict",
          y = "")
 
-ggsave(filename = file.path("output", "plots", "model_metrics.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "model_metrics.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
 
 #
@@ -225,7 +225,7 @@ plot <- ggplot(data = plot_data, aes(x = state, y = V1, color = type, group = ty
          x = "Last date included in the model",
          y = "SCRPS")
 
-ggsave(filename = file.path("output", "plots", "SCRPS_over_states.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "SCRPS_over_states.pdf"),
[…]

### 2021-03-04 src/plots.R
moved plots to paper/plots

@@ -58,7 +58,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
          x = "Death date",
          y = "Number of deaths")
 
-ggsave(filename = file.path("output", "paper", "plots", "latest_prediction.pdf"),
+ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
 
 
@@ -150,7 +150,7 @@ plot <- day_plot(plot_data[date %in% example_dates],
                  reported_dead[date %in% example_dates],
                  "")
                  #"Predicting the number of deaths in a given day reported within 14 days.")
-ggsave(filename = file.path("output", "paper", "plots", "lag_prediction_by_date.pdf"),
+ggsave(filename = file.path("output", "plots", "lag_prediction_by_date.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w)
 
 # For verification, plot each date as well
@@ -160,7 +160,7 @@ for (i in seq_along(dates)) {
     plot <- day_plot(plot_data[date == dates[i]],
                      reported_dead[date == dates[i]],
                      plot.title = dates[i])
-    ggsave(filename = file.path("output", "paper", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
+    ggsave(filename = file.path("output", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
            plot = plot, device = cairo_pdf, width = w, height = w)
 }
 #
@@ -199,7 +199,7 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
          x = "Days of lag to predict",
          y = "")
 
-ggsave(filename = file.path("output", "paper", "plots", "model_metrics.pdf"),
+ggsave(filename = file.path("output", "plots", "model_metrics.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
 
 #
@@ -225,7 +225,7 @@ plot <- ggplot(data = plot_data, aes(x = state, y = V1, color = type, group = ty
          x = "Last date included in the model",
          y = "SCRPS")
 
-ggsave(filename = file.path("output", "paper", "plots", "SCRPS_over_states.pdf"),
+ggsave(filename = file.path("output", "plots", "SCRPS_over_states.pdf"),
[…]

### 2021-03-04 src/plots.R
reran plots to check (no change)

@@ -24,7 +24,7 @@ source(file.path("src", "util","functions.R"))
 
 #
 w <- 11 # plot width (inches)
-my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000") # wes_palette("Darjeeling2") # (replaced #ECCBAE)
+my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000")
 
 # Plot 1 = Predictions and current stats
 deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
@@ -58,7 +58,7 @@ plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
          x = "Death date",
          y = "Number of deaths")
 
-ggsave(filename = file.path("output", "plots", "latest_prediction.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "latest_prediction.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w/1.9)
 
 
@@ -150,7 +150,7 @@ plot <- day_plot(plot_data[date %in% example_dates],
                  reported_dead[date %in% example_dates],
                  "")
                  #"Predicting the number of deaths in a given day reported within 14 days.")
-ggsave(filename = file.path("output", "plots", "lag_prediction_by_date.pdf"),
+ggsave(filename = file.path("output", "paper", "plots", "lag_prediction_by_date.pdf"),
        plot = plot, device = cairo_pdf, width = w, height = w)
 
 # For verification, plot each date as well
@@ -160,7 +160,7 @@ for (i in seq_along(dates)) {
     plot <- day_plot(plot_data[date == dates[i]],
                      reported_dead[date == dates[i]],
                      plot.title = dates[i])
-    ggsave(filename = file.path("output", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
+    ggsave(filename = file.path("output", "paper", "plots", "daily", paste0("prediction_", dates[i], ".pdf")),
            plot = plot, device = cairo_pdf, width = w, height = w)
 }
 #
@@ -199,7 +199,7 @@ plot <- ggplot(data = plot_data, aes(x = factor(days_left), color = type, group 
          x = "Days of lag to predict",
          y = "")
 
-ggsave(filename = file.path("output", "plots", "model_metrics.pdf"),
[…]

### 2021-03-04 src/main.R
added new data and tried to get main.R to run

@@ -1,7 +1,7 @@
 ##
 # run file to create and reproduce the figures in the article
-# make sure that correct FHM files exists at data/FHM/
-# they are aviable from <https://github.com/adamaltmejd/covid>
+# make sure that correct updated FHM files exist at data/FHM/
+# new files can be downloaded from <https://github.com/adamaltmejd/covid>
 ##
 source("src/data_processing.R")
 source("src/buildData.R")

### 2021-03-18 src/buildData.R
merge merge

@@ -18,49 +18,61 @@ library(stringr)
 library(dplyr)
 library(fst)
 
-deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
-deaths_dt <- deaths_dt[date > "2020-04-01", .(date, publication_date, N)]
-##
-#only relevant stuff below
-##
-res2 <- deaths_dt %>% tidyr::spread(publication_date, N)
 
-# dayes with no reporting
-index_NoReport <- res2[,1]$date%in%unique(deaths_dt$publication_date)==F
+buildData <- function(country = "sweden"){
 
-deaths_dt[, .N, date]
-deaths_dt[is.na(date), .N]
-deaths_dt[date < "2020-03-11"]
+    if(country == "sweden"){
+        deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt.fst"), as.data.table = TRUE)
+        deaths_dt <- deaths_dt[date > "2020-04-01", .(date, publication_date, N)]
+    }else if(country=="uk"){
+        deaths_dt <- fread(file.path("data", "uk", "uk.csv"))
+        deaths_dt <- deaths_dt[, .(date, publication_date, N)]
+        pub_dates <- unique(deaths_dt$publication_date)
+        # uk never publish on the same date
+        deaths_dt <- rbind(deaths_dt,
+                           data.table( publication_date = pub_dates,
+                                       date             = pub_dates,
+                                       N = 0))
+    }
+    ##
+    #only relevant stuff below
+    ##
+    res2 <- deaths_dt %>% tidyr::spread(publication_date, N)
 
-# create a detected matrix for date x date
-n.days <- dim(res2)[1]
-detected <- matrix(NA, nrow= n.days,
-                      ncol =n.days)
-detected[,index_NoReport == F] = as.matrix(res2[,2:dim(res2)[2]])
-colnames(detected) <- c(res2[,1]$date)
-#d_ <- diag(detected)
-#d_[is.na(d_)] <- 0
-#d_[index_NoReport] <- NA
-#diag(detected) <- d_
-repeated = 1
-while(repeated != 0){
-    repeated = 0
-    for(i in 1:(dim(detected)[1]-1)){
-        for(j in i:(dim(detected)[2]-1)){
-            if(is.na(detected[i,j])==F & is.na(detected[i,j+1])==F){
-                if(detected[i,j] > detected[i,j+1]){
[…]

### 2021-03-18 src/model2/dataanalysis/ProbabilityAnalysis.R

@@ -5,8 +5,8 @@ source(file.path("src","model2","v5","regression.R"))
 source(file.path("src","model2","v5","model.R"))
 
 graphics.off()
-#data <- readRDS(file.path("data", "processed", "processed_data.rds"))
-data <- readRDS(file.path("data", "processed", "processed_data_icu.rds"))
+data <- readRDS(file.path("data", "processed", "processed_data.rds"))
+#data <- readRDS(file.path("data", "processed", "processed_data_icu.rds"))
 
 
 report_cleaned <- report_clean(data$detected,data$dates)
@@ -16,13 +16,15 @@ colnames(new_cases) <- as.character(data$dates)
 diag(new_cases) <- NA
 N <- apply(new_cases,1,sum, na.rm=T)
 lag <- 15
-to. <- 25
+to. <- 30
 Prob <- matrix(0,nrow=dim(new_cases)[1]-to.,ncol=lag)
 for(i in 1:(dim(new_cases)[1]-to.)){
   cases <- new_cases[i,]
+  m <- min(length(cases)-i,lag)
+  cases <- cases[i+ (1:m)]
   cases <- cases[is.na(cases)==F ]
-  m <- min(length(cases),lag)
-  Prob[i,1:m] = cases[1:m]/sum(cases[1:m])
+
+  Prob[i,1:m] = cases[1:m]/sum(new_cases[i,], na.rm=T)
   #Prob[i,1:m] = cases[1:m]/N[i]
 }
 dates <- as.Date(data$dates[1:(dim(new_cases)[1]-to.)])
@@ -30,10 +32,10 @@ N <- N[1:(dim(new_cases)[1]-to.)]
 N_min <- 6
 x11()
 par(mfrow=c(2,2))
-plot(dates[N>N_min],apply(Prob[N>N_min,1,drop=F],1,sum))
-plot(dates[N>N_min],apply(Prob[N>N_min,1:2,drop=F],1,sum))
-plot(dates[N>N_min],apply(Prob[N>N_min,1:4,drop=F],1,sum))
-plot(dates[N>N_min],apply(Prob[N>N_min,5:10, drop=F],1,sum))
+plot(dates[N>N_min],apply(Prob[N>N_min,1:3,drop=F],1,sum))
+plot(dates[N>N_min],apply(Prob[N>N_min,1:5,drop=F],1,sum))
[…]

### 2021-03-18 src/model2/stan/INLA_time.R

@@ -18,7 +18,7 @@ icu_pre <- data.frame(date = as.Date(icu_pre$Datum_vårdstart),
 
 Reported = icu$detected
 N <- dim(Reported)[1]
-incomplete <- 10 # remove the ten latest
+incomplete <- 8 # remove the ten latest
 incomplete_deaths <- 30
 icu_est <- rep(0,N-incomplete)
 for(j in 1:(N -incomplete)){
@@ -76,5 +76,10 @@ glm.fit <- glm(y~x,poisson, data= data.frame(y=y,x=X))
 lines(Cov_analysis$dates,exp(cbind(1,X)%*%glm.fit$coefficients),col='red',lty=3)
 
 icu_cov <- data.frame(date= dates[dates%in%Cov_analysis$dates], icu = X)
-saveRDS(icu_cov, file.path("data", "processed", "icu_covariates.rds"))
+#saveRDS(icu_cov, file.path("data", "processed", "icu_covariates.rds"))
+
+X <- c(rep(0,i-1), theta)
+dates_pred <- c(dates,dates[length(dates)]+1:7)
+plot(dates_pred,exp(glm.fit$coefficients[1] + glm.fit$coefficients[2]*X),type='l',col='red')
+points(Cov_analysis$dates,Cov_analysis$deaths)
 

### 2021-03-18 src/model2/v10/example.run.R

@@ -8,9 +8,9 @@ source(file.path("src","model2","v10","model.R"))
 days_run <- 30
 # todo addapt sample N
 ##
-sim <- 40000
+sim <- 10000
 data <- readRDS(file.path("data", "processed", "processed_data.rds"))
-j <- dim(data$detected)[1] -29
+j <- dim(data$detected)[1] -41
 start_ = j - days_run # run the last 31 days
 
 report_cleaned <- report_clean(data$detected[start_:j,start_:j],data$dates[start_:j])
@@ -23,8 +23,8 @@ model_parameters <- list(sim           = sim,
                          N.days.fixed  =  3,
                          quantile      = c(0.025,0.975))
 
-prior_list <- list(mu_beta        = c(0,0,0,0,0),
-                   Sigma_beta     = 50*diag(5),
+prior_list <- list(mu_beta        = c(0,0,0,0),
+                   Sigma_beta     = 50*diag(4),
                    a_sigma        = c(3,3,3),
                    b_sigma        = c(10/2,10/2,10/2),
                    theta_mu         = c(0,0),

### 2021-03-18 src/model2/v10/model.R

@@ -114,8 +114,7 @@ model <- function(new_cases, model_paramters, prior_list, startvalue_list=NULL){
                                       X_time$fourth[i, index_]+
                                   X_time$third[i, index_],
                                   X_time$rest[i, index_],
-                                  X_time$Tuesday[i,index_],
-                                  X_time$Wednesday[i,index_])
+                                  X_time$Tuesday[i,index_])
     X_full <- rbind(X_full  , timePoints_MH[[i]]$X)
     #X_fixed <- rbind(X_fixed, timePoints_MH[[i]]$X_fixed)
 
@@ -216,8 +215,8 @@ model <- function(new_cases, model_paramters, prior_list, startvalue_list=NULL){
 
         Nstar <- sample((N[i]-alpha.MCMC[i]):(N[i]+alpha.MCMC[i] ), 1)
         if(Nstar >= sum(timePoints_MH[[i]]$n.obs)){
-          lik      <- density_t((timePoints_MH[[i]]$theta), timePoints_MH[[i]]$n.obs, N[i], Prob)   + dnegbin(N[i], lambda[i],phi)
-          lik_star <-  density_t((timePoints_MH[[i]]$theta), timePoints_MH[[i]]$n.obs, Nstar, Prob) + dnegbin(Nstar, lambda[i],phi)
+          lik      <- density_t((timePoints_MH[[i]]$theta), timePoints_MH[[i]]$n.obs, N[i], Prob)   #+ dnegbin(N[i], lambda[i],phi)
+          lik_star <-  density_t((timePoints_MH[[i]]$theta), timePoints_MH[[i]]$n.obs, Nstar, Prob) #+ dnegbin(Nstar, lambda[i],phi)
           if(log(runif(1)) < lik_star-lik ){
             N[i] <- Nstar
             acc_N[i] <- acc_N[i] + 1
@@ -264,7 +263,7 @@ model <- function(new_cases, model_paramters, prior_list, startvalue_list=NULL){
       a_hat  <- a_sigma + colSums(X_sigma)/2
       report_effort = report_effort - X_full%*%beta_time
       b_hat  <- b_sigma + 0.5  *t(X_sigma)%*%(report_effort)^2
-      sigma_time    <- sqrt(rinvgamma(2, shape= a_hat, rate=b_hat))
+      sigma_time    <- sqrt(rinvgamma(length(a_hat), shape= a_hat, rate=b_hat))
 
 
     }

### 2021-03-18 src/model2/v12/analysis.full.R

@@ -4,7 +4,7 @@
 #
 ##
 graphics.off()
-store_data_folder <- file.path("data","tmp","model2","v13")
+store_data_folder <- file.path("data","tmp","model2","v12")
 data_files <- list.files(store_data_folder,pattern='Npost*')
 
 data <- c()
@@ -14,7 +14,8 @@ for(file in data_files){
 data$df <- as.Date(data$State)-as.Date(data$dates)
 
 days = 1:14
-data <- data[order(data$State),]
+data <- data[order(data$State,decreasing = T),]
+data <- data[1:2000,]
 for(day in days){
     index <- data$df==day
     data_temp <- data[index,]
@@ -29,7 +30,7 @@ for(day in days){
     cat(' (',round(mean(CI_under),2),',',round(mean(CI_over),2),")" ,sep="")
     cat('\n')
 }
-day_ <- 2
+day_ <- 7
 index <- data$df==day_
 data_temp <- data[index,]
 #data_temp <- data_temp[80:sum(index),]

### 2021-03-18 src/model2/v12/run.model.all.R

@@ -24,7 +24,7 @@ run.model.all <- function(deaths, icu){
     #i icu_covariates  updateras filer
     cov_ <- icu_covariates(deaths, icu)
 
-    store_data_folder <- file.path("data","tmp","model2","v13")
+    store_data_folder <- file.path("data","model")
 
     ####
 
@@ -45,6 +45,8 @@ run.model.all <- function(deaths, icu){
                            theta_Sigma    =  5*diag(2),
                            mu_phi         = 1,
                            sigma_phi      = 1)
+
+    output <- data.frame()
 
     for(j in (days_run+1):length(deaths$dates)){
         start_ = j - days_run # run the last 31 days
@@ -95,6 +97,46 @@ run.model.all <- function(deaths, icu){
             colnames(Npost)[c(1,7)] <- c("State",'Truth')
             saveRDS(Npost, file = paste(store_data_folder,"/Npost_",j,'.rds',sep=""))
             saveRDS(result$posteriror_list, file = paste(store_data_folder,"/prior_",j,'.rds',sep=""))
+        }else{
+            Npost <-readRDS( file = paste(store_data_folder,"/Npost_",j,'.rds',sep=""))
+            report_cleaned <- report_clean(deaths$detected[start_:j,start_:j],deaths$dates[start_:j])
+            new_cases <- newCases(report_cleaned)
+            rownames(new_cases) <- as.character(deaths$dates[start_:j])
+            colnames(new_cases) <- as.character(deaths$dates[start_:j])
+        }
+        Total_repported <- rowSums(new_cases, na.rm=T)
+
+        output_temp <-data.frame(prediction_date  = Npost$State,
+                                 date             = Npost$dates,
+                                 sure_deaths      = Total_repported[names(Total_repported)%in%Npost$dates],
+                                 predicted_deaths = Npost$median-Total_repported[names(Total_repported)%in%Npost$dates],
+                                 predicted_deaths_lCI = Npost$lCI,
+                                 predicted_deaths_uCI = Npost$uCI,
+                                 total                = Npost$media)
+        date_to_predict <- Npost$State[1] + (max(as.Date(Npost$dates))-Npost$State[1]+1):0
+        cov_index <- cov_$theta_cov$date%in% date_to_predict
+        cov_data  <- cov_$theta_cov$theta[cov_index]
+        icu_pred <- qpois(model_parameters$quantile[1],
[…]

### 2021-03-18 src/model2/v13/Probability_fit.R

@@ -1,0 +1,47 @@
+
+probability_analysis <- function(model_death_dt ,days.reported=5, lag = 20){
+
+    report_cleaned <- report_clean(model_death_dt$detected,model_death_dt$dates)
+    new_cases <- newCases(report_cleaned)
+    rownames(new_cases) <- as.character(model_death_dt$dates)
+    colnames(new_cases) <- as.character(model_death_dt$dates)
+    diag(new_cases) <- NA
+    N <- apply(new_cases,1,sum, na.rm=T)
+    Prob <- matrix(0,nrow=dim(new_cases)[1]-lag,ncol=30)
+    Analysis_data <- c()
+    coeff <- c()
+    dates <- c()
+    for(i in (days.reported+1):(dim(new_cases)[1]-lag)){
+        cases <- sum(new_cases[(i-days.reported),1:i], na.rm=T)
+        Total <- sum(new_cases[i-days.reported,], na.rm=T)
+        if(is.na(cases)==F & Total >0){
+
+
+            Analysis_data <- rbind(Analysis_data, c(cases, Total))
+            Analysis_glm  <- data.frame(Y = Analysis_data[,1],
+                                        N = Analysis_data[,2]-Analysis_data[,1])
+            fit <- glm(cbind(Y,N)~1, data=Analysis_glm, family=binomial)
+            coeff <- c(coeff,fit$coefficients)
+            dates <- c(dates,colnames(new_cases)[i])
+        }
+    }
+
+    plot.data <- data.frame(date = dates, prob = 1/(1+exp(-coeff)))
+    ggfig <- ggplot(data = plot.data, aes(y = prob, x = date)) +
+        geom_point()
+    return(list(fit = plot.data, fig = ggfig, new_cases= new_cases))
+}
+
+graphics.off()
+model_death_dt <- readRDS(file.path("data", "processed", "processed_data.rds"))
+days.reported <- 10
+
+res <- probability_analysis(model_death_dt,days.reported=days.reported)
+print(res$fig)
+
+
+i <- dim(res$new_cases)[1]
+cases <- sum(res$new_cases[(i-days.reported),1:i], na.rm=T)
+size <- 0:200
+prob_size = dbinom(cases,size,prob=res$fit[dim(res$fit)[1],2])
+plot(size, prob_size/sum(prob_size),typ='l')

### 2021-03-18 src/model2/v13/Probability_fit_v2.R

@@ -1,0 +1,184 @@
+
+library(gamm4)
+library(ggplot2)
+source(file.path("src","model2","v13","icu_regression.R"))
+source(file.path("src","model2","v13","eval.coverage_prediction.R"))
+
+#'
+#' @param gamm model
+CI_pred <- function(model, new.data, obs, sim = 2000,sizes=0:200, prior=NULL){
+
+
+    beta <- coef(model);
+    Vb <- vcov(model)
+    ## simulate replicate beta vectors from posterior...
+    Cv <- chol(Vb)
+    nb <- length(beta)
+    br <- t(Cv) %*% matrix(rnorm(sim*nb),nb,sim) + beta
+
+    ## turn these into replicate linear predictors...
+    Xp <- predict(model,newdata=new.data,type="lpmatrix")
+    lp <- Xp%*%br
+    fv <- c(model$family$linkinv(lp)) ## ... finally, replicate expected value vectors
+    Total <- rep(0,sim)
+    if(is.null(prior)){
+
+        for(i in 1:sim){
+            prob_size = dbinom(obs,sizes,prob=fv[i])
+            prob_size = prob_size/sum(prob_size)
+            Total[i] <- sample(sizes,1,prob=prob_size)
+        }
+    }else{
+        prior_ <- prior(sizes)
+        for(i in 1:sim){
+            prob_size = dbinom(obs,sizes,prob=fv[i],log=T) + prior_
+            prob_size = exp(prob_size-max(prob_size))
+            prob_size = prob_size/sum(prob_size)
+            Total[i] <- sample(sizes,1,prob=prob_size)
+        }
+    }
+    return(Total)
+}
+
+probability_analysis <- function(model_dt,
+                                 model_death_dt ,
+                                 days.reported=5,
+                                 lag = 20,
+                                 prior_=NULL){
+
+    report_cleaned <- report_clean(model_death_dt$detected,model_death_dt$dates)
+    new_cases <- newCases(report_cleaned)
+    rownames(new_cases) <- as.character(model_death_dt$dates)
+    colnames(new_cases) <- as.character(model_death_dt$dates)
+    diag(new_cases) <- NA
+    N <- apply(new_cases,1,sum, na.rm=T)
+    Analysis_data <- c()
+    coeff <- c()
+    dates <- c()
+    prediciton.set <-c()
+    vcovs <- list()
+
+
+    for(i in (days.reported+1):(dim(new_cases)[1]-lag)){
+
[…]

### 2021-03-18 src/model2/v13/eval.coverage_prediction.R

@@ -1,0 +1,99 @@
+##
+# output analysis
+#
+##
+
+#'
+#'@param model_death_dt   - data for deaths so far
+#'@param death_prediction - data.frame with predicitions
+#'@param days.ago - how long ago since the predicition was made
+#'@param lag      - only report for result lag days ago
+#'@param max.days - only report the max.days
+coverage_data <- function(model_death_dt,
+                          death_prediction,
+                          days.ago=0,
+                          lag = 20){
+
+    N <- length(model_death_dt$dates)
+    death_reported_so_far <- data.frame(
+                                        T_deaths = apply(model_death_dt$detected,1,max, na.rm=T)[1:(N-lag)],
+                                        dates  =  model_death_dt$dates[1:(N-lag)])
+    df <- as.Date(death_prediction$prediction_date)-as.Date(death_prediction$date)
+    death_prediction_mod <- death_prediction[df==days.ago,]
+    death_prediction_mod <- death_prediction_mod[death_prediction_mod$prediction_date%in%death_reported_so_far$dates,]
+    death_prediction_mod <- merge(death_prediction_mod,
+                                        death_reported_so_far,
+                                        by.x='date',
+                                        by.y='dates')
+    coverage <- data.frame(
+                           date            = death_prediction_mod$date,
+                           reported        = death_prediction_mod$T_deaths,
+                           prediciton      = death_prediction_mod$total,
+                           uCI             = death_prediction_mod$total_uCI,
+                           lCI             = death_prediction_mod$total_lCI,
+                           lmiss           = death_prediction_mod$T_deaths < death_prediction_mod$total_lCI,
+                           umiss           = death_prediction_mod$T_deaths > death_prediction_mod$total_uCI
+                           )
+
+    coverage <- coverage[coverage$date >= "2020-05-01",]
+    return(coverage)
+}
+#'
+#'
+#'@param coverage - data from coverage_data
+coverage.plot <- function(coverage, days.ago, type, theme=NULL){
+    title <- paste0("Evaluation of forecast (", type, " model)")
[…]

### 2021-03-18 src/model2/v13/icu_regression.R

@@ -26,7 +26,7 @@ icu_covariates <- function(deaths, icu){
     if(file.exists(icu_cov_name)==T)
         icu_cov  <- readRDS(icu_cov_name)
 
-    icu_pre <-read_excel(path = file.path("data", "Folkhalsomyndigheten_Covid19.xlsx"),
+    icu_pre <-read_excel(path = file.path("data", "FHM", "Folkhalsomyndigheten_Covid19_2020-12-09.xlsx"),
                          sheet = 3,
                          col_types = c("date", "numeric"),
                          progress = FALSE)
@@ -52,9 +52,9 @@ icu_covariates <- function(deaths, icu){
 
         data.list <- list(date=1:51,
                           Y = icu_est[1:51])
-        model.fit <- gam(Y~s(date), data=data.list, family = poisson)
+        model.fit <- gam(Y~  s(date), data=data.list, family = poisson)
         icu_cov = data.frame(date=dates[1:51],
-                            theta=log(model.fit$fitted.values)[1:51])
+                             theta=log(model.fit$fitted.values)[1:51])
     }
 
     for(i in 52:length(icu_est)){
@@ -65,7 +65,8 @@ icu_covariates <- function(deaths, icu){
         data.list <- list(date=1:i,
                           Y = icu_est[1:i])
 
-        model.fit <- gam(Y~s(date), data=data.list, family = poisson)
+        model.fit <- gam(Y~  s(date), data=data.list, family = nb(), method="REML")
+        print(summary(model.fit))
         icu_cov <- rbind(icu_cov,
                          c(as.character(dates[i]),
                            log(model.fit$fitted.values)[i]))
@@ -85,7 +86,7 @@ icu_covariates <- function(deaths, icu){
                                dates = deaths$dates[1:(N_d-incomplete_deaths)])
     icu_cov_glm <- icu_lag[dates%in%Cov_analysis$dates]
 
-    glm.fit <- glm(y~x,poisson, data= data.frame(y=Cov_analysis$y,x=icu_cov_glm))
+    glm.fit <- gam(y~x,family = nb(), data= data.frame(y=Cov_analysis$y,x=icu_cov_glm), method = "REML")
 
 
 
@@ -94,8 +95,11 @@ icu_covariates <- function(deaths, icu){
 
     return(list(theta_cov = data.frame(dates = dates,
                                        theta = icu_lag),
-                coeff     = glm.fit$coefficients))
+                coeff     = glm.fit$coefficients,
+                model     = glm.fit))
 }
+
+
 
 
 

### 2021-03-18 src/model2/v13/sluten_regression.R

@@ -1,0 +1,64 @@
+##
+# analysis of slutenvard
+#
+##
+library(gamm4)
+
+source(file.path("src","model2","v12","icu_regression.R"))
+lag <- 7
+deaths <- readRDS(file.path("data", "processed", "processed_data.rds"))
+data.vard <- read.csv(file.path("data", "vard", "data-fJECo.csv"))
+
+n = dim(data.vard)[1]
+slutenvard <- as.numeric(data.vard[,3])
+intensvard <-  as.numeric(data.vard[,5])
+days       <- 1:length(slutenvard)
+data.list.sluten <- data.frame(days= days[!is.na(slutenvard)],
+                        date = as.Date(data.vard[!is.na(slutenvard),1]),
+                        sluten   = slutenvard[!is.na(slutenvard)])
+model.fit.sluten <- gam(sluten~  s(days), data=data.list.sluten, family = nb(),method="REML" )
+data.list.iva <- data.frame(days= days[!is.na(intensvard)],
+                        date = as.Date(data.vard[!is.na(intensvard),1]),
+                        iva   = intensvard[!is.na(intensvard)])
+model.fit.iva <- gam(iva~  s(days), data=data.list.iva, family = nb(),method="REML" )
+
+
+cov.data <- data.frame( date  =as.Date(data.vard[,1]),
+                        theta_sluten = predict(model.fit.sluten,newdata = data.frame(days = days)),
+                        theta_iva = predict(model.fit.iva,newdata = data.frame(days = days)))
+Y <- apply(deaths$detected,1,max,na.rm=T)
+death.list = data.frame( date = as.Date(deaths$dates), death = Y)
+
+lags <- 1:20
+lags.iva    <- rep(NA,length(lags))
+lags.sluten <- rep(NA,length(lags))
+for(i in 1:length(lags)){
+    lag <- lags[i]
+    cov.data.lag = data.frame( date         = c(cov.data$date, cov.data$date[n]+1:lag),
+                               theta_sluten = c(rep(0,lag),cov.data$theta_sluten),
+                               theta_iva    = c(rep(0,lag),cov.data$theta_iva))
+
+    data.total <- merge(cov.data.lag,death.list, by ="date")
+
+    data.total.rem <- data.total[30:(dim(data.total)[1]-15),]
+
[…]

### 2021-03-18 src/model2/v2/analysis.full.R

@@ -16,7 +16,7 @@ data$df <- as.Date(data$State)-as.Date(data$dates)
 days = 1:14
 
 data <- data[order(data$State),]
-data <- data[1:896,]
+#data <- data[1:200,]
 for(day in days){
     index <- data$df==day
     data_temp <- data[index,]

### 2021-03-18 src/model2/v2/full.run.R

@@ -9,7 +9,7 @@ library(foreach)
 library(doParallel)
 days_run <- 30
 sim <- 40000
-n.clust <- 3
+n.clust <- 4
 store_data_folder <- file.path("data","tmp","model2","v2")
 model_parameters <- list(sim           = sim,
                          burnin        = ceiling(0.5*sim),
@@ -28,32 +28,35 @@ prior_list0 <- list(mu_beta        = c(0,0,0),
 N_est_true <- apply(data$detected,1,max, na.rm=T)
 N.obs <- length(data$dates)
 
-cl <- parallel::makeCluster(n.clust)
+cl <- parallel::makeCluster(n.clust,outfile="")
 doParallel::registerDoParallel(cl)
-foreach(j = (days_run+1):(N.obs-30)) %dopar%{
+foreach(j = (days_run+1):(N.obs)) %dopar%{
     library(invgamma)
     start_ = j - days_run # run the last 31 days
+    cat('running day  = ', as.character(data$dates[j]),'\n',sep="")
+    day_is_run = file.exists(paste(store_data_folder,"/Npost_",j,'.rds',sep=""))
+    if(day_is_run==F){
+        report_cleaned <- report_clean(data$detected[start_:j,start_:j],data$dates[start_:j])
+        new_cases <- newCases(report_cleaned)
+        rownames(new_cases) <- as.character(data$dates[start_:j])
+        colnames(new_cases) <- as.character(data$dates[start_:j])
 
-    report_cleaned <- report_clean(data$detected[start_:j,start_:j],data$dates[start_:j])
-    new_cases <- newCases(report_cleaned)
-    rownames(new_cases) <- as.character(data$dates[start_:j])
-    colnames(new_cases) <- as.character(data$dates[start_:j])
 
+        if(start_ <= days_run+ 1){
+            prior_list <- prior_list0
+            prior_list$n_obs <- 0
+        }else{
+            prior_list <- readRDS(paste(store_data_folder,"/prior_",start_-1,'.rds',sep=""))
+        }
 
-    if(start_ <= days_run+ 1){
-        prior_list <- prior_list0
-        prior_list$n_obs <- 0
-    }else{
-        prior_list <- readRDS(paste(store_data_folder,"/prior_",start_-1,'.rds',sep=""))
+        result <- model(new_cases, model_parameters, prior_list)
+        result$posteriror_list$n_obs <- prior_list$n_obs + days_run
[…]

### 2021-03-18 src/model2/v6/analysis.full.R

@@ -29,7 +29,7 @@ for(day in days){
     cat(' (',round(mean(CI_under),2),',',round(mean(CI_over),2),")" ,sep="")
     cat('\n')
 }
-day_ <- 7
+day_ <- 1
 index <- data$df==day_
 data_temp <- data[index,]
 #data_temp <- data_temp[80:sum(index),]

### 2021-03-18 output/paper/paper.tex

--- 2020-06-16 output/paper/paper.tex
@@
     \begin{abstract}
-        The new coronavirus disease --- COVID-2019 --- is rapidly spreading through the world.
+        The new corona virus disease --- COVID-2019 --- is rapidly spreading through the world.
 The availability of unbiased timely statistics of trends in disease events are a key to effective responses.
@@
 \section{Pandemic response demands timely data}
-The SARS-CoV-2 pandemic is affecting societies all around the world.
+The new corona virus pandemic is affecting societies all around the world.
 As countries are challenged to control and fight back, they are in need of timely, unbiased, data for monitoring trends and making fast and well-informed decisions \citep{No_author_2020_coronavirus_three}.
@@
 As expected, the model provides estimates of actual deaths considerably above the reported number of deaths.
-Note how the model predicts additional deaths above the moving average line.
+Not how the model predicts additional deaths above the moving average line.
 
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{../plots/latest_prediction}
+    \includegraphics[width=0.9\textwidth]{plots/latest_prediction}
     \caption{Swedish Covid-19 deaths as of 2020-06-04 and model predictions}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{../plots/lag_prediction_by_date}
+    \includegraphics[width=0.9\textwidth]{plots/lag_prediction_by_date}
     \caption{Model accuracy over time for four dates, compared to benchmark}
@@
     \centering
-    \includegraphics[width=0.9\textwidth]{../plots/model_metrics}
+    \includegraphics[width=0.9\textwidth]{plots/model_metrics}
     \caption{Model accuracy over time for four dates, compared to a benchmark}
@@
 In Sweden, delays are considerable and exhibit a weekday and holiday pattern that need to be accounted for to draw conclusions from the data.
-The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the SARS-CoV-2 virus pandemic.
+The method and algorithm proposed overcomes major shortcomings in the daily interpretation and practice analyzing and controlling the novel Corona virus pandemic.
 It also provides valuable measures of uncertainty around these estimates, showing users how large the range of possible outcomes can be.
@@
 Given there exists many regions in Sweden this is unlikely to be the case.
-For example, it is easy to see that the Swedish region Västra Götland follows a differewnt reporting structure than Stockholm.
+For example, it is easy to see that the Swedish region Västra Götland follows a different reporting structure than Stockholm.
 Building a model for each region separately would most likely give better results and make the assumptions more reasonable.
@@
 
-Moreover, deaths seem to be reported from two distinct populations that follow different trends.
+Moreover, deaths are reported from two distinct populations that seem to follow different trends.
 At the time of writing, the daily deaths in elderly care, reported with a longer delay, seem to be decreasing slower than hospital deaths.
@@
 However, we have noted a clear decline in proportions of deaths reported the two first working days.
-For example, out of all deaths that occurred on the second of April, $\approx 30\%$ were reported within the first two workdays, whereas for the eighteenth of May only $\approx 10\%$ were reported that fast.
-We address this by assuming that the deaths reported during the two first working days comes from a different population than the remainder of days.
+For example the number of deaths occurring at the second of April $\approx 30\%$ of deaths where reported within the first two working days whereas for the eighteens of May only $\approx 10\%$ where reported during the two first working days.
+We address this by assuming that the deaths reported during the two first working days comes from a different population then the remainder of days.
 

### 2021-05-27 src/constant_benchmark.R
set CRPS to zero

@@ -79,7 +79,7 @@ generate_predictions <- function(deaths_dt) {
     predictions[, `:=`(predicted_deaths_cum = reported_dead + cumsum(tmp.deaths_added),
                        predicted_deaths_SD_cum = sqrt(cumsum(tmp.deaths_added_sd^2))), # assuming independently normal
                 by = .(state, date)]
-    predictions[predicted_deaths_SD_cum == 0, predicted_deaths_SD_cum := NA_real_]
+    #predictions[predicted_deaths_SD_cum == 0, predicted_deaths_SD_cum := NA_real_]
 
     # Don't include predictions states where we already know more than we are predicting
     predictions <- predictions[state - date < prediction_date - date + 1]
@@ -112,6 +112,7 @@ out <- DT[, .(state,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
 
+out[is.nan(CRPS), CRPS := 0]
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_SWE.fst"))
 
 ######
@@ -135,5 +136,6 @@ out <- DT[, .(state,
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
+out[is.nan(CRPS), CRPS := 0]
 
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_UK.fst"))

### 2021-05-27 src/plots.R
latest prediction cowplot

@@ -20,56 +20,89 @@ library(forcats)
 library(ggplot2)
 library(hrbrthemes)
 library(wesanderson)
-source(file.path("src", "util","functions.R"))
+library(extrafont)
+library(cowplot)
+source(file.path("src", "util", "functions.R"))
 
 #
 w <- 11 # plot width (inches)
 my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000")
 
 # Plot 1 = Predictions and current stats
-deaths_dt <- read_fst(file.path("data", "processed", "deaths_dt_SWE.fst"), as.data.table = TRUE)
-model_predict <- read_fst(file.path("data", "processed", "model_latest.fst"), as.data.table = TRUE)
-DT1 <- model_predict[date %between% c("2020-05-20", "2020-06-03")]
-DT2 <- deaths_dt[!is.na(N) & !is.na(date) & publication_date == max(model_predict[, date])]
-DT2[, avg := frollmean(N, 7, algo = "exact", align = "right")]
-DT2 <- DT2[date > "2020-03-15"]
+latest_prediction <- function(deaths_dt, model_predict) {
+    # Predict over the last 15 days, plot last two months
+    plot_state <- model_predict[, max(state)]
+    model_predict <- model_predict[state == plot_state & date %between% c(plot_state - 15, plot_state)]
 
-colors <- c("gray80", "#ECCBAE")
-colors <- setNames(colors, c("Reported dead", "Model prediction"))
+    deaths_dt[, avg := frollmean(N, 7, algo = "exact", align = "center"), by = publication_date]
+    deaths_dt <- deaths_dt[publication_date == plot_state & date %between% c(plot_state - 60, plot_state)]
 
-plot <- ggplot(data = DT1, aes(x = date, y = predicted_deaths)) +
-    geom_bar(aes(fill = "Model prediction"), stat = "identity") +
-    geom_bar(data = DT2,
-             aes(y = N, fill = "Reported dead"), stat = "identity") +
-    geom_line(data = DT2[!is.na(avg) & date <= max(model_predict[, date]) - 10],
-              aes(y = avg, linetype = "7-day moving average"), color = "grey50") +
-    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
-                  width = 0.4, color = "#e0ac7e") +
-    set_default_theme() +
-    scale_fill_manual(values = colors) +
[…]

### 2021-05-27 src/constant_benchmark.R
crps at sd 0 should be mean error

@@ -112,7 +112,7 @@ out <- DT[, .(state,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
 
-out[is.nan(CRPS), CRPS := 0]
+out[is.nan(CRPS), CRPS := abs(predicted_deaths - target)]
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_SWE.fst"))
 
 ######
@@ -136,6 +136,6 @@ out <- DT[, .(state,
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
-out[is.nan(CRPS), CRPS := 0]
+out[is.nan(CRPS), CRPS := abs(predicted_deaths - target)]
 
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_UK.fst"))

### 2021-05-27 src/constant_benchmark.R
neg crps

@@ -112,7 +112,7 @@ out <- DT[, .(state,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
 
-out[is.nan(CRPS), CRPS := abs(predicted_deaths - target)]
+out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_SWE.fst"))
 
 ######
@@ -136,6 +136,6 @@ out <- DT[, .(state,
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
-out[is.nan(CRPS), CRPS := abs(predicted_deaths - target)]
+out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
 
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_UK.fst"))

### 2021-05-27 src/plots.R
new daily prediction plots

@@ -25,7 +25,6 @@ library(cowplot)
 source(file.path("src", "util", "functions.R"))
 
 #
-w <- 11 # plot width (inches)
 my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000")
 
 # Plot 1 = Predictions and current stats
@@ -65,13 +64,13 @@ latest_prediction <- function(deaths_dt, model_predict) {
 deaths_dt_SWE <- read_fst(file.path("data", "processed", "deaths_dt_SWE.fst"), as.data.table = TRUE)
 model_predict_SWE <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_SWE.fst"), as.data.table = TRUE)
 plot_SWE  <- latest_prediction(deaths_dt_SWE, model_predict_SWE)
-# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_SWE.pdf"),
+# ggsave2(filename = file.path("output", "paper", "plots", "latest_prediction_SWE.pdf"),
 #         plot = plot_SWE, device = cairo_pdf, width = w, height = w/1.9)
 
 deaths_dt_UK <- read_fst(file.path("data", "processed", "deaths_dt_UK.fst"), as.data.table = TRUE)
 model_predict_UK <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_UK.fst"), as.data.table = TRUE)
 plot_UK <- latest_prediction(deaths_dt_UK, model_predict_UK)
-# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_UK.pdf"),
+# ggsave2(filename = file.path("output", "paper", "plots", "latest_prediction_UK.pdf"),
 #         plot = plot_UK, device = cairo_pdf, width = w, height = w/1.9)
 
 p <- plot_grid(plot_grid(
@@ -86,41 +85,50 @@ p <- plot_grid(plot_grid(
 save_plot(filename = file.path("output", "paper", "plots", "latest_prediction.pdf"),
           plot = p, ncol = 2, nrow = 2, base_height = 2.5, device = cairo_pdf)
 
-break
-
 # Load data
-data/processed/constant_model_predictions_full_SWE.fst
-
-benchmark_SWE <- read_fst(file.path("data", "processed", "constant_model_predictions_full_SWE.fst"), as.data.table = TRUE)
+min_date <- as.Date("2020-10-10")
[…]

### 2021-06-07 src/plots.R
rerun with joined plots and cleanup

@@ -70,8 +70,8 @@ model_predict_UK <- read_fst(file.path("data", "processed", "model_predictions_f
 plot_UK <- latest_prediction(deaths_dt_UK, model_predict_UK)
 
 p <- plot_grid(plot_grid(
-    plot_SWE + guides(fill = "none", linetype = "none"),
-    plot_UK + guides(fill = "none", linetype = "none") + ylab(NULL),
+    plot_SWE + guides(fill = "none", linetype = "none") + theme(plot.margin = unit(c(25, 5, 5, 5), "pt")),
+    plot_UK + guides(fill = "none", linetype = "none") + ylab(NULL) + theme(plot.margin = unit(c(25, 5, 5, 5), "pt")),
     labels = c("Sweden", "United Kingdom"), label_fontfamily = "EB Garamond",
     hjust = -0.5, align = "hv",
     nrow = 1, ncol = 2),
@@ -79,7 +79,7 @@ p <- plot_grid(plot_grid(
     nrow = 2, ncol = 1, rel_heights = c(0.9, 0.1)
 )
 save_plot(filename = file.path("output", "paper", "plots", "latest_prediction.pdf"),
-          plot = p, ncol = 2, nrow = 2, base_height = 2.5, device = cairo_pdf)
+          plot = p, ncol = 2, nrow = 1, base_height = 4, device = cairo_pdf)
 
 # Load data
 min_date <- as.Date("2020-10-10")
@@ -156,7 +156,7 @@ day_plot <- function(DT, reported, plot.title) {
         scale_x_reverse(breaks = scales::extended_breaks(), expand = expansion(add = c(1, 1))) +
         labs(title = plot.title,
              color = "Model",
-             x = "Days of lag to predict",
+             x = "Days of lag to predict", # lag prediction distance (days)
              y = "Number of deaths")
 
     if (DT[, uniqueN(date)] > 1) {
@@ -174,21 +174,33 @@ plot_data_UK[, type := factor(type)]
 
 # Figure 1 - Pick 4 dates at random to plot
 set.seed(1234)
-example_dates <- sample(seq(model_UK[, min(date)], model_SWE[, max(date)], 1), 4)
-plot <- day_plot(plot_data_SWE[date %in% example_dates],
+example_dates <- sample(seq(model_UK[, min(date)], model_SWE[, max(date)], 1), 3)
+plot_SWE <- day_plot(plot_data_SWE[date %in% example_dates],
                  reported_dead_SWE[date %in% example_dates],
[…]

### 2021-06-07 output/paper/paper.tex
Updates from Overleaf

@@
     \centering
-    \includegraphics[width=0.9\textwidth]{plots/latest_prediction}
+    \includegraphics[width=0.95\textwidth]{plots/latest_prediction}
     \caption{Swedish Covid-19 deaths as of 2020-06-04 and model predictions}
@@
     \includegraphics[width=0.9\textwidth]{plots/lag_prediction_by_date}
-    \caption{Model accuracy over time for four dates, compared to benchmark}
+    \caption{Model accuracy over time for three randomly chosen dates, compared to the benchmark model}
     \label{fig:four_dates}
@@
 Each data point is the average of all dates where predictions can be evaluated.
-SCRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}.
+CRPS is a measure of accuracy that rewards precision, it is a proper scoring rule like the continuous probability rank score or the Brier score (see definition in Appendix) \citep{bolin2019scale}.
 The central plot shows the width of the prediction intervals, and the rightmost one the proportion of PI\'s that cover the true value.

### 2021-06-07 src/constant_benchmark.R

@@ -107,12 +107,12 @@ out <- DT[, .(state,
               predicted_deaths_SD = predicted_deaths_SD_cum,
               reported_dead,
               target,
-              days_left = as.integer(30 - as.integer(state - date)),
+              days_left = 30 - as.integer(state - date),
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
 
-out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
+out[is.nan(CRPS), CRPS := 0]
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_SWE.fst"))
 
 ######
@@ -132,10 +132,10 @@ out <- DT[, .(state,
               predicted_deaths_SD = predicted_deaths_SD_cum,
               reported_dead,
               target,
-              days_left = as.integer(30 - as.integer(state - date)),
+              days_left = 30 - as.integer(state - date),
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
-out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
+out[is.nan(CRPS), CRPS := 0]
 
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_UK.fst"))

### 2021-06-07 src/model_predictions.R

@@ -26,8 +26,7 @@ buildData("uk", file.path("data", "processed", "processed_data_uk.rds"))
 source(file.path("src","util","util_swe.r"))
 swe_data <- readRDS(file.path("data", "processed", "processed_data_sweden.rds"))
 
-#remove data before 2020-07-01
-# (from the first wave as reporting style was different then)
+#remove data before  2020-07-01
 start_date <- as.Date("2020-07-01")
 index <- swe_data$dates_report >= start_date
 swe_data$detected             = swe_data$detected[index,index]
@@ -42,7 +41,7 @@ target_swe$dates <- swe_data$dates_report[1:length(target_swe$reported)]
 swe_data$detected[row(swe_data$detected)+max.days.to.report<col(swe_data$detected)]=NA
 
 dts_swe <- lapply(
-    as.Date(swe_data$dates_report[swe_data$dates_report >= "2020-10-10"]),
+    as.Date(swe_data$dates_report[swe_data$dates_report >= start_date + 90]),
     FUN = function(x, ...) swe.prediction(report.dates = x, ...),
     max.days.to.report = max.days.to.report,
     result = swe_data,

### 2021-06-07 src/plots.R

@@ -8,9 +8,9 @@
 # data/processed/model_benchmark.fst.fst (Beta_GP_lag_benchmark_build.R)
 #
 # generates files:
-# output/plots/CRPS_over_weekdays.pdf
+# output/plots/SCRPS_over_weekdays.pdf
 # output/plots/daily/prediction_YYYY-MM-DD.pdf
-# output/plots/CRPS_over_states.pdf
+# output/plots/SCRPS_over_states.pdf
 # output/plots/model_metrics.pdf
 ##
 
@@ -25,6 +25,7 @@ library(cowplot)
 source(file.path("src", "util", "functions.R"))
 
 #
+w <- 11 # plot width (inches)
 my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000")
 
 # Plot 1 = Predictions and current stats
@@ -64,14 +65,18 @@ latest_prediction <- function(deaths_dt, model_predict) {
 deaths_dt_SWE <- read_fst(file.path("data", "processed", "deaths_dt_SWE.fst"), as.data.table = TRUE)
 model_predict_SWE <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_SWE.fst"), as.data.table = TRUE)
 plot_SWE  <- latest_prediction(deaths_dt_SWE, model_predict_SWE)
+# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_SWE.pdf"),
+#         plot = plot_SWE, device = cairo_pdf, width = w, height = w/1.9)
 
 deaths_dt_UK <- read_fst(file.path("data", "processed", "deaths_dt_UK.fst"), as.data.table = TRUE)
 model_predict_UK <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_UK.fst"), as.data.table = TRUE)
 plot_UK <- latest_prediction(deaths_dt_UK, model_predict_UK)
+# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_UK.pdf"),
+#         plot = plot_UK, device = cairo_pdf, width = w, height = w/1.9)
 
 p <- plot_grid(plot_grid(
-    plot_SWE + guides(fill = "none", linetype = "none") + theme(plot.margin = unit(c(25, 5, 5, 5), "pt")),
-    plot_UK + guides(fill = "none", linetype = "none") + ylab(NULL) + theme(plot.margin = unit(c(25, 5, 5, 5), "pt")),
[…]

### 2021-06-07 src/swe_prediction.R

@@ -27,11 +27,11 @@ result$detected[row(result$detected)+max.days.to.report<col(result$detected)]=NA
 
 pred <- swe.prediction(result = result,
               max.days.to.report = max.days.to.report,
-              report.dates = c(as.Date("2020-11-10")),
+              report.dates = c(as.Date("2021-03-20")),
               target= target)
 pred.smooth <- gp.smooth(pred,
                          max.days.to.report = max.days.to.report)
-plot(pred$date,pred$ci_lower, type='l',col='blue',ylim=c(0,max(pred.smooth$ci_upper)*2))
+plot(pred$date,pred$ci_lower, type='l',col='blue',ylim=c(0,140))
 lines(pred$date,pred$ci_upper, type='l',col='red')
 
 lines(pred.smooth$date,pred.smooth$ci_lower, type='l',col='green')

### 2021-06-07 src/uk_prediction.R

@@ -20,14 +20,13 @@ result$detected[row(result$detected)+max.days.to.report<col(result$detected)]=NA
 
 pred <- uk.prediction(result = result,
               max.days.to.report = max.days.to.report,
-              report.dates = c(as.Date("2020-11-26")),
+              report.dates = c(as.Date("2020-11-01")),
               target= target)
 pred.smooth <- gp.smooth(pred,
                          max.days.to.report = max.days.to.report)
-plot(pred$date,pred$target,ylim=c(0,max(pred.smooth$ci_upper)))
+plot(pred$date,pred$target,ylim=c(0,max(pred.smooth$ci_lower)))
 lines(pred$date,pred$predicted_deaths, type='l',col='blue')
 lines(pred$date,pred$ci_upper, type='l',col='red')
-lines(pred$date,pred$ci_lower, type='l',col='red')
 
 lines(pred.smooth$date,pred.smooth$ci_upper, type='l',col='green')
 lines(pred.smooth$date,pred.smooth$ci_lower, type='l',col='green')

### 2021-06-07 src/util/util_swe.R

@@ -882,9 +882,7 @@ swe.prediction <- function(result,
                 }
             }
         }
-        zero_dates <- result_i$dates[rowSums(result_i$detected, na.rm=T)==0] # take out zero reporitng
-        zero_dates <- zero_dates[report.dates[i]-zero_dates < 10] # if more then ten days prob zero
-        zero.report = state%in%zero_dates
+        zero.report = state%in%result_i$dates[rowSums(result_i$detected, na.rm=T)==0]
         deaths.i <- data.table(
             date          = state,
             state         =  report.dates[i],
@@ -914,16 +912,15 @@ swe.prediction <- function(result,
 ##
 likelihood_death <- function(x, D, y){
     kappa       <- exp(x[1])
-    nu          <- 3/(1+exp(x[2]))
+    nu          <- exp(x[2])
     sigma       <- exp(x[3])
     sigma_noise <- exp(x[4])
-    mu          <- x[5]
     Sigma <- matern.covariance(D, kappa, nu, sigma)
     diag(Sigma)<- diag(Sigma) + sigma_noise^2
     R <- chol(Sigma, pivot=T)
     if(attr(R,"rank" )< length(y))
         return(-Inf)
-    v <- solve(t(R),y[attr(R,"pivot")] - mu)
+    v <- solve(t(R),y[attr(R,"pivot")])
     return( -sum(log(diag(R))) - 0.5*t(v)%*%v)
 }
 
@@ -940,13 +937,9 @@ likelihood_death <- function(x, D, y){
 
 gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_width = 0.95){
 
-    #transform <- function(x){log(pmax(x,1))}
-    #itransform <- function(x){exp(x)}
-    transform <- function(x){x^{1/2}}
-    itransform <- function(x){x^2}
 
     if(is.null(theta)){
-        theta <- c(0,0,0,0,0)
+        theta <- c(0,0,0,0)
     }
 
     report.date <- unique(death_prediction$state)
@@ -965,55 +958,54 @@ gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_wid
                                         date    = death_prediction$date[index.known])
 
[…]

### 2021-06-07 src/util/util_uk.R

@@ -907,9 +907,7 @@ uk.prediction <- function(result,
                 }
             }
         }
-        zero_dates <- result_i$dates[rowSums(result_i$detected, na.rm=T)==0] # take out zero reporitng
-        zero_dates <- zero_dates[report.dates[i]-zero_dates < 10] # if more then ten days prob zero
-        zero.report = state%in%zero_dates
+        zero.report = state%in%result_i$dates[rowSums(result_i$detected, na.rm=T)==0]
         deaths.i <- data.table(
             date          = state,
             state         =  report.dates[i],
@@ -938,17 +936,16 @@ uk.prediction <- function(result,
 #' @param y data
 ##
 likelihood_death <- function(x, D, y){
-    kappa       <- exp(x[1]) + 10^-2
-    nu          <- 3/(1+exp(x[2]))
+    kappa       <- exp(x[1])
+    nu          <- exp(x[2])
     sigma       <- exp(x[3])
     sigma_noise <- exp(x[4])
-    mu          <- 0#x[5]
     Sigma <- matern.covariance(D, kappa, nu, sigma)
     diag(Sigma)<- diag(Sigma) + sigma_noise^2
     R <- chol(Sigma, pivot=T)
     if(attr(R,"rank" )< length(y))
         return(-Inf)
-    v <- solve(t(R),y[attr(R,"pivot")]-mu)
+    v <- solve(t(R),y[attr(R,"pivot")])
     return( -sum(log(diag(R))) - 0.5*t(v)%*%v)
 }
 
@@ -965,10 +962,7 @@ likelihood_death <- function(x, D, y){
 
 gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_width = 0.95){
 
-    transform <- function(x){log(pmax(x,1))}
-    itransform <- function(x){exp(x)}
-    #transform <- function(x){x^{1/3}}
-    #itransform <- function(x){x^3}
+
     if(is.null(theta)){
         theta <- c(0,0,0,0)
     }
@@ -990,7 +984,7 @@ gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_wid
                                         date    = death_prediction$date[index.known])
 
[…]

### 2021-06-07 src/constant_benchmark.R
Revert "Updates from Overleaf"
This reverts commit 0495f56240ec6e3eeaa9a3a5d8b2baf2cd8ca8b8.

@@ -107,12 +107,12 @@ out <- DT[, .(state,
               predicted_deaths_SD = predicted_deaths_SD_cum,
               reported_dead,
               target,
-              days_left = 30 - as.integer(state - date),
+              days_left = as.integer(30 - as.integer(state - date)),
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
 
-out[is.nan(CRPS), CRPS := 0]
+out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_SWE.fst"))
 
 ######
@@ -132,10 +132,10 @@ out <- DT[, .(state,
               predicted_deaths_SD = predicted_deaths_SD_cum,
               reported_dead,
               target,
-              days_left = 30 - as.integer(state - date),
+              days_left = as.integer(30 - as.integer(state - date)),
               ci_upper = predicted_deaths_cum + qnorm(1-alpha/2) * predicted_deaths_SD_cum,
               ci_lower = predicted_deaths_cum + qnorm(alpha/2) * predicted_deaths_SD_cum,
               CRPS = CRPS(target, predicted_deaths_cum, predicted_deaths_SD_cum))]
-out[is.nan(CRPS), CRPS := 0]
+out[is.nan(CRPS), CRPS := - abs(predicted_deaths - target)]
 
 write_fst(out, file.path("data", "processed", "constant_model_predictions_full_UK.fst"))

### 2021-06-07 src/model_predictions.R

@@ -26,7 +26,8 @@ buildData("uk", file.path("data", "processed", "processed_data_uk.rds"))
 source(file.path("src","util","util_swe.r"))
 swe_data <- readRDS(file.path("data", "processed", "processed_data_sweden.rds"))
 
-#remove data before  2020-07-01
+#remove data before 2020-07-01
+# (from the first wave as reporting style was different then)
 start_date <- as.Date("2020-07-01")
 index <- swe_data$dates_report >= start_date
 swe_data$detected             = swe_data$detected[index,index]
@@ -41,7 +42,7 @@ target_swe$dates <- swe_data$dates_report[1:length(target_swe$reported)]
 swe_data$detected[row(swe_data$detected)+max.days.to.report<col(swe_data$detected)]=NA
 
 dts_swe <- lapply(
-    as.Date(swe_data$dates_report[swe_data$dates_report >= start_date + 90]),
+    as.Date(swe_data$dates_report[swe_data$dates_report >= "2020-10-10"]),
     FUN = function(x, ...) swe.prediction(report.dates = x, ...),
     max.days.to.report = max.days.to.report,
     result = swe_data,

### 2021-06-07 src/plots.R

@@ -8,9 +8,9 @@
 # data/processed/model_benchmark.fst.fst (Beta_GP_lag_benchmark_build.R)
 #
 # generates files:
-# output/plots/SCRPS_over_weekdays.pdf
+# output/plots/CRPS_over_weekdays.pdf
 # output/plots/daily/prediction_YYYY-MM-DD.pdf
-# output/plots/SCRPS_over_states.pdf
+# output/plots/CRPS_over_states.pdf
 # output/plots/model_metrics.pdf
 ##
 
@@ -25,7 +25,6 @@ library(cowplot)
 source(file.path("src", "util", "functions.R"))
 
 #
-w <- 11 # plot width (inches)
 my_palette <- c("#d1ae90", "#046C9A", "#D69C4E", "#ABDDDE", "#000000")
 
 # Plot 1 = Predictions and current stats
@@ -65,18 +64,14 @@ latest_prediction <- function(deaths_dt, model_predict) {
 deaths_dt_SWE <- read_fst(file.path("data", "processed", "deaths_dt_SWE.fst"), as.data.table = TRUE)
 model_predict_SWE <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_SWE.fst"), as.data.table = TRUE)
 plot_SWE  <- latest_prediction(deaths_dt_SWE, model_predict_SWE)
-# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_SWE.pdf"),
-#         plot = plot_SWE, device = cairo_pdf, width = w, height = w/1.9)
 
 deaths_dt_UK <- read_fst(file.path("data", "processed", "deaths_dt_UK.fst"), as.data.table = TRUE)
 model_predict_UK <- read_fst(file.path("data", "processed", "model_predictions_full_smooth_UK.fst"), as.data.table = TRUE)
 plot_UK <- latest_prediction(deaths_dt_UK, model_predict_UK)
-# ggsave(filename = file.path("output", "paper", "plots", "latest_prediction_UK.pdf"),
-#         plot = plot_UK, device = cairo_pdf, width = w, height = w/1.9)
 
 p <- plot_grid(plot_grid(
-    plot_SWE + guides(fill = "none", linetype = "none"),
-    plot_UK + guides(fill = "none", linetype = "none") + ylab(NULL),
+    plot_SWE + guides(fill = "none", linetype = "none") + theme(plot.margin = unit(c(25, 5, 5, 5), "pt")),
[…]

### 2021-06-07 src/swe_prediction.R

@@ -27,11 +27,11 @@ result$detected[row(result$detected)+max.days.to.report<col(result$detected)]=NA
 
 pred <- swe.prediction(result = result,
               max.days.to.report = max.days.to.report,
-              report.dates = c(as.Date("2021-03-20")),
+              report.dates = c(as.Date("2020-11-10")),
               target= target)
 pred.smooth <- gp.smooth(pred,
                          max.days.to.report = max.days.to.report)
-plot(pred$date,pred$ci_lower, type='l',col='blue',ylim=c(0,140))
+plot(pred$date,pred$ci_lower, type='l',col='blue',ylim=c(0,max(pred.smooth$ci_upper)*2))
 lines(pred$date,pred$ci_upper, type='l',col='red')
 
 lines(pred.smooth$date,pred.smooth$ci_lower, type='l',col='green')

### 2021-06-07 src/uk_prediction.R

@@ -20,13 +20,14 @@ result$detected[row(result$detected)+max.days.to.report<col(result$detected)]=NA
 
 pred <- uk.prediction(result = result,
               max.days.to.report = max.days.to.report,
-              report.dates = c(as.Date("2020-11-01")),
+              report.dates = c(as.Date("2020-11-26")),
               target= target)
 pred.smooth <- gp.smooth(pred,
                          max.days.to.report = max.days.to.report)
-plot(pred$date,pred$target,ylim=c(0,max(pred.smooth$ci_lower)))
+plot(pred$date,pred$target,ylim=c(0,max(pred.smooth$ci_upper)))
 lines(pred$date,pred$predicted_deaths, type='l',col='blue')
 lines(pred$date,pred$ci_upper, type='l',col='red')
+lines(pred$date,pred$ci_lower, type='l',col='red')
 
 lines(pred.smooth$date,pred.smooth$ci_upper, type='l',col='green')
 lines(pred.smooth$date,pred.smooth$ci_lower, type='l',col='green')

### 2021-06-07 src/util/util_swe.R

@@ -882,7 +882,9 @@ swe.prediction <- function(result,
                 }
             }
         }
-        zero.report = state%in%result_i$dates[rowSums(result_i$detected, na.rm=T)==0]
+        zero_dates <- result_i$dates[rowSums(result_i$detected, na.rm=T)==0] # take out zero reporitng
+        zero_dates <- zero_dates[report.dates[i]-zero_dates < 10] # if more then ten days prob zero
+        zero.report = state%in%zero_dates
         deaths.i <- data.table(
             date          = state,
             state         =  report.dates[i],
@@ -912,15 +914,16 @@ swe.prediction <- function(result,
 ##
 likelihood_death <- function(x, D, y){
     kappa       <- exp(x[1])
-    nu          <- exp(x[2])
+    nu          <- 3/(1+exp(x[2]))
     sigma       <- exp(x[3])
     sigma_noise <- exp(x[4])
+    mu          <- x[5]
     Sigma <- matern.covariance(D, kappa, nu, sigma)
     diag(Sigma)<- diag(Sigma) + sigma_noise^2
     R <- chol(Sigma, pivot=T)
     if(attr(R,"rank" )< length(y))
         return(-Inf)
-    v <- solve(t(R),y[attr(R,"pivot")])
+    v <- solve(t(R),y[attr(R,"pivot")] - mu)
     return( -sum(log(diag(R))) - 0.5*t(v)%*%v)
 }
 
@@ -937,9 +940,13 @@ likelihood_death <- function(x, D, y){
 
 gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_width = 0.95){
 
+    #transform <- function(x){log(pmax(x,1))}
+    #itransform <- function(x){exp(x)}
+    transform <- function(x){x^{1/2}}
+    itransform <- function(x){x^2}
 
     if(is.null(theta)){
-        theta <- c(0,0,0,0)
+        theta <- c(0,0,0,0,0)
     }
 
     report.date <- unique(death_prediction$state)
@@ -958,54 +965,55 @@ gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_wid
                                         date    = death_prediction$date[index.known])
 
[…]

### 2021-06-07 src/util/util_uk.R

@@ -907,7 +907,9 @@ uk.prediction <- function(result,
                 }
             }
         }
-        zero.report = state%in%result_i$dates[rowSums(result_i$detected, na.rm=T)==0]
+        zero_dates <- result_i$dates[rowSums(result_i$detected, na.rm=T)==0] # take out zero reporitng
+        zero_dates <- zero_dates[report.dates[i]-zero_dates < 10] # if more then ten days prob zero
+        zero.report = state%in%zero_dates
         deaths.i <- data.table(
             date          = state,
             state         =  report.dates[i],
@@ -936,16 +938,17 @@ uk.prediction <- function(result,
 #' @param y data
 ##
 likelihood_death <- function(x, D, y){
-    kappa       <- exp(x[1])
-    nu          <- exp(x[2])
+    kappa       <- exp(x[1]) + 10^-2
+    nu          <- 3/(1+exp(x[2]))
     sigma       <- exp(x[3])
     sigma_noise <- exp(x[4])
+    mu          <- 0#x[5]
     Sigma <- matern.covariance(D, kappa, nu, sigma)
     diag(Sigma)<- diag(Sigma) + sigma_noise^2
     R <- chol(Sigma, pivot=T)
     if(attr(R,"rank" )< length(y))
         return(-Inf)
-    v <- solve(t(R),y[attr(R,"pivot")])
+    v <- solve(t(R),y[attr(R,"pivot")]-mu)
     return( -sum(log(diag(R))) - 0.5*t(v)%*%v)
 }
 
@@ -962,7 +965,10 @@ likelihood_death <- function(x, D, y){
 
 gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_width = 0.95){
 
-
+    transform <- function(x){log(pmax(x,1))}
+    itransform <- function(x){exp(x)}
+    #transform <- function(x){x^{1/3}}
+    #itransform <- function(x){x^3}
     if(is.null(theta)){
         theta <- c(0,0,0,0)
     }
@@ -984,7 +990,7 @@ gp.smooth <- function(death_prediction, max.days.to.report, theta = NULL, CI_wid
                                         date    = death_prediction$date[index.known])
 
[…]
